---
title: "TX Water Systems - Bridge Grant Analysis"
author: "EmmaLi Tsai"
date: "2024-09-23"
output: 
  html_document:
    toc: true
    theme: united
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# 0.0 Set-up
```{r message = F, results='hide'}
# No scientific notation
options(scipen = 99999999)
options(tigris_use_cache = TRUE)

# Libraries: 
library(aws.s3)
library(sf)
library(tidyverse)
library(tidycensus)
library(scales)
library(leaflet)
library(plotly)
library(tigris)

# grabbing national xwalk: 
# NOTE: the xwalk has been updated with the fixed epa block parcel crosswalk 
# from blue conduit, and therefore likely has more "accurate" data for tier 
# 2 crosswalked systems. Going to resort to primarily using the xwalk 
xwalk <- aws.s3::s3read_using(st_read, 
                              object = "/service_area_boundaries/epa-sabs/epa-sabs-crosswalk.geojson",
                              bucket = "tech-team-data",
                              quiet = TRUE)

# grabbing data from TX application: 
app_data <-  aws.s3::s3read_using(st_read, 
                                  object = "/state-drinking-water/TX/clean/app/app-data-prod.geojson",
                                  bucket = "tech-team-data",
                                  quiet = TRUE)

# filtering xwalk for pwsids in app_data, and adding water system information 
# that's missing from the EPA crosswalk
app_data_ws_info <- app_data %>%
  as.data.frame() %>%
  select(pwsid:primary_source_code) 

tx_xwalk <- xwalk %>%
  filter(pwsid %in% app_data$pwsid) %>%
  left_join(app_data_ws_info, by = "pwsid")

# grabbing baseline 2020 census stats for TX for percentages: 
tx_county_pop <- tidycensus::get_acs(
  geography = "county",
  variables = c(total_pop = "B01003_001"),
  state = "TX",
  year = 2020,
  geometry = TRUE
)

# financial data from the tx project: 
fin <- aws.s3::s3read_using(readRDS, 
                            object = "s3://tech-team-data/state-drinking-water/TX/clean/TX_financial_list.RData")


# themes: 
sysfonts::font_add_google("Lato")
showtext::showtext_auto()

cat_palette <- colorRampPalette(c("#172f60","#1054a8",
                                  "#791a7b","#de9d29", 
                                  "#b15712","#4ea324"))
cont_palette <- colorRampPalette(c("#172f60","#4ea324"))

epic_chart_theme <- theme_minimal() + 
  theme(legend.position = "right", 
        text = element_text(size = 13, family = "Lato"), 
        legend.text = element_text(size = 10, family = "Lato"), 
        legend.title = element_text(size = 11, family = "Lato"), 
        axis.text.x = element_text(margin = margin(t = 10, r = 0, 
                                                   b = 0, l = 0)), 
        axis.title.x = element_text(margin = margin(t = 10, r = 0, 
                                                    b = 0, l = 0)), 
        axis.text.y = element_text(margin = margin(t = 0, r = 10, 
                                                   b = 0, l = 0)), 
        axis.title.y = element_text(margin = margin(t = 0, r = 10, 
                                                    b = 0, l = 0))) 

# function for getting sab xwalked stats over time: 
source("/Users/emmalitsai/TX-drinking-water/code/functions/state_sab_crosswalk_year.R")
```

# 1.0: What is the size of the opportunity?
## 1.1: What is the number of Texans served by CWS?
```{r}
# what source codes are we working with?
# unique(tx_xwalk$primary_source_code)

# summing population where source code is surface or surface purchased: 
tx_pop_sw <- tx_xwalk %>%
  as.data.frame() %>%
  # filter(primary_source_code %in% c("Surface", "Surface-Purchased")) %>%
  summarize(total_population_surface = sum(estimate_total_pop, na.rm = T))

# how does this compare to the total population in 2020?
tx_pop_total <- sum(tx_county_pop$estimate)
```
The number of Texans served by CWS is `r round(tx_pop_sw, 0)`. With `r tx_pop_total` people living in Texas, this reflects approximately `r round(100*(tx_pop_sw/tx_pop_total), 2)`% of the population. 

## 1.2: What is the amount of investment that goes into CWS across the state? 
```{r}
# From my notes, this is the TX DWSRF analysis from Phil & Danielle: 
# this is from https://github.com/Environmental-Policy-Innovation-Center/water-data/tree/main/clean-data/dwsrf-awards
# analysis doc: https://docs.google.com/document/d/1zzS5IASdb7CtN8WcIs8u17CaFpKJYjNrmriGB6_WVZQ/edit
# contains FOIA'd DWSRF awards from October 2009 - 2020
srf <- fin$TX_DWSRF_Analysis 
tx_srf <- srf %>% 
  filter(pwsid %in% tx_xwalk$pwsid)

# TODO OR - use the srf dashboard to pull 2020-2023?

# which ones have received funding?
srf_funded <- tx_xwalk %>%
  filter(pwsid %in% tx_srf$pwsid)

# what % of the population does this reflect?
# TODO-  is this the project_population_served?
srf_funded_pop <-  tx_xwalk %>%
  as.data.frame() %>%
  # select(pwsid, project_population_served) %>%
  # unique() %>%
  filter(pwsid %in% tx_srf$pwsid) %>%
  summarize(funded_pop = sum(estimate_total_pop, na.rm = T))

# what % of the population served by CWS does this reflect?
total_cws_pop <- sum(tx_xwalk$estimate_total_pop, na.rm = T)


# how much total assistance has been provided?
assistance_stats <- tx_srf %>%
  summarize(total_assistance = dollar(sum(assistance_amt)), 
            total_pf = dollar(sum(prin_forgive_amt)), 
            mean_assistance = dollar(mean(assistance_amt)), 
            mean_pf = dollar(mean(prin_forgive_amt)))

```
Out of `r formatC(length(unique(tx_xwalk$pwsid)), format = "d", big.mark = ",")` CWS in Texas, `r length(unique(srf_funded$pwsid))` have received Drinking Water State Revolving Funds over 2009-2020. This reflects `r round((length(unique(srf_funded$pwsid))/length(unique(tx_xwalk$pwsid))*100), 2)`% of CWS in Texas. 

Based on water system populations, systems that have been funded by DWSRF serve `r formatC(round(srf_funded_pop$funded_pop, 0), format = "d", big.mark = ",")` people, which reflects `r round((srf_funded_pop$funded_pop / total_cws_pop)*100, 0)`% of the population served by CWS in Texas, and `r round((srf_funded_pop$funded_pop / tx_pop_total)*100, 0)`% of the total population in Texas. 

```{r}
# how has the amount of total assistance/year changed through time?
srf_year_summary <- tx_srf %>% 
  group_by(year) %>%
  summarize(total_amt = sum(assistance_amt))

# plot to show results through time: 
ggplot(srf_year_summary, aes(x = year, y = total_amt)) + 
  ggchicklet::geom_chicklet(stat = "identity", 
                            color = "white", 
                            fill = "#1054a8", 
                            alpha = 0.7) + 
  scale_y_continuous(labels = scales::dollar_format(prefix="$")) + 
  epic_chart_theme + 
  labs(x = "Year", 
       y = "Total Assistance ($)") + 
  ggtitle("Amount of DWSRF Assistance has Increased Since 2009")

```

From 2009-2020, the amount of total assistance received by CWS in Texas is `r assistance_stats$total_assistance`, with an average of `r assistance_stats$mean_assistance` across our study period. The total amount of principal forgiveness is `r assistance_stats$total_pf`, with an average of `r assistance_stats$mean_pf`.

```{r}
# what counties have received the most funding? 
tx_funded_counties <- tx_xwalk %>%
  as.data.frame() %>%
  filter(pwsid %in% tx_srf$pwsid) %>%
  select(pwsid, county_served)

tx_srf_counties <- merge(tx_srf, tx_funded_counties, all = T)

tx_srf_county_summary <- tx_srf_counties %>%
  group_by(county_served) %>%
  summarize(total_assistance = sum(assistance_amt), 
            total_yrs_funded = length(unique(year)), 
            assistance_per_yr = total_assistance/total_yrs_funded) %>%
  mutate(county_clean = paste0(county_served, " County, Texas"))

# merging with TX county geometries for mapping: 
tx_county_simple <- tx_county_pop %>%
  select(NAME, geometry) %>%
  unique()
tx_srf_county_geom <- merge(tx_srf_county_summary, tx_county_pop, 
                            by.x = "county_clean", by.y = "NAME", 
                            all.x = T) %>%
  st_as_sf() 


# plotting: 
assistance_pal <- colorNumeric(
  palette = cont_palette(5),
  domain = tx_srf_county_geom$total_assistance)

leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron, 
                   group = "Toner Lite") %>%
    addPolygons(data = tx_county_pop,
              # slight opacity
              fillOpacity = 0.2,
              color = "gray",
              weight = 1, 
              smoothFactor = 0) %>%
  addPolygons(data = tx_srf_county_geom,
              fillColor = ~assistance_pal(total_assistance),
              # slight opacity
              fillOpacity = 0.8,
              color = "white",
              weight = 1, 
              smoothFactor = 0, 
              label = paste0("County: ", tx_srf_county_geom$county_served, 
                             "; Total Assistance : ",
                             dollar(tx_srf_county_geom$total_assistance), 
                             "; Total Years Funded: ",
                             (tx_srf_county_geom$total_yrs_funded), 
                             "; Assistance/Year: ",
                             dollar(tx_srf_county_geom$assistance_per_yr))) %>%
  addLegend("bottomright",
            pal = assistance_pal,
            values = tx_srf_county_geom$total_assistance,
            # if working with money or percentages, use the 
            # follow label formats: 
            labFormat = labelFormat(prefix = "$"),
            # labFormat = labelFormat(prefix = "%"),
            title = "Total Assistance ($)\n - 2009 to 2020",
            opacity = 1)

```
Which counties have received the most funding?

The counties that have received the largest total assistance over our study period are located around large urban centers such as Dallas (both Dallas & Tarrant county), San Antonio (Bexar county), and Houston (Brazoria county). Bexar county around San Antonio has received the largest amount of funding over our study period, with ~$200 million received over the course of 7 years. 
```{r}
# what are the characteristics of utilities that have received funding?
tx_funding_flag <- tx_xwalk %>%
  as.data.frame() %>%
  mutate(funded = case_when(
    pwsid %in% tx_srf$pwsid ~ "Funded", 
    TRUE ~ "Not Funded"
  )) %>%
  select(pwsid, funded, estimate_poc_alone_per, estimate_mhi, 
         estimate_laborforce_unemployed_per, population_served_count,
         dwater, 
         # primary_source_code, 
         estimate_hh_below_pov_per) %>%
  rename("%POC" = estimate_poc_alone_per, 
          "MHI" = estimate_mhi, 
         `Population Served` = population_served_count,
         `% Unemployment` = estimate_laborforce_unemployed_per, 
          `Drinking Water Metric` = dwater,
          # `Water Source` = primary_source_code, 
          `% Households in Poverty` = estimate_hh_below_pov_per)


# summary stats for text: 
summary_stats <- tx_funding_flag %>%
  group_by(funded) %>%
  summarize(across(2:7, ~mean(.x, na.rm = T))) %>%
  rename("poc" = `%POC`, 
         "unemp" =  `% Unemployment`, 
         "pop" =`Population Served`, 
         "dw" = `Drinking Water Metric`,
         "poverty" = `% Households in Poverty`)

funded <- summary_stats %>%
  filter(funded == "Funded")
not_funded <- summary_stats %>%
  filter(funded != "Funded")

# is there any difference by source water? 
by_source <- tx_xwalk %>%
  as.data.frame() %>%
  mutate(funded = case_when(
    pwsid %in% tx_srf$pwsid ~ "Funded", 
    TRUE ~ "Not Funded"
  )) %>%
  select(pwsid, funded, primary_source_code) %>%
  group_by(funded, primary_source_code) %>%
  tally()

# long for plotting: 
tx_funding_long <- tx_funding_flag %>%
  pivot_longer(., cols = 3:8)

ggplot(tx_funding_long, aes(x = funded, y = value, 
                            color = funded)) + 
  geom_jitter(width = 0.05, alpha = 0.2) + 
  geom_boxplot(alpha = 0.6, 
                 outliers = FALSE) + 
  facet_wrap(~name, scales = "free") + 
  scale_color_manual(values = cat_palette(2), 
                     name = "") + 
  epic_chart_theme +
  labs(x = "", y = "Value") + 
  theme(legend.position = "bottom")
```

What are the characteristics of utilities that have received funding?

Based on the metrics plotted above, utilities that have received funding have a larger mean percentage of households in poverty by `r round(funded$poverty - not_funded$poverty, 2)`%. These utilities also have a larger mean population they serve by `r round(funded$pop - not_funded$pop, 0)` people, likely because many of the utilities that have received funding provide services to large population centers, as shown in the question above. Funded systems also have a lower median household income than unfunded systems by `r dollar(abs(funded$MHI - not_funded$MHI))`, and serve less diverse communities. Both funded and unfunded systems were found to be primarily pulling groundwater. 

One of the most striking difference across all of our indicators was EPA's new drinking water metric, where systems that received funding had a mean value that was `r round(funded$dw - not_funded$dw, 2)` larger than systems that have not received funding. 


# 2.0: What are the characteristics of a fragile system?
```{r}
# TX SRF DAC definition: https://www.epa.gov/system/files/documents/2022-10/DWSRF%20DAC%20Definitions%20Report_October%202022%20Updates_FINAL_508.pdf
# affordability criteria: 
# (a) Has an Annual Median Household Income (AMHI) that is no more than 75 percent of the state median household income using an acceptable source of socioeconomic data, and
# (b) the Household Cost Factor (HCF) that considers income, unemployment rates, and population trends must be greater than or equal to 1 percent if only water or sewer service is provided or greater than or equal to 2 percent if both water and sewer service are provided.

# The Household Cost Factor in the IUP is the cost of water and wastewater utilities per household (after completion of the proposed project) in the project service area compared to the project area’s current median household income level, with two adjustments. Specifically, the formula is:

# A=Average annual water bill per household
# B= Average annual wastewater bill per household
# C= Annual loan cost per household of the new financing
# D= AMHI for the project service area
# HCF = (A + B + C) / D

# In addition, the state includes two adjustments that can only increase the HCF (i.e., benefit the applicant) - this includes unemployment rates & population adjustments 


# could ask water team for loan costs 


# TX definition of "safe, adequate, and reliable"
# Texas health definition code - approved or superior 
# test <- aws.s3::s3read_using(read.csv, 
#                               object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_TCEQ_DWW/TX_20231217/WaterSystemDetail/TX_Water System Details_20231217_20231229203427.csv")
# test %>%
#   filter(Type == "System  Recognition") %>%
#   group_by(Value) %>%
#   tally()
# lol nevermind: 
#   Value           n
#   <chr>       <int>
# 1 APPROVED       19
# 2 NO DATA     13625
# 3 PROVISIONAL     8
# 4 SUPERIOR      738

# check this: s3://tech-team-data/state-drinking-water/TX/raw/TX_TCEQ_DWW/TX_20231217/WaterSystemDetail/TX_Water System Details_20231217_20231229203427.csv

# TX definition of a failing system 
```

# 3.0: How do water systems change over time in terms of fragility and durability? 
```{r eval = FALSE}
# prep for this section: 

# which system have been around for that long?
# NOTE - absolute file paths needed to knitting rmd 
sdwa_info <- read.csv("/Users/emmalitsai/TX-drinking-water/data/raw/sdwa_download/SDWA_FACILITIES.csv") %>%
  mutate(start_date = as.Date(FIRST_REPORTED_DATE, tryFormats = c("%m/%d/%Y")),
         year_date = year(start_date))

# # NOTE - this is pwsids with FACILITIES where the first report is at least 2010
pwsids_since_2010 <- sdwa_info %>%
  filter(year_date <= 2010) %>%
  filter(PWSID %in% tx_xwalk$pwsid) %>%
  distinct(PWSID)

# # filtering SABs for just these 
tx_xwalk_2010_pwsids <- tx_xwalk %>%
  filter(pwsid %in% pwsids_since_2010$PWSID) %>%
  select(pwsid:shape_length)

# NOTE - only need to run this if running the xwalk again - shouldn't need
# to run while knitting 
# need to reconstruct block housing and population weights - the tigris block()
# function does not return this information from 2010 - 2019. 
# grabbing 2020: 
state_blocks_2020 <- tigris::blocks(
  state = "TX",
  year = 2020)

# grabbing 2010:
# NOTE - with going back to 2010, we've lost education categories and 
# % unemployment 
# dec_vars_all <- load_variables(2010, "sf1", cache = TRUE)
dec_vars <- c(HOUSING10 = "H001001",
              POP10 = "P001001")
state_blocks_2010 <- get_decennial(
  geography = "block",
  variables = dec_vars,
  year = 2010,
  state = "TX",
  geometry = TRUE
)

# pivoting to wide to work with xwalk: 
state_blocks_2010_wide <- state_blocks_2010 %>%
  pivot_wider(., names_from = variable, values_from = value)

# 2015 is going to be tricky - grabbing a relationship file:
file <-"https://www2.census.gov/geo/docs/maps-data/data/rel2020/t10t20/TAB2010_TAB2020_ST48.zip"
file_loc <- "/Users/emmalitsai/TX-drinking-water/data/raw/census_rel"
download.file(file,
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(fsile_loc, ".zip"), exdir = file_loc)
file.remove(paste0(file_loc, ".zip"))

# reading in block relationships from 2010 to 2020 in TX:
block_2010_2020_rel <- read.table("/Users/emmalitsai/TX-drinking-water/data/raw/census_rel/tab2010_tab2020_st48_tx.txt", 
                                  sep = "|")
# correcting weird column names: 
names(block_2010_2020_rel) <- block_2010_2020_rel[1,]
block_2010_2020_rel <- block_2010_2020_rel[2:nrow(block_2010_2020_rel),]

# building block relationships
block_rel <- block_2010_2020_rel %>%
  mutate(GEOID_10 = paste0(STATE_2010, COUNTY_2010, TRACT_2010, BLK_2010), 
         GEOID_20 = paste0(STATE_2020, COUNTY_2020, TRACT_2020, BLK_2020))

# want to keep 2010 block geometries, but grab mean housing and pop between two 
# time periods
block_2020_simple <- state_blocks_2020 %>%
  as.data.frame() %>%
  select(GEOID20, HOUSING20, POP20) 
block_2020_rel <- merge(block_2020_simple, block_rel, by.x = "GEOID20", 
                        by.y = "GEOID_20", all = T)
block_2010_rel <- merge(state_blocks_2010_wide, block_2020_rel, 
                        by.x = "GEOID", by.y = "GEOID_10", all = T)

# grabbing mean between years of housing data, AND grouping by 2010 
# geoid to calculate mean values for each original 2010 geoid (some blocks 
# were either merged or split over the decade)
state_blocks_2015 <- block_2010_rel %>%
  as.data.frame() %>%
  filter(!(is.na(HOUSING10))) %>%
  select(GEOID, GEOID20, HOUSING10, HOUSING20, POP10, POP20) %>%
  rowwise() %>%
  mutate(HOUSING15 = mean(c(HOUSING10, HOUSING20)), 
         POP15 = mean(c(POP10, POP20))) %>%
  group_by(GEOID) %>%
  summarize(HOUSING15 = mean(HOUSING15), 
            POP15 = mean(POP15))

# adding original 2010 geometries back: 
geom_2010 <- state_blocks_2010_wide %>%
  select("GEOID")
state_blocks_2015 <- merge(state_blocks_2015, geom_2010, by = "GEOID") %>%
  st_as_sf()

# crosswalkin' 2015 and 2010: 
tx_xwalk_2015 <- state_sab_crosswalk_year(tx_xwalk_2010_pwsids, "TX", 2015,
                                          state_blocks_2015,
                                          "WGS84")
# st_write(tx_xwalk_2015, "/Users/emmalitsai/TX-drinking-water/data/raw/tx_xwalk_2015.geojson")
# put_object(
#   file = "/Users/emmalitsai/TX-drinking-water/data/raw/tx_xwalk_2015.geojson",
#   object = "/state-drinking-water/TX/raw/TX_xwalk_2015.geojson",
#   bucket = "tech-team-data"
# )

tx_xwalk_2010 <- state_sab_crosswalk_year(tx_xwalk_2010_pwsids, "TX", 2010,
                                            state_blocks_2010_wide,
                                            "WGS84")
# st_write(tx_xwalk_2010, "/Users/emmalitsai/TX-drinking-water/data/raw/tx_xwalk_2010.geojson")
# put_object(
#   file = "/Users/emmalitsai/TX-drinking-water/data/raw/tx_xwalk_2010.geojson",
#   object = "/state-drinking-water/TX/raw/TX_xwalk_2010.geojson",
#   bucket = "tech-team-data"
# )

# download SDWA database and look over entire timescale 
file_loc <- "/Users/emmalitsai/TX-drinking-water/data/raw/sdwa_download"
# download.file("https://echo.epa.gov/files/echodownloads/SDWA_latest_downloads.zip", 
#               destfile = paste0(file_loc, ".zip"))
# unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
# file.remove(paste0(file_loc, ".zip"))

```

## 3.1: What changes do we see to population growth in fragile systems? 
```{r fig.height = 9, fig.width = 10}
# population growth through time, and ideally this would also include 
# MHI 
tx_xwalk_2010 <- aws.s3::s3read_using(st_read, 
                                      object = "/state-drinking-water/TX/raw/TX_xwalk_2010.geojson",
                                      bucket = "tech-team-data",
                                      quiet = TRUE)

tx_xwalk_2015 <- aws.s3::s3read_using(st_read, 
                                      object = "/state-drinking-water/TX/raw/TX_xwalk_2015.geojson",
                                      bucket = "tech-team-data",
                                      quiet = TRUE)

# tidying and mutating: 
tx_xwalk_2015 <- tx_xwalk_2015 %>% 
  mutate(year = "2015")
tx_xwalk_2010 <- tx_xwalk_2010 %>% 
  mutate(year = "2010")
tx_xwalk_2020 <- tx_xwalk %>%
  filter(pwsid %in% tx_xwalk_2010$pwsid) %>%
  mutate(year = "2020")

# selecting only columns of interest for now: 
tx_2015_simple <- tx_xwalk_2015 %>%
  as.data.frame() %>%
  select(pwsid, tier_crosswalk, estimate_total_pop, estimate_mhi)
names(tx_2015_simple) <- paste0(names(tx_2015_simple), "_2015")

tx_2010_simple <- tx_xwalk_2010 %>%
  as.data.frame() %>%
  select(pwsid, tier_crosswalk, estimate_total_pop, estimate_mhi)
names(tx_2010_simple) <- paste0(names(tx_2010_simple), "_2010")

tx_2020_simple <- tx_xwalk_2020 %>%
  as.data.frame() %>%
  select(pwsid, tier_crosswalk, estimate_total_pop, estimate_mhi)
names(tx_2020_simple) <- paste0(names(tx_2020_simple), "_2020")

# comparisons over a decade: 
tx_decade <- merge(tx_2010_simple, tx_2015_simple, 
                   by.x = "pwsid_2010", 
                   by.y = "pwsid_2015", all = T) 

tx_decade <- merge(tx_decade, tx_2020_simple, by.x = "pwsid_2010", by.y = "pwsid_2020", all = T) %>%
# reducing analysis to ONLY tier 1 crosswalk for ALL time periods 
  filter(tier_crosswalk_2010 == "tier_1" & tier_crosswalk_2020 == "tier_1" & tier_crosswalk_2015 == "tier_1") %>%
  # adjusting for inflation - using: https://data.bls.gov/cgi-bin/cpicalc.pl?cost1=2000&year1=201001&year2=202001
  mutate(estimate_mhi_2010_inf_adj = estimate_mhi_2010*1.190525, 
         estimate_mhi_2015_inf_adj = estimate_mhi_2015*1.10283, 
         pct_change_pop_2010_2015 = 100*((estimate_total_pop_2015 - estimate_total_pop_2010)/estimate_total_pop_2010),
         pct_change_mhi_2010_2015 = 100*((estimate_mhi_2015_inf_adj - estimate_mhi_2010_inf_adj)/estimate_mhi_2010_inf_adj),
         pct_change_pop_2015_2020 = 100*((estimate_total_pop_2020 - estimate_total_pop_2015)/estimate_total_pop_2015),
         pct_change_mhi_2015_2020 = 100*((estimate_mhi_2020 - estimate_mhi_2015_inf_adj)/estimate_mhi_2015_inf_adj),
         pct_change_pop_2010_2020 = 100*((estimate_total_pop_2020 - estimate_total_pop_2010)/estimate_total_pop_2010), 
         pct_change_mhi_2010_2020 = 100*((estimate_mhi_2020 - estimate_mhi_2010_inf_adj)/estimate_mhi_2010_inf_adj)) %>%
  rename(pwsid = pwsid_2010)

# adding geoms back: 
tx_geoms <- tx_xwalk %>% 
  select(pwsid, county_served)

tx_decade_geoms <- tx_decade %>%
  left_join(tx_geoms, by = "pwsid") %>%
  st_as_sf()

# plotting: 
pop_pal <- colorBin(
  bins = c(-90, -20, 0, 20, 50, 100, 20000),
  palette = c("#d73027", "#fc8d59", "white", "#d9ef8b", "#91cf60", "#1a9850", "darkgreen"),
  domain = tx_decade_geoms$pct_change_pop_2010_2020)

leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron, 
                   group = "Toner Lite") %>%
  addPolygons(data = tx_decade_geoms,
              fillColor = ~pop_pal(pct_change_pop_2010_2020),
              # slight opacity
              fillOpacity = 0.8,
              stroke = T, 
              color = "black",
              opacity = 0.2, 
              weight = 1,
              # smoothFactor = 0.5, 
              label = paste0("pwsid: ", tx_decade_geoms$pwsid, 
                             "; % pop change : ",
                             round(tx_decade_geoms$pct_change_pop_2010_2020, 0))) %>%
  addLegend("bottomright",
            pal = pop_pal,
            values = tx_decade_geoms$pct_change_pop_2010_2020,
            # if working with money or percentages, use the 
            # follow label formats: 
            # labFormat = labelFormat(prefix = "$"),
            labFormat = labelFormat(prefix = "%"),
            title = "% Population Change 2010 to 2020",
            opacity = 1)
```

Broadly, systems in suburban areas around cities have grown, while populations within cities have shrunk since 2010. Water systems located in NE Texas have generally exhibited population decline since 2010, along with more rural areas of Texas. 

```{r fig.height = 9, fig.width = 10}
mhi_pal <- colorBin(
  bins = c(-100, -50, -20, 0, 20, 100, 300),
  palette = c("#d73027", "#fc8d59", "white", "#d9ef8b", "#91cf60", "#1a9850", "darkgreen"),
  domain = tx_decade_geoms$pct_change_mhi_2010_2020)

leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron, 
                   group = "Toner Lite") %>%
  addPolygons(data = tx_decade_geoms,
           fillColor = ~mhi_pal(pct_change_mhi_2010_2020),
              # slight opacity
              fillOpacity = 0.8,
              stroke = T, 
              color = "black",
              opacity = 0.2, 
              weight = 1,
              # smoothFactor = 0.5, 
              label = paste0("pwsid: ", tx_decade_geoms$pwsid, 
                             "; % MHI change : ",
                             round(tx_decade_geoms$pct_change_mhi_2010_2020, 0))) %>%
  addLegend("bottomright",
            pal = mhi_pal,
            values = tx_decade_geoms$pct_change_mhi_2010_2020,
            # if working with money or percentages, use the 
            # follow label formats: 
            # labFormat = labelFormat(prefix = "$"),
            # labFormat = labelFormat(prefix = "%"),
            title = "% MHI Change 2010 to 2020",
            opacity = 1)
```

Median household income (already adjusted for inflation) within CWS in Texas have broadly increased. There are some water systems located in mostly rural areas that have exhibited slight declines in MHI over the past decade. 

```{r fig.height = 9, fig.width = 10}

# looking at % change in population and % change in median household income: 
leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron, 
                   group = "Toner Lite") %>%
  bivariatechoropleths::addBivariateChoropleth(
    map_data = tx_decade_geoms,
    var1_name = pct_change_pop_2010_2020,
    var2_name = pct_change_mhi_2010_2020,
    ntiles= 3,
    var1_label = "Population % Change",
    var2_label = "MHI % Change",
    # region_name = "CSDNAME",
    weight = 1,
    fillOpacity = 0.7,
    color = "grey",
    highlightOptions = leaflet::highlightOptions(color = "black",
                                                 weight = 2,
                                                 opacity = 1)) %>%
  # if I don't add this, we can't see pwsids :(
  addPolygons(data = tx_decade_geoms, 
              fillOpacity = 0, 
              opacity = 0, 
              label = paste0("pwsid: ", tx_decade_geoms$pwsid,
                             "; % MHI change : ",
                             round(tx_decade_geoms$pct_change_mhi_2010_2020, 0),
                             "; % Pop change : ",
                             round(tx_decade_geoms$pct_change_pop_2010_2020, 0)))
```

Above, you can see CWS boundaries mapped by % change in population and MHI from 2010-2020. The areas of particular interest are in bright teal, where the CWS exhibited increases in population, but little to no change in MHI. Areas located in white are also of interest, as this indicates a system that is losing both populations and income from 2010-2020. 

```{r, results = "asis"}
cws_of_interest <- tx_decade_geoms %>%
  mutate(Category = case_when(
    (pct_change_pop_2010_2020 > 0 & pct_change_mhi_2010_2020 < 0) ~ "Pop Growing, MHI Decreasing", 
    (pct_change_pop_2010_2020 < 0 & pct_change_mhi_2010_2020 < 0) ~ "Pop & MHI Decreasing", 
  TRUE ~ "No Concern Yet"))

table <- cws_of_interest %>% 
  as.data.frame() %>%
  group_by(Category) %>%
  summarize(CWS = n())
knitr::kable(table, caption = "Number of CWS in each Category")
```

While most CWS exhibit positive MHI and growth over the past decade, there are ~836 CWS in TX where the population has grown over time, but the MHI has decreased. There are 490 that have decreased in both population and MHI over the course of the past decade - indicating the system may be serving less people and with lower income now than in 2010. 

```{r fig.height = 9, fig.width = 10}
# plotting 
# ggplot(tx_decade, aes(x = pct_change_pop, y = pct_change_mhi)) + 
#   geom_point() + 
#   labs(x = "% Population Change", 
#        y = "% MHI Change") + 
#   epic_chart_theme

# combining all of the data: 
tx_2010_all <- tx_xwalk_2010 %>%
  as.data.frame() %>%
  select(-starts_with("geom"))

tx_2015_all <- tx_xwalk_2015 %>%
  as.data.frame() %>%
  select(-starts_with("geom"))

tx_2020_all <- tx_xwalk_2020 %>%
  as.data.frame() %>%
  select(-starts_with("geom")) %>%
  select(names(tx_2010_all)) %>%
  mutate(date_modified = as.character(date_modified), 
         verification_date = as.character(verification_date))

tx_decade_all <- bind_rows(tx_2010_all, tx_2015_all, 
                           tx_2020_all) %>%
  select(pwsid, estimate_total_pop:year) %>%
  select(-contains("age")) %>%
  select(pwsid, estimate_white_alone_per:estimate_hh_below_pov_per,
         estimate_pop_below_200_pov_per, year)


# pivoting to long for formatting:  
tx_decade_long <- tx_decade_all %>%
  pivot_longer(., estimate_white_alone_per:estimate_pop_below_200_pov_per)

ggplot(tx_decade_long, aes(x = year, 
                           y = value, color = year)) + 
  geom_jitter(width = 0.2, alpha = 0.1) +
  geom_violin(alpha = 0.8) +
  facet_wrap(~name, scales = "free") + 
  epic_chart_theme + 
  scale_color_manual(values = cat_palette(3), 
                     name = "") +
  labs(x = "Year", y = "Value")
```

Demographic data for CWS in TX are relatively similar across the study period. Broadly, the %mixed population served by CWS in TX have increased sinc 2010, and the % of households below the poverty level have decreased. 


## 3.2: What changes in violations do we see in fragile systems? 
```{r fig.height = 9, fig.width = 10}
file_loc <- "/Users/emmalitsai/TX-drinking-water/data/raw/sdwa_download"

# reading in violations: 
sdwa_viols <- read.csv(paste0(file_loc, "/SDWA_VIOLATIONS_ENFORCEMENT.csv")) %>%
  janitor::clean_names() 
tx_viols <- sdwa_viols %>% 
  filter(pwsid %in% tx_xwalk$pwsid) %>%
  mutate(rule_code = as.character(rule_code))

# violation codes:
sdwa_codes <- read.csv(paste0(file_loc, "/SDWA_REF_CODE_VALUES.csv")) %>%
  janitor::clean_names() 
rule_codes <- sdwa_codes %>% 
  filter(value_type %in% c("RULE_FAMILY_CODE")) %>%
  rename(rule = value_description)

# merging by rule code
tx_viol_rulecode <- merge(tx_viols, rule_codes, 
                          by.x = "rule_family_code", 
                          by.y = "value_code", 
                          all.x = T)

# grabbing summaries by rule code & creating stacked bar plot: 
health_viol <- tx_viol_rulecode %>%
  mutate(start_date = as.Date(viol_first_reported_date, 
                              tryFormats = c("%m/%d/%Y")), 
         start_year = year(start_date)) %>%
  filter(is_health_based_ind == "Y") %>%
  group_by(start_year, rule) %>%
  summarize(total = n())

health_viol_plot <- ggplot(health_viol, aes(x = start_year, y = total, 
                        fill = rule)) + 
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = cat_palette(11), 
                     name = "") +
  theme_minimal() + 
  theme(legend.position = "right", 
        text = element_text(size = 13), 
        legend.text = element_text(size = 10), 
        legend.title = element_text(size = 11), 
        axis.text.x = element_text(margin = margin(t = 10, r = 0, 
                                                   b = 0, l = 0)), 
        axis.title.x = element_text(margin = margin(t = 10, r = 0, 
                                                    b = 0, l = 0)), 
        axis.text.y = element_text(margin = margin(t = 0, r = 10, 
                                                   b = 0, l = 0)), 
        axis.title.y = element_text(margin = margin(t = 0, r = 10, 
                                                    b = 0, l = 0))) +
  labs(x = "Year of Health Violation", y = "Total Number of Rule Violations")

plotly::ggplotly(health_viol_plot)
```

Prior to 2017, the majority of health-based violations in Texas appear to be stage 1 disinfectants & inorganic chemicals. After 2016, the majority of violations switched to stage 2 disinfectants & inorganic chemicals. Health-based violations peaked in 2017, but has decreased over time. 

Maybe chat w/ EPA folks or do some research online??? 
1991 stage 2 - maybe chat with Michael Pastino? (dw information systems branch manager)
How much of this is just showing the history of SDWIS?

## 3.3: What changes in water rates and MHI do we see in fragile systems? 
```{r fig.height = 9, fig.width = 10}
# order of operations: 
# filter pwsids to those that have been around for the entire time period of interest
# populations, MHI, and water rates for census tracts through time (even when boundaries have changed !)
# crosswalk data to pwids for each year (which might've not even existed at a certain time period...?) this also involves writing a whole new xwalk function for just a couple 
# of variables 

# summarize sdwis violation data by appropriate year bins 

# -----
# water rates: B25134 - only go back to 2021, and only in 1yr acs which
# "The 1-year ACS provides data for geographies with populations of 65,000 and greater."
# so it only has data for large counties
# all_census_vars <- load_variables(2023, "acs1", cache = TRUE)

census_vars <- c(total_pop = "B01003_001", 
                 mhi = "B19013_001", 
                 total_water = "B25134_003", 
                 less_125 = "B25134_004", 
                 between_125_249 = "B25134_005", 
                 between_250_499 = "B25134_006", 
                 between_500_749 = "B25134_007", 
                  between_750_999 = "B25134_008", 
                 over_1000 = "B25134_009")

counties_2023 <- tidycensus::get_acs(
  geography = "county", 
  variables = census_vars, 
  survey = "acs1",
  year = 2023
) 

counties_2022 <- tidycensus::get_acs(
  geography = "county", 
  variables = census_vars, 
  survey = "acs1",
  year = 2022
)

counties_2021 <- tidycensus::get_acs(
  geography = "county", 
  variables = census_vars, 
  survey = "acs1",
  year = 2021
)

# quick combining for county data: 
tx_2023 <- counties_2023 %>%
  filter(grepl("^48", GEOID)) %>%
  select(-moe) %>%
  pivot_wider(., names_from = variable, values_from = estimate) %>%
  mutate(acs1yr_year = "2023")
tx_2022 <- counties_2022 %>%
  filter(grepl("^48", GEOID)) %>%
  select(-moe) %>%
  pivot_wider(., names_from = variable, values_from = estimate) %>%
  mutate(acs1yr_year = "2022")
tx_2021 <- counties_2021 %>%
  filter(grepl("^48", GEOID)) %>%
  select(-moe) %>%
  pivot_wider(., names_from = variable, values_from = estimate) %>%
  mutate(acs1yr_year = "2021")
# 2020 had a low response rate
# yeah - can confirm we don't have water rate data prior to 2021
# counties_2019 <- tidycensus::get_acs(
#   geography = "county", 
#   variables = census_vars, 
#   survey = "acs1",
#   year = 2019
# )


# calculating % of total population paying for water in each $ category: 
all_rates <- bind_rows(tx_2023, tx_2022, tx_2021) %>%
  mutate_at(vars(!c("GEOID", "NAME", "total_pop", 
                    "mhi", "acs1yr_year", "total_water")), 
            ~((./total_water)*100))

# pivoting to long for plotting: 
all_rates_long <- all_rates %>%
  pivot_longer(., cols = less_125:over_1000) %>%
  mutate(mhi_cat = case_when(
    mhi < 40000 ~ "Less than $40,000", 
    mhi >= 40000 & mhi < 60000 ~ "$40,000 - $59,999",
    mhi >= 60000 & mhi < 80000 ~ "$60,000 - $79,999",
    mhi >= 80000 & mhi < 100000 ~ "$80,000 - $99,999",
    mhi >= 100000 & mhi < 120000 ~ "$100,000 - $120,000",
    mhi >= 120000 ~ "Over $120,000"
  ), 
  mhi_acs = paste0(mhi_cat, acs1yr_year))

# plotting: 
# ggplot(all_rates_long, aes(x = mhi_cat, y = value, 
#                            color = acs1yr_year, 
#                            fill = acs1yr_year, 
#                            group = mhi_acs)) + 
#   geom_point( alpha = 0.2, 
#               position = position_dodge(width = 0.75)) + 
#   geom_boxplot(alpha = 0.4, outliers = F) + 
#   facet_wrap(~name) + 
#   coord_flip() + 
#   epic_chart_theme + 
#   labs(y = "% of Rate Payers", 
#        x = "MHI Category")

# rates_barplot <- ggplot(all_rates_long, aes(x = value, y = mhi_cat, fill = name)) + 
#   geom_bar(stat = "identity") + 
#   # geom_smooth(se = F) + 
#   facet_wrap(~acs1yr_year, ncol = 1) +
#   epic_chart_theme +  
#   labs(y = "% of Rate Payers", 
#        x = "MHI")
# 
# plotly::ggplotly(rates_barplot) 

# looking at counties where there is a high % of rate payers paying >750, 
# but income is < 40,000
low_mhi_high_rates <- all_rates_long %>%
  filter(name %in% c("between_750_999", "over_1000") & mhi < 60000) %>%
  pivot_wider(., names_from = name, values_from = value) %>%
  mutate(total_750_1000 = (between_750_999 + over_1000))

counties_of_interest <- unique(low_mhi_high_rates$NAME) 
counties_tidy <- counties_of_interest %>%
  gsub(" County, Texas", "", .)

# which ones serve Starr county?
starr_cws <- tx_xwalk %>% 
  filter(county_served == "Starr") 
starr_pwsid <- unique(starr_cws$pwsid)


of_interest <- ggplot(low_mhi_high_rates, 
                      aes(x = mhi, y = total_750_1000, 
                          color = NAME, 
                          text = NAME)) + 
  geom_point() + 
  facet_wrap(~acs1yr_year, ncol = 1) + 
  labs(y = "% of Rate Payers Paying > $750 Annually", 
       x = "MHI") + 
  theme_minimal() 
plotly::ggplotly(of_interest, tooltip = "NAME") 

# TODO - calculate water rate burden 
# calculate % change in water rate burden 
```

Counties where the median household income is less than $60,000 and the annual rate for water and sewer is greater than 750/year: `r counties_tidy`. Starr county, in particular, had counties where the MHI was less than 45,000 but nearly 40% of rate payers were spending more than 750/year for water and sewer. CWS that serve this county include: `r starr_pwsid`

```{r fig.height = 9, fig.width = 10}
# can I calculate mean water rates?
all_rates_calc <- bind_rows(tx_2023, tx_2022, tx_2021) 

# one way to do this is multiply the midpoint of each bin by # of people, 
# sum that information, and then divide by total number of people 
# NOTE - this is not perfect! and could be improved! 
all_rates_calc <- all_rates_calc %>%
  rowwise() %>%
  mutate(
    # finding total $ contributed by each bin using the bin's midpoint:
    less_125_mp = less_125 *62,
    between_125_249_mp = between_125_249*187, 
    between_250_499_mp = between_250_499 *374.5,
    between_500_749_mp = between_500_749*624.5, 
    between_750_999_mp = between_750_999*874.5, 
    # NOTE - assuming "midpoint" for >100 cat is just a half of a full
    # bin above 1000
    over_1000_mp = over_1000*1124.5,
    # summing total amount: 
    total = sum(less_125_mp, between_125_249_mp, between_250_499_mp,
                between_500_749_mp, between_750_999_mp, over_1000_mp), 
    # diving by total people to get a ROUGH mean
    mean_rate = total/total_water, 
    # finding pct of mhi spent on water annually
    rate_div_mhi = (mean_rate/mhi)*100
  )

# combien with census geos: 
tx_county_geo <- tx_county_pop %>%
  select(NAME)

all_rates_geo <- all_rates_calc %>%
  left_join(tx_county_geo) %>%
  st_as_sf() 

# plotting: 
rate_pal <- colorNumeric(
  palette = cont_palette(5),
  domain = all_rates_geo$rate_div_mhi)

leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron, 
                   group = "Toner Lite") %>%
    addPolygons(data = tx_county_pop,
              # slight opacity
              fillOpacity = 0.2,
              color = "gray",
              weight = 1, 
              smoothFactor = 0) %>%
  addPolygons(data = all_rates_geo[all_rates_geo$acs1yr_year == "2023",],
              fillColor = ~rate_pal(rate_div_mhi),
              # slight opacity
              fillOpacity = 0.8,
              color = "white",
              weight = 1, 
              smoothFactor = 0, 
              group = "2023", 
              label = paste0("County: ", all_rates_geo[all_rates_geo$acs1yr_year == "2023",]$NAME, 
                             "; % MHI spent on Water & Sewer : ",
                            round(all_rates_geo[all_rates_geo$acs1yr_year == "2023",]$rate_div_mhi,
                                  3), "%")) %>%
    addPolygons(data = all_rates_geo[all_rates_geo$acs1yr_year == "2022",],
              fillColor = ~rate_pal(rate_div_mhi),
              # slight opacity
              fillOpacity = 0.8,
              color = "white",
              weight = 1, 
              smoothFactor = 0, 
              group = "2022", 
              label = paste0("County: ", all_rates_geo[all_rates_geo$acs1yr_year == "2022",]$NAME, 
                             "; % MHI spent on Water & Sewer : ",
                            round(all_rates_geo[all_rates_geo$acs1yr_year == "2022",]$rate_div_mhi,
                                  3), "%")) %>%
      addPolygons(data = all_rates_geo[all_rates_geo$acs1yr_year == "2021",],
              fillColor = ~rate_pal(rate_div_mhi),
              # slight opacity
              fillOpacity = 0.8,
              color = "white",
              weight = 1, 
              smoothFactor = 0, 
              group = "2021", 
              label = paste0("County: ", all_rates_geo[all_rates_geo$acs1yr_year == "2021",]$NAME, 
                             "; % MHI spent on Water & Sewer : ",
                            round(all_rates_geo[all_rates_geo$acs1yr_year == "2021",]$rate_div_mhi, 3), "%")) %>%
  addLegend("bottomright",
            pal = rate_pal,
            values = all_rates_geo$rate_div_mhi,
            # if working with money or percentages, use the 
            # follow label formats: 
            # labFormat = labelFormat(prefix = "$"),
            labFormat = labelFormat(prefix = "%"),
            title = "% MHI spent on Water",
            opacity = 1) %>% 
  addLayersControl(baseGroups = c('2023', '2022', "2021"),
                   options = layersControlOptions(collapsed = FALSE),
                   position = 'bottomleft')
```

Across the years and counties in which we have data, counties closer to the southern border appear to spend a larger percentage of their income of water in comparison to the rest of the counties included in our study. Starr county had the largest percentage in our study, where 17% of income was going towards sewer and water in 2021. 


```{r fig.height = 9, fig.width = 10}
# plotting by % change from 2021 to 2023 
all_rates_geo_simple <- all_rates_geo %>%
  select(NAME, acs1yr_year, rate_div_mhi) %>%
  pivot_wider(., names_from = acs1yr_year, values_from = rate_div_mhi) %>%
  mutate(pct_change =  100*((`2023` - `2021`)/`2023`))
     
# plotting: 
pct_mhi_rate_pal <- colorBin(
  bins = c(-40, -20, -5, 0, 5, 20),
  palette = rev(c("#d73027", "#fc8d59", "white", "#d9ef8b", "#91cf60")),
  domain = all_rates_geo_simple$pct_change)

leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron, 
                   group = "Toner Lite") %>%
  addPolygons(data = tx_county_pop,
              # slight opacity
              fillOpacity = 0.2,
              color = "gray",
              weight = 1, 
              smoothFactor = 0) %>%
  addPolygons(data = all_rates_geo_simple,
              fillColor = ~pct_mhi_rate_pal(pct_change),
              # slight opacity
              fillOpacity = 0.8,
              stroke = T, 
              color = "black",
              opacity = 0.2, 
              weight = 1,
              # smoothFactor = 0.5, 
              label = paste0("County: ", all_rates_geo_simple$NAME, 
                             "; % change: ",
                             round(all_rates_geo_simple$pct_change, 3))) %>%
  addLegend("bottomright",
            pal = pct_mhi_rate_pal,
            values = all_rates_geo_simple$pct_change,
            # if working with money or percentages, use the 
            # follow label formats: 
            # labFormat = labelFormat(prefix = "$"),
            labFormat = labelFormat(prefix = "%"),
            title = "% Change in % of Income Spent on Water",
            opacity = 1)
```

When looking at the percentage in the % of income spent on water from 2021 - 2023, most counties in our study period decreased over time, indicating that the relative financial burden of water on annual income has decreased. There are a few counties where financial burden has increased, particularly in more rural areas in NE Texas. The largest % change in our study period was Tom Green County with an increase of 18.6%. 


# 4.0: What are examples of success stories? 
## 4.1: What are the trends [for this specific CWS] in terms of population, violations, water rates, and investments? 

Looking the county collection of pwsids that have received funding 
```{r fig.height = 9, fig.width = 10}
# grabbing demographic and socioeconomic trends - 2010, 2015, 2020
# head(tx_decade)
# # funding data - 2009, 2020
# head(tx_srf)
# violation data 
viol_summary <- tx_viol_rulecode %>%
  mutate(start_date = as.Date(viol_first_reported_date, 
                              tryFormats = c("%m/%d/%Y")), 
         start_year = year(start_date)) %>%
  filter(is_health_based_ind == "Y") %>%
  group_by(pwsid, start_year) %>%
  summarize(total_health_viols = n())

# ok let's just zoom in on one pwsid and look at trends over time -
# ideally one that has received funding around 2015ish 

# looking at pwsids in a particular county: 
county <- "Lubbock"
pwsids_county <- app_data_ws_info %>%
  filter(county_served == county)
# grabbing rates: 
pwsid_rates <- all_rates_calc %>% 
  filter(NAME ==  paste0(county, " County, Texas"))
# grabbing demographic info 
pwsid_demo <- tx_decade %>% 
  filter(pwsid %in% pwsids_county$pwsid)
# SRF funding history
pwsid_srf <- tx_srf %>% 
  filter(pwsid %in% pwsids_county$pwsid) %>%
  select(pwsid, year, assistance_amt, prin_forgive_amt) %>%
  group_by(pwsid, year) %>%
  summarize(assistance_amt = sum(assistance_amt), 
            prin_forgive_amt = sum(prin_forgive_amt))
# violation history: 
pwsid_viol <- viol_summary %>%
  filter(pwsid %in% pwsids_county$pwsid) %>%
  filter(start_year > 2005) %>%
  mutate(start_year = as.character(start_year)) 

# ok gotta reorg to see all of this information: 
pwsid_demo_long <- pwsid_demo %>%
  select(pwsid, estimate_total_pop_2010, estimate_mhi_2010_inf_adj, 
         estimate_total_pop_2015, estimate_mhi_2015_inf_adj, estimate_mhi_2020, 
         estimate_total_pop_2020) %>%
  pivot_longer(., cols = estimate_total_pop_2010:estimate_total_pop_2020) %>%
  mutate(year = case_when(
    grepl("2010", name) ~ "2010", 
    grepl("2015", name) ~ "2015", 
    TRUE ~ "2020"
  ), 
  demo = case_when(
    grepl("pop", name) ~ "total_pop", 
    TRUE ~ "mhi"
  )) %>%
  select(-name) %>%
  pivot_wider(., names_from = demo, values_from = value) 

# adding violations: 
pwsid_demo_viol <- merge(pwsid_demo_long, pwsid_viol, 
                         by.x = c("pwsid", "year"), 
                         by.y = c("pwsid", "start_year"), all = T)

# adding srf: 
pwsid_profile <- merge(pwsid_demo_viol, pwsid_srf,
                       by = c("pwsid", "year"), all = T) 

pwsid_profile_long <- pwsid_profile %>% 
  pivot_longer(., cols = total_pop:prin_forgive_amt) %>%
  # just looking at ones that have received funding 
  filter(pwsid %in% pwsid_srf$pwsid) %>%
  mutate(year = as.numeric(year))

# looking at multiple pwsids for a particualar county: 
pwsid_plot <- ggplot(pwsid_profile_long, aes(x = year, y = value, 
                               color = pwsid, group = pwsid)) + 
  geom_smooth(se = F)+ 
  geom_point() + 
  facet_wrap(~name, scales = "free_y", ncol = 1) + 
  epic_chart_theme + 
  ggtitle(paste0("Utilities Funded in ", county, " County")) + 
  theme_minimal() + 
  theme(legend.position = "right", 
        text = element_text(size = 13), 
        legend.text = element_text(size = 10), 
        legend.title = element_text(size = 11), 
        axis.text.x = element_text(margin = margin(t = 10, r = 0, 
                                                   b = 0, l = 0)), 
        axis.title.x = element_text(margin = margin(t = 10, r = 0, 
                                                    b = 0, l = 0)), 
        axis.text.y = element_text(margin = margin(t = 0, r = 10, 
                                                   b = 0, l = 0)), 
        axis.title.y = element_text(margin = margin(t = 0, r = 10, 
                                                    b = 0, l = 0))) +
  labs(x = "Year", y = "Value")
ggplotly(pwsid_plot)
```

This figure can be used for looking at overall trends for utilities serving a particular county. 

Zooming in on some pwsids: 
```{r fig.height = 9, fig.width = 10}
# looking at pwsids in a particular county: 
county <- "Tarrant"

pwsids_county <- app_data_ws_info %>%
  filter(county_served == county)
# grabbing rates: 
pwsid_rates <- all_rates_calc %>% 
  filter(NAME ==  paste0(county, " County, Texas"))
# grabbing demographic info 
pwsid_demo <- tx_decade %>% 
  filter(pwsid %in% pwsids_county$pwsid)
# SRF funding history
pwsid_srf <- tx_srf %>% 
  filter(pwsid %in% pwsids_county$pwsid) %>%
  select(pwsid, year, assistance_amt, prin_forgive_amt) %>%
  group_by(pwsid, year) %>%
  summarize(assistance_amt = sum(assistance_amt), 
            prin_forgive_amt = sum(prin_forgive_amt))
# violation history: 
pwsid_viol <- viol_summary %>%
  filter(pwsid %in% pwsids_county$pwsid) %>%
  filter(start_year > 2005) %>%
  mutate(start_year = as.character(start_year)) 

# ok gotta reorg to see all of this information: 
pwsid_demo_long <- pwsid_demo %>%
  select(pwsid, estimate_total_pop_2010, estimate_mhi_2010_inf_adj, 
         estimate_total_pop_2015, estimate_mhi_2015_inf_adj, estimate_mhi_2020, 
         estimate_total_pop_2020) %>%
  pivot_longer(., cols = estimate_total_pop_2010:estimate_total_pop_2020) %>%
  mutate(year = case_when(
    grepl("2010", name) ~ "2010", 
    grepl("2015", name) ~ "2015", 
    TRUE ~ "2020"
  ), 
  demo = case_when(
    grepl("pop", name) ~ "total_pop", 
    TRUE ~ "mhi"
  )) %>%
  select(-name) %>%
  pivot_wider(., names_from = demo, values_from = value) 

# adding violations: 
pwsid_demo_viol <- merge(pwsid_demo_long, pwsid_viol, 
                         by.x = c("pwsid", "year"), 
                         by.y = c("pwsid", "start_year"), all = T)

# adding srf: 
pwsid_profile <- merge(pwsid_demo_viol, pwsid_srf,
                       by = c("pwsid", "year"), all = T) 

# long format for plotting: 
pwsid_profile_long <- pwsid_profile %>% 
  pivot_longer(., cols = total_pop:prin_forgive_amt) %>%
  # just looking at ones that have received funding 
  filter(pwsid %in% pwsid_srf$pwsid) %>%
  mutate(year = as.numeric(year))

# zooming in a particular pwsid: 
pwsid_i <- "TX2200069"

pwsid_focus <- pwsid_profile_long %>% 
  filter(pwsid == pwsid_i) %>%
  filter(!(name %in% c("assistance_amt", "prin_forgive_amt")))
pwsid_srf_i <- pwsid_srf %>% 
  filter(pwsid == pwsid_i)

ggplot(pwsid_focus, aes(x = year, y = value, group = pwsid)) + 
  geom_smooth(se = F)+ 
  geom_point() + 
  geom_vline(xintercept = unique(pwsid_srf_i$year),
             color = "red",
             size=1, 
             linetype = "dashed") + 
  facet_wrap(~name, scales = "free_y", ncol = 1) + 
  epic_chart_theme + 
  ggtitle(paste0("Trends for ", pwsid_i, ", serving ", county, " County"))
```

Red dashed line indicates the year in which the utility received funding. For this particular utility, violations dipped after the utility received funding, populations continued to increase, and MHI started increasing since 2015. Utilities serving Tarrant county have also exhibited a decrease in the % of income spent on MHI from 2021-2023, indicating either lower rates or higher MHI over that time period. 

```{r fig.height = 9, fig.width = 10}
county <- "Lubbock"
pwsid_i <- "TX1520067"

pwsids_county <- app_data_ws_info %>%
  filter(county_served == county)
# grabbing rates: 
pwsid_rates <- all_rates_calc %>% 
  filter(NAME ==  paste0(county, " County, Texas"))
# grabbing demographic info 
pwsid_demo <- tx_decade %>% 
  filter(pwsid %in% pwsids_county$pwsid)
# SRF funding history
pwsid_srf <- tx_srf %>% 
  filter(pwsid %in% pwsids_county$pwsid) %>%
  select(pwsid, year, assistance_amt, prin_forgive_amt) %>%
  group_by(pwsid, year) %>%
  summarize(assistance_amt = sum(assistance_amt), 
            prin_forgive_amt = sum(prin_forgive_amt))
# violation history: 
pwsid_viol <- viol_summary %>%
  filter(pwsid %in% pwsids_county$pwsid) %>%
  filter(start_year > 2005) %>%
  mutate(start_year = as.character(start_year)) 

# ok gotta reorg to see all of this information: 
pwsid_demo_long <- pwsid_demo %>%
  select(pwsid, estimate_total_pop_2010, estimate_mhi_2010_inf_adj, 
         estimate_total_pop_2015, estimate_mhi_2015_inf_adj, estimate_mhi_2020, 
         estimate_total_pop_2020) %>%
  pivot_longer(., cols = estimate_total_pop_2010:estimate_total_pop_2020) %>%
  mutate(year = case_when(
    grepl("2010", name) ~ "2010", 
    grepl("2015", name) ~ "2015", 
    TRUE ~ "2020"
  ), 
  demo = case_when(
    grepl("pop", name) ~ "total_pop", 
    TRUE ~ "mhi"
  )) %>%
  select(-name) %>%
  pivot_wider(., names_from = demo, values_from = value) 

# adding violations: 
pwsid_demo_viol <- merge(pwsid_demo_long, pwsid_viol, 
                         by.x = c("pwsid", "year"), 
                         by.y = c("pwsid", "start_year"), all = T)

# adding srf: 
pwsid_profile <- merge(pwsid_demo_viol, pwsid_srf,
                       by = c("pwsid", "year"), all = T) 

# long format for plotting: 
pwsid_profile_long <- pwsid_profile %>% 
  pivot_longer(., cols = total_pop:prin_forgive_amt) %>%
  # just looking at ones that have received funding 
  filter(pwsid %in% pwsid_srf$pwsid) %>%
  mutate(year = as.numeric(year))

# zooming in a particular pwsid: 
pwsid_focus <- pwsid_profile_long %>% 
  filter(pwsid == pwsid_i) %>%
  filter(!(name %in% c("assistance_amt", "prin_forgive_amt")))
pwsid_srf_i <- pwsid_srf %>% 
  filter(pwsid == pwsid_i)

ggplot(pwsid_focus, aes(x = year, y = value, group = pwsid)) + 
  geom_smooth(se = F)+ 
  geom_point() + 
  geom_vline(xintercept = unique(pwsid_srf_i$year),
             color = "red",
             size=1, 
             linetype = "dashed") + 
  facet_wrap(~name, scales = "free_y", ncol = 1) + 
  epic_chart_theme + 
  ggtitle(paste0("Trends for ", pwsid_i, ", serving ", county, " County"))
```

Red dashed line indicates the year in which the utility received funding. For this particular utility, violations dipped after the utility received funding, and MHI and populations continued to increase. Utilities serving Lubbock county have exhibited an increase in the % of income spent on MHI from 2021-2023, indicating either higher rates or lower MHI over that time period. 


# 5.0 What are examples of cautionary tales?
## 5.1: What are the trends [for this specific CWS] in terms of population, violations, water rates, and investments? 
```{r fig.height = 9, fig.width = 10}
# zooming in on utilities that have been declining in pop and MHI 
# and counties where financial burden of water have been increasing: 
pwsid_caution <- cws_of_interest %>% 
  filter(Category %in% c("Pop & MHI Decreasing", 
                         "Pop Growing, MHI Decreasing")) %>% 
  mutate(county_name = paste0(county_served, " County, Texas")) 

# which ones are serving counties where financial burden is increasing?
county_caution <- all_rates_geo_simple %>% 
  filter(NAME %in% pwsid_caution$county_name) %>%
  filter(pct_change > 0)

# these utilites not only have declining populations and mhi, but financial 
# burden of water is increasing; 
pwsid_decdemo_rates <- pwsid_caution %>%
  filter(county_name %in% county_caution$NAME)

# grabbing violations for these pwsids: 
pwsid_viol_caution <- viol_summary %>%
  filter(pwsid %in% pwsid_decdemo_rates$pwsid) %>%
  group_by(pwsid, start_year) %>%
  filter(start_year > 2005) %>%
  summarize(total_health_viols = sum(total_health_viols))

# SRF funding history
pwsid_srf_caution <- tx_srf %>% 
  filter(pwsid %in% pwsid_decdemo_rates$pwsid) %>%
  select(pwsid, year, assistance_amt, prin_forgive_amt) %>%
  group_by(pwsid, year) %>%
  summarize(assistance_amt = sum(assistance_amt), 
            prin_forgive_amt = sum(prin_forgive_amt))


# what has funding looked like?
pwsid_demo_long_c <- pwsid_decdemo_rates %>%
  select(pwsid, county_name, Category, 
         estimate_total_pop_2010, estimate_mhi_2010_inf_adj, 
         estimate_total_pop_2015, estimate_mhi_2015_inf_adj, estimate_mhi_2020, 
         estimate_total_pop_2020) %>%
  pivot_longer(., cols = estimate_total_pop_2010:estimate_total_pop_2020) %>%
  mutate(year = case_when(
    grepl("2010", name) ~ "2010", 
    grepl("2015", name) ~ "2015", 
    TRUE ~ "2020"
  ), 
  demo = case_when(
    grepl("pop", name) ~ "total_pop", 
    TRUE ~ "mhi"
  )) %>%
  select(-name) %>%
  pivot_wider(., names_from = demo, values_from = value) 

# adding violations: 
pwsid_demo_viol_c <- merge(pwsid_demo_long_c, pwsid_viol_caution, 
                         by.x = c("pwsid", "year"), 
                         by.y = c("pwsid", "start_year"), all = T)

# adding srf: 
pwsid_profile_c <- merge(pwsid_demo_viol_c, pwsid_srf_caution,
                       by = c("pwsid", "year"), all = T) %>%
  as.data.frame() %>%
  select(-starts_with("geom")) %>% 
  mutate(year = as.numeric(year))

pwsid_profile_c_long <- pivot_longer(pwsid_profile_c, 
                                     cols = total_pop:prin_forgive_amt)

# plottin'
of_caution <- ggplot(pwsid_profile_c_long, aes(x = year, y = value, 
                                 color = pwsid,
                                 group = pwsid)) + 
  geom_smooth(se = F)+ 
  geom_point() + 
  # geom_vline(xintercept = unique(pwsid_srf_i$year),
  #            color = "red",
  #            size=1, 
  #            linetype = "dashed") + 
  facet_wrap(~name, scales = "free_y") + 
  epic_chart_theme + 
  ggtitle(paste0("Trends for Utilities in Possible Need for Support")) + 
  theme_minimal() + 
  theme(legend.position = "right", 
        text = element_text(size = 13), 
        legend.text = element_text(size = 10), 
        legend.title = element_text(size = 11), 
        axis.text.x = element_text(margin = margin(t = 10, r = 0, 
                                                   b = 0, l = 0)), 
        axis.title.x = element_text(margin = margin(t = 10, r = 0, 
                                                    b = 0, l = 0)), 
        axis.text.y = element_text(margin = margin(t = 0, r = 10, 
                                                   b = 0, l = 0)), 
        axis.title.y = element_text(margin = margin(t = 0, r = 10, 
                                                    b = 0, l = 0))) +
  labs(x = "Year", y = "Value")

ggplotly(of_caution)
```

Chart above shows utilies where populations and/or MHI have been decreasing from 2010 - 2020, and serve a county in which the relative annual financial burden of water has increased from 2021 - 2023 (this of course limits our universe to only counties where we had rate data). Of utilities that meet these criteria, only 6 utilities have recieved funding. 
