---
title: "Texas Drinking Water Data Collection"
author: "EmmaLi Tsai"
date: "2024-01-29"
output: html_document
---

Packages: 
```{r packages}
library(tidyverse)
library(leaflet)
library(sf)
library(aws.s3) 
library(janitor)
library(tidycensus)
library(areal)
library(mapview)
library(httr)
library(readxl)
library(jsonlite)
library(googlesheets4)
# remotes::install_github("environmental-policy-innovation-center/stateDW", 
#                         auth_token = "insert-github-pat-here")
library(stateDW)

# Store keys on local machine in .Renviron file! 
```

Functions: 
```{r functions}
# Migrated to R package workflow but keeping these here as a backup 
# source("./code/functions/get_sab.R")
# source("./code/functions/get_SDWIS.R")
# source("./code/functions/interp_sab_census.R")
```

_______________________________________________________________________________
### Data Download - code for web scraping & saving raw files. Data here either take a long time to download, require extra tidying, are updated frequently (i.e., reservoir or aquifer levels) or need to be manually downloaded from a website. Everything in this section is added to S3.
_______________________________________________________________________________
## Data Download: Demographic
```{r data collection}
# SOP for TX-specific data: ####################################################
# PWSIDs should have "TX" at the beginning of them 
# All data must have either pwsid, aquifer, census id (block group/tract), and well number (or any match-able 
# columns)
# Group data collections by category (demographic, environmental, financial, etc. )
# All data should be in wide format 
# All data should have a pulled link to note where it came from
# Add data to the TX_data list 
# Variable naming convention - janitor::clean_names()

## Demographic #################################################################
# grabbing sabs: 
sab <- get_sab(states = c("TX"), crs = ("ESRI:102296"))
# making note the of the pwsids located in TX: 
pwsids <- unique(sab$pwsid)

# census data - grabbing block group stats for now 
# vars_acs <- load_variables(year = 2021, "acs5")
census_vars <- c(total_pop = "B01003_001",
                 black_alone = "B02001_003", 
                 asian_alone = "B02001_005", 
                 white_alone = "B02001_002", 
                 AIAN_alone = "B02001_004", 
                 NAPI_alone = "B02001_006", 
                 other_alone = "B02001_007", 
                 mixed_alone = "B02001_008",
                 hisp_alone = "B03003_003", 
                 mhi = "B19013_001", 
                 under_6  = "B23008_002", 
                 over_65 = "B09021_022", 
                 only_english = "B99162_002", 
                 other_lang = "B99162_003", 
                 no_school = "B15003_002", 
                 prof_degree = "B15003_024", 
                 foreign = "B99051_005")
census <- tidycensus::get_acs(
  geography = "block group", 
  variables = census_vars, 
  state = unique(sab$state_code), 
  year = 2021,
  geometry = TRUE
)

# crosswalking census stats with pwsids: 
sab_census <- interp_sab_census(census_df = census, sab_df = sab, 
                                total_pop_var = "estimate_total_pop", 
                                white_alone_var = "estimate_white_alone")
# SAB & Census:
tmp <- tempfile()
st_write(sab_census, dsn = paste0(tmp, ".geojson"))
on.exit(unlink(tmp))
put_object(
  file = paste0(tmp, ".geojson"),
  object = "/state-drinking-water/TX/raw/TX_sab_census.geojson",
  bucket = "tech-team-data",
)
```
## Data Download: Environmental
```{r}
## Environmental ###############################################################
## Groundwater database: https://www.twdb.texas.gov/groundwater/data/GWDBDownload.zip
# Pulled Jan 22 2024 
file_loc <- "./data/raw/TX_GWDB"
download.file("https://www.twdb.texas.gov/groundwater/data/GWDBDownload.zip", 
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
file.remove(paste0(file_loc, ".zip"))
## TODO: add code that downloads & uploads entire folder to aws
# put_folder(
#   folder = "./data/raw/TX_GWDB/GWDBDownload/",
#   bucket = "s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB_test"
# )

## Reservoir levels: https://www.waterdatafortexas.org/reservoirs/statewide
# downloaded Jan 24 2024 
# NOTE: only recent conditions - update more frequently?
res_levels <- read.csv("http://www.waterdatafortexas.org/reservoirs/recent-conditions.csv")
# writing to aws: 
tmp <- tempfile()
write.csv(res_levels, file = paste0(tmp, ".csv"))
on.exit(unlink(tmp))
put_object(
  file = paste0(tmp, ".csv"),
  object = "/state-drinking-water/TX/raw/TX_reservoir_levels_recent.csv",
  bucket = "tech-team-data",
)

# grabbing historic reservoir levels: 
tx_res <- c("brazos", "canadian", "colorado", "cypress", "guadalupe", "lavaca", 
            "neches", "nueces", "red", "rio-grande", "sabine", "san-antonio", "san-jacinto", 
            "sulphur", "trinity")
# grabbing historic reservoir levels: 
res_data <- data.frame()
for(i in 1:length(tx_res)){
  res_i <- tx_res[i]
  print(res_i)
  download_link <- paste0("https://www.waterdatafortexas.org/reservoirs/basin/", res_i, ".csv")
  res_data_i <- read.csv(download_link, skip = 29) 
  res_data_i$reservoir <- res_i
  res_data <- rbind(res_data, res_data_i)
}
# filtering for the past 10 years, since some records go back to 1938
res_data$date <- as.Date(res_data$date, tryFormats = "%Y-%m-%d")
res_data$year <- year(res_data$date)
res_data_filtered <- res_data %>%
  filter(year >= 2013)
# writing to aws: 
tmp <- tempfile()
write.csv(res_data_filtered, file = paste0(tmp, ".csv"))
on.exit(unlink(tmp))
put_object(
  file = paste0(tmp, ".csv"),
  object = "/state-drinking-water/TX/raw/TX_reservoir_levels_10yr.csv",
  bucket = "tech-team-data",
)

## Groundwater levels: https://www.waterdatafortexas.org/groundwater
# downloaded Jan 24 2024 
# NOTE: only recent conditions - update more frequently?
well_levels <- read.csv("https://www.waterdatafortexas.org/groundwater/recent-conditions.csv") %>%
  janitor::clean_names()
# writing to aws: 
tmp <- tempfile()
write.csv(well_levels, file = paste0(tmp, ".csv"))
on.exit(unlink(tmp))
put_object(
  file = paste0(tmp, ".csv"),
  object = "/state-drinking-water/TX/raw/TX_recent_well_levels.csv",
  bucket = "tech-team-data",
)
```
## Data Download: Water Delivery System 
```{r}
## Water Delivery System #######################################################
# grabbing SDWIS data - currently looking at all of them by leaving cont_code
# as blank
sdwis <- get_SDWIS(pwsids = pwsids, 
                   cont_code = c())
tmp <- tempfile()
write.csv(sdwis, file = tmp)
on.exit(unlink(tmp))
put_object(
  file = tmp,
  object = "/state-drinking-water/TX/raw/TX_sdwis.csv",
  bucket = "tech-team-data",
)
```
## Data Download: Financial
```{r}
# TX DAC - whether they are likely to qualify for DAC status: 
## Water rates: https://raw.githubusercontent.com/NIEPS-Water-Program/water-affordability/main/www/data/all_rates_table_current.csv
# downloaded Jan 31 2024
raw_link <- "https://raw.githubusercontent.com/NIEPS-Water-Program/water-affordability/main/www/data/all_rates_table_current.csv"
temp_file <- tempfile(fileext = ".csv")
water_rates <- GET(raw_link, write_disk(path = temp_file))
water_rates <- read.csv(temp_file) %>%
  janitor::clean_names()
# filtering for just TX data, and household use == 6000 
tx_rates <- water_rates[grepl("TX", water_rates$pwsid),] %>%
  filter(hh_use == 6000) %>%
  select(c("pwsid", "category", "hh_use", "total_water", "total_sewer")) %>%
  mutate(total_water_year = total_water*12, 
         total_sewer_year = total_sewer*12)
```

## Data Download: Requires manually downloading the data 
```{r}
# Climate Vulnerability Index: https://map.climatevulnerabilityindex.org/map/cvi_overall/usa?mapBoundaries=Tract&mapFilter=0&reportBoundaries=Tract&geoContext=State
# manually downloaded from dropbox on Jan 26 2024 (there is no api) and 
# added to my ./data/raw folder: 
cvi <- readxl::read_excel("./data/raw/MasterCVIDataset_Oct2023.xlsx", 
           sheet = "Domain CVI Values") %>%
  filter(State == "TX") %>%
  janitor::clean_names()
# writing to aws: 
tmp <- tempfile()
write.csv(cvi, file = paste0(tmp, ".csv"))
on.exit(unlink(tmp))
put_object(
  file = paste0(tmp, ".csv"),
  object = "/state-drinking-water/TX/raw/TX_CVI.csv",
  bucket = "tech-team-data",
)

## Historic use estimates: https://www.twdb.texas.gov/waterplanning/waterusesurvey/estimates/index.asp
# downloaded Jan 24 2024
# Without an api or data download link, I manually downloaded data from 
# 2016 - 2020 from here: https://www3.twdb.texas.gov/apps/reports/WU_REP/PWS_Categorical_Connections_and_Volumes
# and dropped it in my local ./data/raw folder: 
folder <- "./data/raw/TX_historic_use_estimates/"
files <- list.files(folder)
year_use_estimates <- lapply(paste0(folder, files), read.csv, header = TRUE)
historic_use <- do.call(rbind, year_use_estimates) %>%
  janitor::clean_names()
# writing to aws: 
tmp <- tempfile()
write.csv(historic_use, file = paste0(tmp, ".csv"))
on.exit(unlink(tmp))
put_object(
  file = paste0(tmp, ".csv"),
  object = "/state-drinking-water/TX/raw/TX_historic_use_estimates.csv",
  bucket = "tech-team-data",
)

## Water loss audits: https://www.twdb.texas.gov/conservation/municipal/waterloss/historical-annual-report.asp
# downloaded Jan 24 2024
# This requires merging in downloaded file that is stored in s3 as 
# TX_historic_use_estimates.csv, looping over to cleaning and then merge 
```

_______________________________________________________________________________
### Data Collating - pulling data from APIs, web links, or S3 buckets:  Data here are either pulled from S3 (see code above for how they got there) or downloaded easily from websites or APIs
_______________________________________________________________________________
## Data Collating: Demographic
```{r data collating}
## Demographic #################################################################
# SABs
sab_census <- aws.s3::s3read_using(st_read, 
                           object = "state-drinking-water/TX/raw/TX_sab_census.geojson",
                           bucket = "tech-team-data")
# CEJST
cejst_raw <- aws.s3::s3read_using(read.csv, object = "s3://tech-team-data/ej-data/cejest-v1/1.0-communities.csv") %>%
  janitor::clean_names() %>%
  filter(state_territory == "Texas") %>%
  select(c("census_tract_2010_id","total_threshold_criteria_exceeded",
           "total_categories_exceeded","identified_as_disadvantaged",
           "percentage_of_tract_that_is_disadvantaged_by_area","total_population"))%>%
  rename("total_population_cejst" = "total_population")

# Social Vulnerability Index 
svi <- aws.s3::s3read_using(read.csv, 
                            object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_SVI_2020.csv")

# Climate Vulnerability Index: 
cvi <- aws.s3::s3read_using(read.csv, 
                            object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_CVI.csv")
```
## Data Collating: Environmental
```{r}
## Environmental ###############################################################
## Surface water intake locations: https://gis-tceq.opendata.arcgis.com/datasets/TCEQ::surface-intakes/api
# Pulled Jan 22 2024 
sw_intake <- GET("https://gisweb.tceq.texas.gov/arcgis/rest/services/Public/PWS/MapServer/1/query?where=1%3D1&outFields=*&outSR=4326&f=json")
sw_intake_tidy <- fromJSON(rawToChar(sw_intake$content))$features$attribute %>%
  as.data.frame() %>%
  janitor::clean_names() %>%
  rename(pwsid = pws_id) %>%
  mutate(pwsid = paste0("TX", pwsid))

## groundwater quality & levels: 
# function for reading gwdb tables, which have unique separators: 
read_gwdb_table <- function(x) {
     read.table(x, header = TRUE, fill = TRUE, sep = "|")
}
# grabbing well metadata: 
well_main <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WellMain.txt") %>%
  janitor::clean_names()

# water quality information: 
wq_major <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WaterQualityMajor.txt") %>%
  janitor::clean_names()
wq_minor <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WaterQualityMinor.txt") %>%
  janitor::clean_names()
wq_other <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WaterQualityOtherUnassigned.txt") %>%
  janitor::clean_names()
# let's just combine all of these 
well_wq <- rbind(wq_major, wq_minor, wq_other)

# water level information: 
wl_major <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WaterLevelsMajor.txt") %>%
  janitor::clean_names()
wl_minor <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WaterLevelsMinor.txt") %>%
  janitor::clean_names()
wl_other <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WaterLevelsOtherUnassigned.txt") %>%
  janitor::clean_names()
# let's just combine all of these 
well_wl <- rbind(wl_major, wl_minor, wl_other)

## Major aquifers: https://www.twdb.texas.gov/mapping/gisdata/doc/major_aquifers.zip
# downloaded 1/23/2024
file_loc <- "./data/raw/TX_major_aquifers"
download.file("https://www.twdb.texas.gov/mapping/gisdata/doc/major_aquifers.zip", 
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
file.remove(paste0(file_loc, ".zip"))
maj_aquifers <- st_read(paste0(file_loc, "/NEW_major_aquifers_dd.shp")) %>%
  janitor::clean_names()

## Major aquifers: https://www.twdb.texas.gov/mapping/gisdata/doc/major_aquifers.zip
# downloaded 1/23/2024
file_loc <- "./data/raw/TX_major_aquifers"
download.file("https://www.twdb.texas.gov/mapping/gisdata/doc/major_aquifers.zip", 
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
file.remove(paste0(file_loc, ".zip"))
maj_aquifers <- st_read(paste0(file_loc, "/NEW_major_aquifers_dd.shp")) %>%
  janitor::clean_names()

## Minor aquifers: https://www.twdb.texas.gov/mapping/gisdata/doc/minor_aquifers.zip
# downloaded 1/23/2024
file_loc <- "./data/raw/TX_minor_aquifers"
download.file("https://www.twdb.texas.gov/mapping/gisdata/doc/minor_aquifers.zip", 
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
file.remove(paste0(file_loc, ".zip"))
minor_aquifers <- st_read(paste0(file_loc, "/Minor_Aquifers.shp")) %>%
  janitor::clean_names()

## Major river basins: https://www.twdb.texas.gov/mapping/gisdata/doc/Major_River_Basins_Shapefile.zip
# downloaded 1/23/2024
file_loc <- "./data/raw/TX_major_river_basins"
download.file("https://www.twdb.texas.gov/mapping/gisdata/doc/Major_River_Basins_Shapefile.zip", 
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
file.remove(paste0(file_loc, ".zip"))
major_river_basins <- st_read(paste0(file_loc, "/TWDB_MRBs_2014.shp")) %>%
  janitor::clean_names()

## Existing reservoirs: https://www.twdb.texas.gov/mapping/gisdata/doc/Existing_Reservoirs.zip
# downloaded 1/23/2024
file_loc <- "./data/raw/TX_reservoirs"
download.file("https://www.twdb.texas.gov/mapping/gisdata/doc/Existing_Reservoirs.zip", 
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
file.remove(paste0(file_loc, ".zip"))
reservoirs <- st_read(paste0(file_loc, "/TWDB_SWP2012_Major_Reservoirs.shp")) %>%
  janitor::clean_names()

## Surface water 303(d) river basin impairment: https://texaswaterexplorer.tnc.org/index.html
# downloaded 1/24/2024
file_loc <- "./data/raw/TX_basin_impairment_2012"
download.file("https://texaswaterexplorer.tnc.org/downloads/Data_WQImpairment.zip", 
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
file.remove(paste0(file_loc, ".zip"))
basin_impairment <- readxl::read_excel(paste0(file_loc, "/Data_WQImpairment.xlsx"), sheet = "Basins")%>%
  janitor::clean_names()

# reservoir levels - recent
recent_reservoir_levels <- aws.s3::s3read_using(read.csv, 
                                         object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_reservoir_levels_recent.csv")

# reservoir levels - historic
reservoir_levels_10yr <- aws.s3::s3read_using(read.csv, 
                                         object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_reservoir_levels_10yr.csv")

# groundwater levels
recent_well_levels <- aws.s3::s3read_using(read.csv, 
                                           object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_recent_well_levels.csv")

# TCEQ DWW system data: 
# NOTE: there is A TON more information in s3 for us to pull like enforcement 
# actions, etc. but starting the structure here for grabbing that information
# pulling data from two different webscraping events and locating pwsids that 
# are community water systems: 
water_details_dec17 <- aws.s3::s3read_using(read.csv, 
                                      object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_TCEQ_DWW/TX_20231217/WaterSystemDetail/TX_Water System Details_20231217_20231229203427.csv") 
water_details_dec3rd <- aws.s3::s3read_using(read.csv, 
                                      object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_TCEQ_DWW/TX_20231203/WaterSystemDetail/TX_Water System Details_20231203.csv") 
combined_details <- rbind(water_details_dec17, water_details_dec3rd)
water_details <- pivot_wider(combined_details, names_from = "Type", values_from = "Value") %>%
  janitor::clean_names()
# grabbing just community water systems (tceq has 10k that are not cws): 
tceq_cws <- water_details %>%
  filter(system_type == "C") # 4666

# pulling violations: 
viol_dec17 <- aws.s3::s3read_using(read.csv, 
                                      object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_TCEQ_DWW/TX_20231217/Violations/TX_Individual Violations_20231217_20231229203427.csv") 
viol_dec3rd <- aws.s3::s3read_using(read.csv, 
                                      object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_TCEQ_DWW/TX_20231203/Violations/TX_Individual Violations_20231203.csv") 
comined_viols <- rbind(viol_dec17, viol_dec3rd) %>%
  janitor::clean_names() %>%
  filter(water_system_no %in% tceq_cws$water_system_no) # 118,659
# SDWIS contains 20,000 more violations  
```
## Data Collating: Water Delivery System
```{r}
## Water Delivery System ######################################################
# historic use estimates: 
historic_use_estimates <-  aws.s3::s3read_using(read.csv, 
                                                object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_historic_use_estimates.csv")

# water loss audits
# vector of cleaned column names 
waterlosscolnames <- c("Name of Utility",
                       "Real Loss GMD (<32 conn/mi)",
                       "Real Loss GCD",
                       "Apparent Loss GCD",
                       "Water Loss GCD",
                       "ILI (>= 3,000 connections)", 
                       "Total GPCD", 
                       "GPCD Loss",
                       "Real Loss Cost in dollars",
                       "Apparent Loss Cost in dollars","Year")

#making empty dataframe 
waterloss_all <- setNames(data.frame(matrix(ncol = length(waterlosscolnames), nrow = 0)), waterlosscolnames)

year <- 2010 

# ! edit this when adding additional years 
stopyear <- 2022 

while(year <= stopyear)
{
print(year)
waterloss_raw <- aws.s3::s3read_using(read_csv, 
                               object = paste0("s3://tech-team-data/state-drinking-water/TX/raw/TCEQ_Loss_Audits/SummaryAuditsByCategory_",year,".csv"))

waterloss <- waterloss_raw %>%
             select(1:10)

colnames(waterloss) <- waterlosscolnames

waterloss <- waterloss %>%
             janitor::clean_names()%>%
             filter(name_of_utility != "UtilityName")%>%
             mutate(Year = year)

waterloss_all <- rbind(waterloss_all,waterloss)
year <- year + 1
}

## 2022 TX State Water Plan: https://2022.texasstatewaterplan.org/statewide?
# downloaded Jan 29 2024
pop <- read.csv("https://2022.texasstatewaterplan.org/download/statewide/population") %>%
  janitor::clean_names()
demands <- read.csv("https://2022.texasstatewaterplan.org/download/statewide/demands") %>%
  janitor::clean_names()
supp <- read.csv("https://2022.texasstatewaterplan.org/download/statewide/supplies") %>%
  janitor::clean_names()
needs <- read.csv("https://2022.texasstatewaterplan.org/download/statewide/needs") %>%
  janitor::clean_names()
strat <- read.csv("https://2022.texasstatewaterplan.org/download/statewide/strategies") %>%
  janitor::clean_names()
```
## Data Collating: Financial 
```{r}
## Financial ##################################################################
## Water affordability: https://github.com/NIEPS-Water-Program/water-affordability/raw/main/data/rates_data/rates_tx.xlsx
# downloaded Jan 24 2024
link <- "https://github.com/NIEPS-Water-Program/water-affordability/raw/main/data/rates_data/rates_tx.xlsx"
temp_file <- tempfile(fileext = ".xlsx")
water_aff <- GET(link, write_disk(path = temp_file))
water_aff <- readxl::read_excel(temp_file) %>%
  janitor::clean_names()

# PPL data & tidying 
TX_PPL <- aws.s3::s3read_using(read.csv, 
                               object = "s3://water-team-data/clean_data/srf_project_priority_lists/web_ppl_combined_clean_v1-1.csv") %>%
  mutate(across('Project.Type', 
                str_replace, 'Other', 'General')) %>%
  filter(State == "Texas") %>%
  janitor::clean_names()

## Danielle's SRF analysis: 
gs4_deauth()
URL <- "https://docs.google.com/spreadsheets/d/1kUHLdzFjwmKx4r6YTu2qjxwNuuG5JeUi/edit#gid=1618705547"
non_applying_cities <- read_sheet(URL, sheet = "Analysis")
```
_______________________________________________________________________________
### Organizing data into a list:
_______________________________________________________________________________
```{r}
# TX_data <- list()
# TX_data[[1]] <- sab_census
# TX_data[[2]] <- sdwis_tidy
# names(TX_data) <- c("census", "sdwis")
# ... 
# # here's how you get the list entry out and into a data frame: 
# sdwis <- do.call(data.frame, TX_data[[2]])
```





