---
title: "TX Drinking Water Data Analysis - Hydrology"
author: "EmmaLi Tsai"
date: "2024-02-23"
output: html_document
---
## Loading packages 
```{r}
library(tidyverse)
library(leaflet)
library(sf)
library(aws.s3)
```

## Loading data lists 
```{r}
# reading demographic, environmental (gw and sw), water delivery system (wds)
# and financial lists generated in TX_dw_collating: 
demo <- aws.s3::s3read_using(readRDS, 
                            object = "s3://tech-team-data/state-drinking-water/TX/clean/TX_demographic_list.RData")
enviro_sw <- aws.s3::s3read_using(readRDS, 
                            object = "s3://tech-team-data/state-drinking-water/TX/clean/TX_enviro_surface_list.RData")
enviro_gw <- aws.s3::s3read_using(readRDS, 
                            object = "s3://tech-team-data/state-drinking-water/TX/clean/TX_enviro_ground_list.RData")
wds <- aws.s3::s3read_using(readRDS, 
                            object = "s3://tech-team-data/state-drinking-water/TX/clean/TX_water_delivery_list.RData")
fin <- aws.s3::s3read_using(readRDS, 
                            object = "s3://tech-team-data/state-drinking-water/TX/clean/TX_financial_list.RData")

keys <- aws.s3::s3read_using(readRDS, 
                            object = "s3://tech-team-data/state-drinking-water/TX/clean/TX_merging_keys_list.RData")

crosswalk <- aws.s3::s3read_using(read.csv, 
                                  object = "s3://tech-team-data/pws_crosswalk/pws_census_tract_weighted_crosswalk.csv")%>%
  # if the geoid is missing prefix 0, add it back in
  mutate(tract_geoid = as.character(tract_geoid),
         tract_geoid = case_when(
           str_length(tract_geoid) == 10 ~ paste0(0, tract_geoid),
           TRUE ~ tract_geoid
         )) %>%
  # some tract geoids in the crosswalk are NAs - assuming these are errors 
  filter(!(is.na(tract_geoid)))
```

## Does the system have adequate water supply?
```{r}
# NOTE - there MIGHT be some pwsid & names we can link together based on 
# entity names, but would require some fuzzy name matching (again)

# TODO: make a analysis key for county-level stats for comparisons with county 
# MHI, % POC, etc. 

# comprehensive report here: https://www.twdb.texas.gov/waterplanning/swp/2022/docs/SWP22-Water-For-Texas.pdf?d=126542.70000004768

# projected values to 2070 - values are in acre-feet/year
proj_pop <- wds$TXSWP_pop
proj_dem <- wds$TXSWP_demands
proj_supp <- wds$TXSWP_supp
# needs also = potential shortages
proj_needs <- wds$TXSWP_needs
# not entirely sure how to incorporate strategies just yet 
proj_strat <- wds$TXSWP_strat

# checking out projected population change for each county: 
pop <- proj_pop %>%
  select(wug_county, wug_type, p2020:p2070) %>% 
  filter(wug_type == "MUNICIPAL") %>%
  group_by(wug_county) %>%
  summarize(p2020 = sum(p2020),
            p2030 = sum(p2030), 
            p2040 = sum(p2040), 
            p2050 = sum(p2050), 
            p2060 = sum(p2060), 
            p2070 = sum(p2070)) %>%
  mutate(pop_change = p2070 - p2020)
# gotta map this - grabbing census county geographies
census <- tidycensus::get_acs(
  geography = "county", 
  variables = c(total_pop = "B01003_001"), 
  state = c("TX"), 
  year = 2021,
  geometry = TRUE
)
# adding counties: 
census_tidy <- census
counties <- unlist(strsplit(census_tidy$NAME, split = ","))
census_tidy$counties <- trimws(counties[grepl("County", counties)])
counties <- census_tidy %>% 
  group_by(counties) %>% 
  mutate(county_geo = st_union(geometry)) %>%
  select(counties) %>%
  mutate(counties = toupper(counties), 
         counties = gsub(" COUNTY", "", counties)) %>%
  rename(wug_county = counties)

# combining with pop trends: 
pop_sf <- pop %>%
  left_join(counties) %>%
  st_as_sf() %>%
  st_transform(., st_crs(census))

# plotting to see projected population change by county: 
pop_pal <- colorBin(
  palette = viridis::viridis(10),
  bins = c(0, 100, 500, 1000, 50000, 100000, 500000, 1000000, 1564476))

leaflet() %>%
  addProviderTiles(providers$CartoDB.VoyagerNoLabels,
                   group = "Toner Lite") %>%
  addPolygons(data = pop_sf,
              opacity = 9,
              color = ~pop_pal(pop_change),
              weight = 1,
              label = paste0("county: ", pop_sf$wug_county,
                "; pop change: ", pop_sf$pop_change)) %>%
  addLegend("bottomright",
            pal = pop_pal,
            values = c(0, 100, 500, 1000, 50000, 100000, 500000, 1000000, 1564476),
            title = "pop change",
            opacity = 1)

# Combining demand, supplies, and needs stats 
dem <- proj_dem %>%
  select(wug_county, entity_name, wug_type, d2020:d2070) %>%
  pivot_longer(d2020:d2070) %>%
  mutate(name = gsub("d", "", name)) %>%
  rename(decade = name, 
         demands = value)

sup <- proj_supp %>%
  select(wug_county, entity_name, wug_type, ws2020:ws2070) %>%
  pivot_longer(ws2020:ws2070) %>%
  mutate(name = gsub("ws", "", name)) %>%
  rename(decade = name, 
         supplies = value)

needs <- proj_needs %>%
  select(wug_county, entity_name, wug_type, n2020:n2070) %>%
  pivot_longer(n2020:n2070) %>%
  mutate(name = gsub("n", "", name)) %>%
  rename(decade = name, 
         needs = value)

# grabbing summaries to understand demand, supplies, and projected needs: 
dem_summ <- dem %>%
  group_by(wug_type, decade) %>%
  summarize(sum_dem = sum(demands))
sup_summ <- sup %>%
  group_by(wug_type, decade) %>%
  summarize(sum_sup = sum(supplies))
needs_summ <- needs %>%
  group_by(wug_type, decade) %>%
  summarize(sum_need = sum(needs))

swp <- merge(dem_summ, sup_summ)
swp <- merge(swp, needs_summ)
swp_long <- swp %>%
  pivot_longer(sum_dem:sum_need)

ggplot(swp_long, aes(x = decade, y = value, group = name, 
                       color = name)) + 
  geom_point() + 
  geom_smooth(se = FALSE) +
  facet_wrap(~wug_type) + 
  theme_minimal() 

# Overall findings: 
#   - Out of all six possible water use types, municipal has the largest
#     increase in projected needs from 2020 - 2070
#   - Largest projected population increases are mostly near cities 
#     which is expected. Harris, Collin, and Montgomery have the largest 
#     projected population increases. 
#   - for both irrigation and municipal water categories, the demand 
#     largely outweighs water supplies (aside from 2020 in the municipal 
#     category)

# notes from sync with Charles: 
# not as important for E TX deliverable but state-wide analysis 
# mine to 2050 
# demand calculations are not always easy due to tech advancements - take 
# it with a grain of salt 
# we can't assume boundaries with aquifers - but we don't really have a choice 
# https://iopscience.iop.org/article/10.1088/1748-9326/ab0db7/meta
```

## Has the system reported any water quality violations?
```{r}
# grabbing sdiws violations by pwsid: 
viols <- keys$analysis_keys

# creating a quick graph to investigate results: 
viol_long <- viols %>%
  as.data.frame() %>%
  select(-geometry) %>%
  pivot_longer(paperwork_violations_10yr:total_violations_5yr)

# violation stats: 
viol_long %>%
  group_by(name) %>%
  filter(value > 0) %>% 
  summarize(total = length(unique(pwsid)))

# checking out trends with healthbased viols over the past 5 years and 
# relationship to % POC and MHI
ggplot(viols, aes(x = estimate_poc_alone_per, y = estimate_mhi)) + 
  geom_point() +
  geom_point(data = viols %>% filter(healthbased_violations_5yr > 0), 
             color = "red")

# grabbing stats for 5yr health-based violation bins: 
viols %>%
    as.data.frame() %>%
    select(-geometry) %>%
    mutate(viol_bin = cut(healthbased_violations_5yr, 7)) %>%
    group_by(viol_bin) %>%
    summarize(total_utilities = n(),
              mean_mhi = mean(estimate_mhi, na.rm = T), 
              mean_poc = mean(estimate_poc_alone_per, na.rm = T))

# what about all of the other categories of violations? 
ggplot(viol_long, aes(x = estimate_poc_alone_per, y = value, color = name)) +
  geom_point() +
  facet_wrap(~name, scales = "free") +
  theme_bw() +
  theme(legend.position = "none") +
  labs(x = "% POC", y = "Number of Violations") +
  ggtitle("SDWIS Violations By % POC")

ggplot(viol_long, aes(x = estimate_mhi, y = value, color = name)) +
  geom_point() +
  facet_wrap(~name, scales = "free") +
  theme_bw() +
  labs(x = "MHI", y = "Number of Violations") +
  ggtitle("SDWIS Violations By Median Household Income")


# Quick Summary: 
# 4542 pwsids in total 
#   - 703 have had health-based violations over past 5 years (15.48%)
#   - 1201 have had health-based violations over past 10 years (26.44%)
#   
#   - 2673 have had paperwork violations over the past 5 years (58.85)
#   - 3637 have had paperwork violations over the past 10 years (80.07%)
# 
#   Qualitatively, if your mHI is < 100,000 and % POC is < 30%, you're more
#   likely to have more health-based violations over the past 5 years. 
# 
#   When looking at bins of health-based violations over the past 5 years, 
#   most utilities (4484) have less than 25 violations. As the number of 
#   violation increase the MHI decreases. % POC is highest at low levels of 
#   health-based violations, but that might be because the total number of
#   utilities that fall within this category is higher. Utilities with lots of 
#   violations generally have a %POC population greater than 20%. 
```

## How many boil water notices have there been over the past five years?
```{r}
# don't have this data yet - need to be FOIA'd
```

## Have there been any enforcement actions?
```{r}
sdwis <- wds$sdwis

sdwis_filt_enf %>%
  group_by(pwsid, enf_year) %>%
  distinct(enforcement_action_type_code) 
```

## What is the Climate Vulerability Index?
```{r}
# loading cvi: 
cvi <- demo$cvi 
cvi_tidy <- cvi %>%
  select(-X) %>%
  rename(census_tract = fips_code)

# loading tract crosswalk: 
pwsids <- unique(demo$census$pwsid)
tx_crosswalk <- crosswalk %>%
  select(-X) %>%
  filter(pwsid %in% pwsids)

# merging cvi with the crosswalk: 
cvi_cross <- merge(cvi_tidy, tx_crosswalk, by.x = "census_tract", 
      by.y = "tract_geoid") %>%
  mutate(cvi_weighted  = overall_cvi_score*tract_parcel_weight) 

cvi_weighted <- cvi_cross %>%
  group_by(pwsid) %>%
  summarize(cvi_weighted_score = weighted.mean(cvi_weighted))

# adding geographies for mapping: 
pwsid_geometries <- demo$census %>%
  select(pwsid, geometry)

cvi_weighted_geoms <- merge(cvi_weighted, pwsid_geometries) %>%
  st_as_sf(.) %>%
  st_transform(., crs = st_crs(enviro_sw$major_river_basins))

cvi_pal <- colorNumeric(
  palette = viridis::mako(9),
  domain = cvi_weighted_geoms$cvi_weighted_score)

leaflet() %>%
  addProviderTiles(providers$CartoDB.VoyagerNoLabels, group = "Toner Lite") %>%
  addPolygons(data = cvi_weighted_geoms,
              opacity = 9,
              color = ~cvi_pal(cvi_weighted_score),
              weight = 1,
              label = paste0("pwsid: ", cvi_weighted_geoms$pwsid,
                "; cvi: ", round(cvi_weighted_geoms$cvi_weighted_score, 2))) %>%
  addLegend("bottomright",
            pal = cvi_pal,
            values = cvi_weighted_geoms$cvi_weighted_score,
            title = "cvi",
            opacity = 1)

# Overall findings: 
# cities had a higher CVI, if you're in the suburbs of a city you have a lower 
# CVI, rural areas have higher CVI 
```

## Does the system use green energy/focus on green energy?
```{r}
# Flagging this one for a deeper dive in the future 
```

## How much water loss does the system have?
```{r}
pwsid_keys <- demo$census %>%
  select(pwsid, pws_name) %>%
  mutate(pws_name_tidy = str_to_upper(pws_name), 
         pws_name_tidy = str_squish(pws_name_tidy), 
          # remove hyphens 
         pws_name_tidy = gsub("-", "", pws_name_tidy), 
         # remove all spaces 
         pws_name_tidy = gsub(" ", "", pws_name_tidy), 
         # remove #
         pws_name_tidy = gsub("#", "", pws_name_tidy)) %>%
  unique() %>%
  select(-pws_name)
  # st_as_sf(.) %>%
  # st_transform(., crs = st_crs(enviro_sw$major_river_basins))

wl <- wds$water_loss %>%
  janitor::clean_names() %>%
  select(year, name_of_utility, real_loss_cost_in_dollars, 
         apparent_loss_cost_in_dollars) %>%
  mutate(name_of_utility_tidy = str_to_upper(name_of_utility), 
         name_of_utility_tidy = str_squish(name_of_utility_tidy), 
           # remove hyphens 
         name_of_utility_tidy = gsub("-", "", name_of_utility_tidy), 
         # remove all spaces 
         name_of_utility_tidy = gsub(" ", "", name_of_utility_tidy), 
         # remove #
         name_of_utility_tidy = gsub("#", "", name_of_utility_tidy))

# matching names and pwsids: 
wl_pwsids <- merge(wl, pwsid_keys,
                   by.x = "name_of_utility_tidy", by.y = "pws_name_tidy", 
                   all.x = TRUE)
# 2,773 water systems 
  
# how many don't have a name match? 200
missing_pwsid <- wl_pwsids %>%
  filter(is.na(pwsid)) %>%
  select(name_of_utility_tidy) %>%
  unique()

# pulling name matches from a webscrape: https://dww2.tceq.texas.gov/DWW/JSP/SearchDispatch?number=&name=&ActivityStatusCD=All&county=All&WaterSystemType=All&SourceWaterType=All&SampleType=null&begin_date=2%2F8%2F2022&end_date=2%2F8%2F2024&action=Search+For+Water+Systems
URL <- "https://docs.google.com/spreadsheets/d/1Ds2GeJ8AMWqhBKRcIAXgO0Po8c9l_syVz4JFkeCNyXk/edit#gid=0"
dl <- googledrive::drive_download(
  as_id(URL),
  path = 'temp1.xlsx', 
  type = "xlsx")
pwsid_names <- readxl::read_excel('temp1.xlsx', col_names = TRUE) %>%
  janitor::clean_names() %>%
  mutate(water_system_name = str_to_upper(water_system_name), 
         water_system_name = str_squish(water_system_name)) 
file.remove("temp1.xlsx")

# removing fact sheet summary string and tidying before name merge: 
pwsid_names$utility_name_tidy <- unlist(strsplit(pwsid_names$water_system_name, " FACT SHEET SUMMARY SHEET"))
pwsid_names_tidy <- pwsid_names %>% select(water_system_no, utility_name_tidy) %>%
  rename(pwsid = water_system_no) %>%
  # remove hyphens 
  mutate(utility_name_tidy = gsub("-", "", utility_name_tidy), 
         # remove all spaces 
         utility_name_tidy = gsub(" ", "", utility_name_tidy), 
         # remove #
         utility_name_tidy = gsub("#", "", utility_name_tidy))
       
match_names <- merge(missing_pwsid, pwsid_names_tidy, by.x = "name_of_utility_tidy", 
      by.y = "utility_name_tidy", all.x = TRUE) %>%
  filter(!is.na(pwsid))

# recombining: 
all_pwsid_names <- merge(wl_pwsids, match_names, by = "name_of_utility_tidy", all.x = TRUE) %>%
  mutate(pwsid = case_when(
    is.na(pwsid.x) ~ pwsid.y, 
    is.na(pwsid.y) ~ pwsid.x, 
    TRUE ~ "no match"
  )) %>%
  select(-c(pwsid.x, pwsid.y))

# there are still 115 pwsids missing: 
all_pwsid_names %>%
  filter(is.na(pwsid)) %>%
  distinct(name_of_utility_tidy)

# TODO: check out the fuzzy merging code that Phil sent for matching the 
# remainder of these pwsids - most of them can reasonably match to the 
# excel spreadsheet that I read in but would rather not do that by hand 

# pwsids that require fuzzy matching: 
fuzzy <- all_pwsid_names %>%
  filter(is.na(pwsid)) %>%
  distinct(name_of_utility_tidy) 

library(fuzzyjoin)
name_dist <- stringdist_left_join(fuzzy, pwsid_names_tidy, 
                                  by = c("name_of_utility_tidy" = "utility_name_tidy"),
                                  method = "cosine",
                                  max_dist = 2, distance_col = "dist") 

```

## What type of treatment is in use? 
```{r}
# statewide basis - people pulling from sw are using different treatment 
# techniques than groundwater 
# however - there is likely going to be minimal difference 
#         - chlorine and fluoride and mechanical (membrane)
# double-check 
# filter by counties - the categories would get dropped down to 5 and then 
# determine if it's realisitc or not 

treatments_dec17 <- aws.s3::s3read_using(read.csv, 
                           object = "state-drinking-water/TX/raw/TX_TCEQ_DWW/TX_20231217/WaterSystemFacilities/TX_Treatment Plant Unit Process Flows_20231217_20231229203427.csv",
                           bucket = "tech-team-data") %>%
  clean_names()

treatments_dec03 <- aws.s3::s3read_using(read.csv, 
                           object = "state-drinking-water/TX/raw/TX_TCEQ_DWW/TX_20231203/WaterSystemFacilities/TX_Treatment Plant Unit Process Flows_20231203.csv",
                           bucket = "tech-team-data") %>%
  clean_names()

# this is probably not a comprehensive list 
all_treat <- rbind(treatments_dec03, treatments_dec17)
treat_simple <- all_treatments %>%
  group_by(water_system_no) %>%
  distinct(supply)

# desal tech --> brine is injected into wells?!?! like fracking?!
```

## What is the source of water? 
```{r}
# we can match sw intakes with pwsids: 
sw_intake_keys <- keys$pwsid_sw

census <- demo$census
pwsid_info <- census %>%
  select(pwsid:pop_density) 

pwsid_sw <- merge(pwsid_info, sw_intake_keys, by = c("pwsid"), all.y = TRUE)
# matched with 449 entries - some pwsids seem to take water from multiple 
# surface water reservoirs? Also there's a handful of pwsids that are "GW" but 
# matched with a surface water intake 

# looking at county and groundwater keys: 
gw_keys <- keys$well_gw_county
county_gw <- gw_keys %>%
  group_by(county_name) %>%
  distinct(aquifer_one, aquifer_two, aquifer_three, 
           aquifer_four, aquifer_five)
```

Potential rabbit holes of interest: 
- SDIWS violations - investigating the type of violation (i.e. MON, RPT (?), MR (monitoring and reporting), TT (treatment technique), MCL (max contaminant level), and MRDL (max residual disinfectant))
- Deeper dive into SRF green project reserve
