---
title: "TX Water Systems - Bridge Grant Analysis"
author: "EmmaLi Tsai"
date: "2024-09-23"
output: 
  html_document:
    toc: true
    theme: united
---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
# notes from sync w/ charles: 
# - zoom in on utilities in East TX for final sections of analysis 
# - investigate methods for adjusting for inflation 
# - more text on water rates and how that's a limiting factor 
# - red/green color blind issues on % change graphs 
# - re-adjust for inflation when looking at mhi and water rates?
```

# 0.0 Set-up
```{r message = F, results='hide'}
# No scientific notation
options(scipen = 99999999)
options(tigris_use_cache = TRUE)

# Libraries: 
library(aws.s3)
library(sf)
library(tidyverse)
library(tidycensus)
library(scales)
library(leaflet)
library(plotly)
library(tigris)

# grabbing national xwalk: 
# NOTE: the xwalk has been updated with the fixed epa block parcel crosswalk 
# from blue conduit, and therefore likely has more "accurate" data for tier 
# 2 crosswalked systems. Going to resort to primarily using the xwalk 
xwalk <- aws.s3::s3read_using(st_read, 
                              object = "/service_area_boundaries/epa-sabs/epa-sabs-crosswalk.geojson",
                              bucket = "tech-team-data",
                              quiet = TRUE)

# grabbing data from TX application: 
app_data <-  aws.s3::s3read_using(st_read, 
                                  object = "/state-drinking-water/TX/clean/app/app-data-prod.geojson",
                                  bucket = "tech-team-data",
                                  quiet = TRUE)

# filtering xwalk for pwsids in app_data, and adding water system information 
# that's missing from the EPA crosswalk
app_data_ws_info <- app_data %>%
  as.data.frame() %>%
  select(pwsid:primary_source_code) 

tx_xwalk <- xwalk %>%
  filter(pwsid %in% app_data$pwsid) %>%
  left_join(app_data_ws_info, by = "pwsid")

# grabbing baseline 2020 census stats for TX for percentages: 
tx_county_pop <- tidycensus::get_acs(
  geography = "county",
  variables = c(total_pop = "B01003_001"),
  state = "TX",
  year = 2020,
  geometry = TRUE
)

# financial data from the tx project: 
fin <- aws.s3::s3read_using(readRDS, 
                            object = "s3://tech-team-data/state-drinking-water/TX/clean/TX_financial_list.RData")


# themes: 
sysfonts::font_add_google("Lato")
showtext::showtext_auto()

cat_palette <- colorRampPalette(c("#172f60","#1054a8",
                                  "#791a7b","#de9d29", 
                                  "#b15712","#4ea324"))
cont_palette <- colorRampPalette(c("#172f60","#4ea324"))

epic_chart_theme <- theme_minimal() + 
  theme(legend.position = "right", 
        text = element_text(size = 13, family = "Lato"), 
        legend.text = element_text(size = 10, family = "Lato"), 
        legend.title = element_text(size = 11, family = "Lato"), 
        axis.text.x = element_text(margin = margin(t = 10, r = 0, 
                                                   b = 0, l = 0)), 
        axis.title.x = element_text(margin = margin(t = 10, r = 0, 
                                                    b = 0, l = 0)), 
        axis.text.y = element_text(margin = margin(t = 0, r = 10, 
                                                   b = 0, l = 0)), 
        axis.title.y = element_text(margin = margin(t = 0, r = 10, 
                                                    b = 0, l = 0))) 

# function for getting sab xwalked stats over time: 
source("/Users/emmalitsai/TX-drinking-water/code/functions/state_sab_crosswalk_year.R")
```

# 1.0: What is the size of the opportunity?
## 1.1: What is the number of Texans served by CWS?
```{r}
# what source codes are we working with?
# unique(tx_xwalk$primary_source_code)

# summing population where source code is surface or surface purchased: 
tx_pop_sw <- tx_xwalk %>%
  as.data.frame() %>%
  # filter(primary_source_code %in% c("Surface", "Surface-Purchased")) %>%
  summarize(total_population_surface = sum(estimate_total_pop, na.rm = T))

# how does this compare to the total population in 2020?
tx_pop_total <- sum(tx_county_pop$estimate)
```
The number of Texans served by CWS is `r formatC(round(tx_pop_sw$total_population_surface, 0), format = "d", big.mark = ",")`. With `r formatC(tx_pop_total, format = "d", big.mark = ",")` people living in Texas, this reflects approximately `r round(100*(tx_pop_sw/tx_pop_total), 2)`% of the population. 

## 1.2: What is the amount of investment that goes into CWS across the state? 
```{r}
# From my notes, this is the TX DWSRF analysis from Phil & Danielle: 
# this is from https://github.com/Environmental-Policy-Innovation-Center/water-data/tree/main/clean-data/dwsrf-awards
# analysis doc: https://docs.google.com/document/d/1zzS5IASdb7CtN8WcIs8u17CaFpKJYjNrmriGB6_WVZQ/edit
# contains FOIA'd DWSRF awards from October 2009 - 2020
srf <- fin$TX_DWSRF_Analysis 
tx_srf <- srf %>% 
  filter(pwsid %in% tx_xwalk$pwsid)

# TODO OR - use the srf dashboard to pull 2020-2023?

# which ones have received funding?
srf_funded <- tx_xwalk %>%
  filter(pwsid %in% tx_srf$pwsid)

# what % of the population does this reflect?
# TODO-  is this the project_population_served?
srf_funded_pop <-  tx_xwalk %>%
  as.data.frame() %>%
  # select(pwsid, project_population_served) %>%
  # unique() %>%
  filter(pwsid %in% tx_srf$pwsid) %>%
  summarize(funded_pop = sum(estimate_total_pop, na.rm = T))

# what % of the population served by CWS does this reflect?
total_cws_pop <- sum(tx_xwalk$estimate_total_pop, na.rm = T)


# how much total assistance has been provided?
assistance_stats <- tx_srf %>%
  summarize(total_assistance = dollar(sum(assistance_amt)), 
            total_pf = dollar(sum(prin_forgive_amt)), 
            mean_assistance = dollar(mean(assistance_amt)), 
            mean_pf = dollar(mean(prin_forgive_amt)))

```
Out of `r formatC(length(unique(tx_xwalk$pwsid)), format = "d", big.mark = ",")` CWS in Texas, `r length(unique(srf_funded$pwsid))` have received Drinking Water State Revolving Funds over 2009-2020. This reflects `r round((length(unique(srf_funded$pwsid))/length(unique(tx_xwalk$pwsid))*100), 2)`% of CWS in Texas. 

Based on water system populations, systems that have been funded by DWSRF serve `r formatC(round(srf_funded_pop$funded_pop, 0), format = "d", big.mark = ",")` people, which reflects `r round((srf_funded_pop$funded_pop / total_cws_pop)*100, 0)`% of the population served by CWS in Texas, and `r round((srf_funded_pop$funded_pop / tx_pop_total)*100, 0)`% of the total population in Texas. 

```{r}
# how has the amount of total assistance/year changed through time?
srf_year_summary <- tx_srf %>% 
  group_by(year) %>%
  summarize(total_amt = sum(assistance_amt))

# plot to show results through time: 
ggplot(srf_year_summary, aes(x = year, y = total_amt)) + 
  ggchicklet::geom_chicklet(stat = "identity", 
                            color = "white", 
                            fill = "#1054a8", 
                            alpha = 0.7) + 
  scale_y_continuous(labels = scales::dollar_format(prefix="$")) + 
  epic_chart_theme + 
  labs(x = "Year", 
       y = "Total Assistance ($)") + 
  ggtitle("Amount of DWSRF Assistance has Increased Since 2009")

```

From 2009-2020, the amount of total assistance received by CWS in Texas is `r assistance_stats$total_assistance`, with an average of `r assistance_stats$mean_assistance` across our study period. The total amount of principal forgiveness is `r assistance_stats$total_pf`, with an average of `r assistance_stats$mean_pf`.

```{r}
# what counties have received the most funding? 
tx_funded_counties <- tx_xwalk %>%
  as.data.frame() %>%
  filter(pwsid %in% tx_srf$pwsid) %>%
  select(pwsid, county_served)

tx_srf_counties <- merge(tx_srf, tx_funded_counties, all = T)

tx_srf_county_summary <- tx_srf_counties %>%
  group_by(county_served) %>%
  summarize(total_assistance = sum(assistance_amt), 
            total_yrs_funded = length(unique(year)), 
            assistance_per_yr = total_assistance/total_yrs_funded) %>%
  mutate(county_clean = paste0(county_served, " County, Texas"))

# merging with TX county geometries for mapping: 
tx_county_simple <- tx_county_pop %>%
  select(NAME, geometry) %>%
  unique()
tx_srf_county_geom <- merge(tx_srf_county_summary, tx_county_pop, 
                            by.x = "county_clean", by.y = "NAME", 
                            all.x = T) %>%
  st_as_sf() 


# plotting: 
assistance_pal <- colorNumeric(
  palette = cont_palette(5),
  domain = tx_srf_county_geom$total_assistance)

leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron, 
                   group = "Toner Lite") %>%
    addPolygons(data = tx_county_pop,
              # slight opacity
              fillOpacity = 0.2,
              color = "gray",
              weight = 1, 
              smoothFactor = 0) %>%
  addPolygons(data = tx_srf_county_geom,
              fillColor = ~assistance_pal(total_assistance),
              # slight opacity
              fillOpacity = 0.8,
              color = "white",
              weight = 1, 
              smoothFactor = 0, 
              label = paste0("County: ", tx_srf_county_geom$county_served, 
                             "; Total Assistance : ",
                             dollar(tx_srf_county_geom$total_assistance), 
                             "; Total Years Funded: ",
                             (tx_srf_county_geom$total_yrs_funded), 
                             "; Assistance/Year: ",
                             dollar(tx_srf_county_geom$assistance_per_yr))) %>%
  addLegend("bottomright",
            pal = assistance_pal,
            values = tx_srf_county_geom$total_assistance,
            # if working with money or percentages, use the 
            # follow label formats: 
            labFormat = labelFormat(prefix = "$"),
            # labFormat = labelFormat(prefix = "%"),
            title = "Total Assistance ($)\n - 2009 to 2020",
            opacity = 1)

```
Which counties have received the most funding?

The counties that have received the largest total assistance over our study period are located around large urban centers such as Dallas (both Dallas & Tarrant county), San Antonio (Bexar county), and Houston (Brazoria county). Bexar county around San Antonio has received the largest amount of funding over our study period, with ~$200 million received over the course of 7 years. 


```{r}
# what are the characteristics of utilities that have received funding?
tx_funding_flag <- tx_xwalk %>%
  as.data.frame() %>%
  mutate(funded = case_when(
    pwsid %in% tx_srf$pwsid ~ "Funded", 
    TRUE ~ "Not Funded"
  )) %>%
  select(pwsid, funded, estimate_poc_alone_per, estimate_mhi, 
         estimate_laborforce_unemployed_per, population_served_count,
         dwater, 
         # primary_source_code, 
         estimate_hh_below_pov_per) %>%
  rename("%POC" = estimate_poc_alone_per, 
          "MHI" = estimate_mhi, 
         `Population Served` = population_served_count,
         `% Unemployment` = estimate_laborforce_unemployed_per, 
          `Drinking Water Metric` = dwater,
          # `Water Source` = primary_source_code, 
          `% Households in Poverty` = estimate_hh_below_pov_per)


# summary stats for text: 
summary_stats <- tx_funding_flag %>%
  group_by(funded) %>%
  summarize(across(2:7, ~mean(.x, na.rm = T))) %>%
  rename("poc" = `%POC`, 
         "unemp" =  `% Unemployment`, 
         "pop" =`Population Served`, 
         "dw" = `Drinking Water Metric`,
         "poverty" = `% Households in Poverty`)

funded <- summary_stats %>%
  filter(funded == "Funded")
not_funded <- summary_stats %>%
  filter(funded != "Funded")

# is there any difference by source water? 
by_source <- tx_xwalk %>%
  as.data.frame() %>%
  mutate(funded = case_when(
    pwsid %in% tx_srf$pwsid ~ "Funded", 
    TRUE ~ "Not Funded"
  )) %>%
  select(pwsid, funded, primary_source_code) %>%
  group_by(funded, primary_source_code) %>%
  tally()

# long for plotting: 
tx_funding_long <- tx_funding_flag %>%
  pivot_longer(., cols = 3:8)

ggplot(tx_funding_long, aes(x = funded, y = value, 
                            color = funded)) + 
  geom_jitter(width = 0.05, alpha = 0.2) + 
  geom_boxplot(alpha = 0.6, 
                 outliers = FALSE) + 
  facet_wrap(~name, scales = "free") + 
  scale_color_manual(values = cat_palette(2), 
                     name = "") + 
  epic_chart_theme +
  labs(x = "", y = "Value") + 
  theme(legend.position = "bottom")
```

What are the characteristics of utilities that have received funding?

Based on the metrics plotted above, utilities that have received funding have a larger mean percentage of households in poverty by `r round(funded$poverty - not_funded$poverty, 2)`%. These utilities also have a larger mean population they serve by `r round(funded$pop - not_funded$pop, 0)` people, likely because many of the utilities that have received funding provide services to large population centers, as shown in the question above. Funded systems also have a lower median household income than unfunded systems by `r dollar(abs(funded$MHI - not_funded$MHI))`, and serve less diverse communities. Both funded and unfunded systems were found to be primarily pulling groundwater. 

One of the most striking difference across all of our indicators was EPA's new drinking water metric, where systems that received funding had a mean value that was `r round(funded$dw - not_funded$dw, 2)` larger than systems that have not received funding. 


# 2.0: What are the characteristics of a fragile system?
```{r}
# TX SRF DAC definition: https://www.epa.gov/system/files/documents/2022-10/DWSRF%20DAC%20Definitions%20Report_October%202022%20Updates_FINAL_508.pdf
# affordability criteria: 
# (a) Has an Annual Median Household Income (AMHI) that is no more than 75 percent of the state median household income using an acceptable source of socioeconomic data, and
# (b) the Household Cost Factor (HCF) that considers income, unemployment rates, and population trends must be greater than or equal to 1 percent if only water or sewer service is provided or greater than or equal to 2 percent if both water and sewer service are provided.

# The Household Cost Factor in the IUP is the cost of water and wastewater utilities per household (after completion of the proposed project) in the project service area compared to the project area’s current median household income level, with two adjustments. Specifically, the formula is:

# A=Average annual water bill per household
# B= Average annual wastewater bill per household
# C= Annual loan cost per household of the new financing
# D= AMHI for the project service area
# HCF = (A + B + C) / D

# In addition, the state includes two adjustments that can only increase the HCF (i.e., benefit the applicant) - this includes unemployment rates & population adjustments 


# part a: what is the 75% of state MHI?
tx_mhi <- tidycensus::get_acs(
  geography = "state",
  variables = c(mhi = "B19013_001"),
  state = "TX",
  year = 2020,
  geometry = TRUE
)
pct_75_mhi <- tx_mhi$estimate*0.75


# grabbing d: from acs 2020 estimate
d <- tx_xwalk %>% 
  as.data.frame() %>%
  select(pwsid, estimate_mhi) %>%
  mutate(meets_criteria_a = 
           case_when(estimate_mhi <= pct_75_mhi ~ "Yes", 
                     TRUE ~ "No"))

# grabbing ab: duke's affordability dashboard for water and sewer rates 
# NOTE - these only have data for 706 pwsids
water_rates <- fin$TX_affordability_DAC
# these all have sewer and water 
ab <- water_rates %>%
  as.data.frame() %>%
  mutate(water_sewer_year = total_water_year + total_sewer_year) %>%
  select(pwsid, water_sewer_year)

# grabbing c: annual loan cost per household of new financing
# talked w/ Phil, and this is difficult for a number of reasons. 



# TX definition of "safe, adequate, and reliable"
# Texas health definition code - approved or superior 
# test <- aws.s3::s3read_using(read.csv, 
#                               object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_TCEQ_DWW/TX_20231217/WaterSystemDetail/TX_Water System Details_20231217_20231229203427.csv")
# test %>%
#   filter(Type == "System  Recognition") %>%
#   group_by(Value) %>%
#   tally()
# lol nevermind: 
#   Value           n
#   <chr>       <int>
# 1 APPROVED       19
# 2 NO DATA     13625
# 3 PROVISIONAL     8
# 4 SUPERIOR      738

# TX definition of a failing system 
```

# 3.0: How do water systems change over time in terms of fragility and durability? 
```{r eval = FALSE}
# prep for this section: 

# which system have been around for that long?
# NOTE - absolute file paths needed to knitting rmd 
sdwa_info <- read.csv("/Users/emmalitsai/TX-drinking-water/data/raw/sdwa_download/SDWA_FACILITIES.csv") %>%
  mutate(start_date = as.Date(FIRST_REPORTED_DATE, tryFormats = c("%m/%d/%Y")),
         year_date = year(start_date))

# # NOTE - this is pwsids with FACILITIES where the first report is at least 2010
pwsids_since_2010 <- sdwa_info %>%
  filter(year_date <= 2010) %>%
  filter(PWSID %in% tx_xwalk$pwsid) %>%
  distinct(PWSID)

# # filtering SABs for just these 
tx_xwalk_2010_pwsids <- tx_xwalk %>%
  filter(pwsid %in% pwsids_since_2010$PWSID) %>%
  select(pwsid:shape_length)

# NOTE - only need to run this if running the xwalk again - shouldn't need
# to run while knitting 
# need to reconstruct block housing and population weights - the tigris block()
# function does not return this information from 2010 - 2019. 
# grabbing 2020: 
state_blocks_2020 <- tigris::blocks(
  state = "TX",
  year = 2020)

# grabbing 2010:
# NOTE - with going back to 2010, we've lost education categories and 
# % unemployment 
# dec_vars_all <- load_variables(2010, "sf1", cache = TRUE)
dec_vars <- c(HOUSING10 = "H001001",
              POP10 = "P001001")
state_blocks_2010 <- get_decennial(
  geography = "block",
  variables = dec_vars,
  year = 2010,
  state = "TX",
  geometry = TRUE
)

# pivoting to wide to work with xwalk: 
state_blocks_2010_wide <- state_blocks_2010 %>%
  pivot_wider(., names_from = variable, values_from = value)

# 2015 is going to be tricky - grabbing a relationship file:
file <-"https://www2.census.gov/geo/docs/maps-data/data/rel2020/t10t20/TAB2010_TAB2020_ST48.zip"
file_loc <- "/Users/emmalitsai/TX-drinking-water/data/raw/census_rel"
download.file(file,
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(fsile_loc, ".zip"), exdir = file_loc)
file.remove(paste0(file_loc, ".zip"))

# reading in block relationships from 2010 to 2020 in TX:
block_2010_2020_rel <- read.table("/Users/emmalitsai/TX-drinking-water/data/raw/census_rel/tab2010_tab2020_st48_tx.txt", 
                                  sep = "|")
# correcting weird column names: 
names(block_2010_2020_rel) <- block_2010_2020_rel[1,]
block_2010_2020_rel <- block_2010_2020_rel[2:nrow(block_2010_2020_rel),]

# building block relationships
block_rel <- block_2010_2020_rel %>%
  mutate(GEOID_10 = paste0(STATE_2010, COUNTY_2010, TRACT_2010, BLK_2010), 
         GEOID_20 = paste0(STATE_2020, COUNTY_2020, TRACT_2020, BLK_2020))

# want to keep 2010 block geometries, but grab mean housing and pop between two 
# time periods
block_2020_simple <- state_blocks_2020 %>%
  as.data.frame() %>%
  select(GEOID20, HOUSING20, POP20) 
block_2020_rel <- merge(block_2020_simple, block_rel, by.x = "GEOID20", 
                        by.y = "GEOID_20", all = T)
block_2010_rel <- merge(state_blocks_2010_wide, block_2020_rel, 
                        by.x = "GEOID", by.y = "GEOID_10", all = T)

# grabbing mean between years of housing data, AND grouping by 2010 
# geoid to calculate mean values for each original 2010 geoid (some blocks 
# were either merged or split over the decade)
state_blocks_2015 <- block_2010_rel %>%
  as.data.frame() %>%
  filter(!(is.na(HOUSING10))) %>%
  select(GEOID, GEOID20, HOUSING10, HOUSING20, POP10, POP20) %>%
  rowwise() %>%
  mutate(HOUSING15 = mean(c(HOUSING10, HOUSING20)), 
         POP15 = mean(c(POP10, POP20))) %>%
  group_by(GEOID) %>%
  summarize(HOUSING15 = mean(HOUSING15), 
            POP15 = mean(POP15))

# adding original 2010 geometries back: 
geom_2010 <- state_blocks_2010_wide %>%
  select("GEOID")
state_blocks_2015 <- merge(state_blocks_2015, geom_2010, by = "GEOID") %>%
  st_as_sf()

# crosswalkin' 2015 and 2010: 
tx_xwalk_2015 <- state_sab_crosswalk_year(tx_xwalk_2010_pwsids, "TX", 2015,
                                          state_blocks_2015,
                                          "WGS84")
# st_write(tx_xwalk_2015, "/Users/emmalitsai/TX-drinking-water/data/raw/tx_xwalk_2015.geojson")
# put_object(
#   file = "/Users/emmalitsai/TX-drinking-water/data/raw/tx_xwalk_2015.geojson",
#   object = "/state-drinking-water/TX/raw/TX_xwalk_2015.geojson",
#   bucket = "tech-team-data"
# )

tx_xwalk_2010 <- state_sab_crosswalk_year(tx_xwalk_2010_pwsids, "TX", 2010,
                                            state_blocks_2010_wide,
                                            "WGS84")
# st_write(tx_xwalk_2010, "/Users/emmalitsai/TX-drinking-water/data/raw/tx_xwalk_2010.geojson")
# put_object(
#   file = "/Users/emmalitsai/TX-drinking-water/data/raw/tx_xwalk_2010.geojson",
#   object = "/state-drinking-water/TX/raw/TX_xwalk_2010.geojson",
#   bucket = "tech-team-data"
# )

# download SDWA database and look over entire timescale 
file_loc <- "/Users/emmalitsai/TX-drinking-water/data/raw/sdwa_download"
# download.file("https://echo.epa.gov/files/echodownloads/SDWA_latest_downloads.zip", 
#               destfile = paste0(file_loc, ".zip"))
# unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
# file.remove(paste0(file_loc, ".zip"))

```

## 3.1: What changes do we see to population growth in fragile systems? 
```{r fig.height = 9, fig.width = 10}
# population growth through time, and ideally this would also include 
# MHI 
tx_xwalk_2010 <- aws.s3::s3read_using(st_read, 
                                      object = "/state-drinking-water/TX/raw/TX_xwalk_2010.geojson",
                                      bucket = "tech-team-data",
                                      quiet = TRUE)

tx_xwalk_2015 <- aws.s3::s3read_using(st_read, 
                                      object = "/state-drinking-water/TX/raw/TX_xwalk_2015.geojson",
                                      bucket = "tech-team-data",
                                      quiet = TRUE)

# tidying and mutating: 
tx_xwalk_2015 <- tx_xwalk_2015 %>% 
  mutate(year = "2015")
tx_xwalk_2010 <- tx_xwalk_2010 %>% 
  mutate(year = "2010")
tx_xwalk_2020 <- tx_xwalk %>%
  filter(pwsid %in% tx_xwalk_2010$pwsid) %>%
  mutate(year = "2020")

# selecting only columns of interest for now: 
tx_2015_simple <- tx_xwalk_2015 %>%
  as.data.frame() %>%
  select(pwsid, tier_crosswalk, estimate_total_pop, estimate_mhi)
names(tx_2015_simple) <- paste0(names(tx_2015_simple), "_2015")

tx_2010_simple <- tx_xwalk_2010 %>%
  as.data.frame() %>%
  select(pwsid, tier_crosswalk, estimate_total_pop, estimate_mhi)
names(tx_2010_simple) <- paste0(names(tx_2010_simple), "_2010")

tx_2020_simple <- tx_xwalk_2020 %>%
  as.data.frame() %>%
  select(pwsid, tier_crosswalk, estimate_total_pop, estimate_mhi)
names(tx_2020_simple) <- paste0(names(tx_2020_simple), "_2020")

# comparisons over a decade: 
tx_decade <- merge(tx_2010_simple, tx_2015_simple, 
                   by.x = "pwsid_2010", 
                   by.y = "pwsid_2015", all = T) 

tx_decade <- merge(tx_decade, tx_2020_simple, by.x = "pwsid_2010", by.y = "pwsid_2020", all = T) %>%
# reducing analysis to ONLY tier 1 crosswalk for ALL time periods 
  filter(tier_crosswalk_2010 == "tier_1" & tier_crosswalk_2020 == "tier_1" & tier_crosswalk_2015 == "tier_1") %>%
  # adjusting for inflation - using: https://data.bls.gov/cgi-bin/cpicalc.pl?cost1=2000&year1=201001&year2=202001
  mutate(estimate_mhi_2010_inf_adj = estimate_mhi_2010*1.190525, 
         estimate_mhi_2015_inf_adj = estimate_mhi_2015*1.10283, 
         pct_change_pop_2010_2015 = 100*((estimate_total_pop_2015 - estimate_total_pop_2010)/estimate_total_pop_2010),
         pct_change_mhi_2010_2015 = 100*((estimate_mhi_2015_inf_adj - estimate_mhi_2010_inf_adj)/estimate_mhi_2010_inf_adj),
         pct_change_pop_2015_2020 = 100*((estimate_total_pop_2020 - estimate_total_pop_2015)/estimate_total_pop_2015),
         pct_change_mhi_2015_2020 = 100*((estimate_mhi_2020 - estimate_mhi_2015_inf_adj)/estimate_mhi_2015_inf_adj),
         pct_change_pop_2010_2020 = 100*((estimate_total_pop_2020 - estimate_total_pop_2010)/estimate_total_pop_2010), 
         pct_change_mhi_2010_2020 = 100*((estimate_mhi_2020 - estimate_mhi_2010_inf_adj)/estimate_mhi_2010_inf_adj)) %>%
  rename(pwsid = pwsid_2010)

# adding geoms back: 
tx_geoms <- tx_xwalk %>% 
  select(pwsid, county_served)

tx_decade_geoms <- tx_decade %>%
  left_join(tx_geoms, by = "pwsid") %>%
  st_as_sf()

# plotting: 
pop_pal <- colorBin(
  bins = c(-90, -20, 0, 20, 100, 20000),
    (c("#eab28e", "#f4cbb3", "white", "#b1dac9", "#8cc7b0")),

  # palette = c("#d73027", "#fc8d59", "white", "#d9ef8b", "#91cf60", "#1a9850", "darkgreen"),
  domain = tx_decade_geoms$pct_change_pop_2010_2020)

leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron, 
                   group = "Toner Lite") %>%
  addPolygons(data = tx_decade_geoms,
              fillColor = ~pop_pal(pct_change_pop_2010_2020),
              # slight opacity
              fillOpacity = 0.8,
              stroke = T, 
              color = "black",
              opacity = 0.2, 
              weight = 1,
              # smoothFactor = 0.5, 
              label = paste0("pwsid: ", tx_decade_geoms$pwsid, 
                             "; % pop change : ",
                             round(tx_decade_geoms$pct_change_pop_2010_2020, 0))) %>%
  addLegend("bottomright",
            pal = pop_pal,
            values = tx_decade_geoms$pct_change_pop_2010_2020,
            # if working with money or percentages, use the 
            # follow label formats: 
            # labFormat = labelFormat(prefix = "$"),
            labFormat = labelFormat(prefix = "%"),
            title = "% Population Change 2010 to 2020",
            opacity = 1)
```

Broadly, systems in suburban areas around cities have grown, while populations within cities have shrunk since 2010. Water systems located in NE Texas have generally exhibited population decline since 2010, along with more rural areas of Texas. 

```{r fig.height = 9, fig.width = 10}
mhi_pal <- colorBin(
  bins = c(-100, -50, -20, 0, 20, 300),
  (c("#eab28e", "#f4cbb3", "white", "#b1dac9", "#8cc7b0")),
  # palette = c("#d73027", "#fc8d59", "white", "#d9ef8b", "#91cf60", "#1a9850", "darkgreen"),
  domain = tx_decade_geoms$pct_change_mhi_2010_2020)

leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron, 
                   group = "Toner Lite") %>%
  addPolygons(data = tx_decade_geoms,
           fillColor = ~mhi_pal(pct_change_mhi_2010_2020),
              # slight opacity
              fillOpacity = 0.8,
              stroke = T, 
              color = "black",
              opacity = 0.2, 
              weight = 1,
              # smoothFactor = 0.5, 
              label = paste0("pwsid: ", tx_decade_geoms$pwsid, 
                             "; % MHI change : ",
                             round(tx_decade_geoms$pct_change_mhi_2010_2020, 0))) %>%
  addLegend("bottomright",
            pal = mhi_pal,
            values = tx_decade_geoms$pct_change_mhi_2010_2020,
            # if working with money or percentages, use the 
            # follow label formats: 
            # labFormat = labelFormat(prefix = "$"),
            # labFormat = labelFormat(prefix = "%"),
            title = "% MHI Change 2010 to 2020",
            opacity = 1)
```

Median household income (already adjusted for inflation) within CWS in Texas have broadly increased. There are some water systems located in mostly rural areas that have exhibited slight declines in MHI over the past decade. 

```{r fig.height = 9, fig.width = 10}

# looking at % change in population and % change in median household income: 
leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron, 
                   group = "Toner Lite") %>%
  bivariatechoropleths::addBivariateChoropleth(
    map_data = tx_decade_geoms,
    var1_name = pct_change_pop_2010_2020,
    var2_name = pct_change_mhi_2010_2020,
    ntiles= 3,
    var1_label = "Population % Change",
    var2_label = "MHI % Change",
    # region_name = "CSDNAME",
    weight = 1,
    fillOpacity = 0.7,
    color = "grey",
    highlightOptions = leaflet::highlightOptions(color = "black",
                                                 weight = 2,
                                                 opacity = 1)) %>%
  # if I don't add this, we can't see pwsids :(
  addPolygons(data = tx_decade_geoms, 
              fillOpacity = 0, 
              opacity = 0, 
              label = paste0("pwsid: ", tx_decade_geoms$pwsid,
                             "; % MHI change : ",
                             round(tx_decade_geoms$pct_change_mhi_2010_2020, 0),
                             "; % Pop change : ",
                             round(tx_decade_geoms$pct_change_pop_2010_2020, 0)))
```

Above, you can see CWS boundaries mapped by % change in population and MHI from 2010-2020. The areas of particular interest are in bright teal, where the CWS exhibited increases in population, but little to no change in MHI. Areas located in white are also of interest, as this indicates a system that is losing both populations and income from 2010-2020. 

```{r, results = "asis"}
cws_of_interest <- tx_decade_geoms %>%
  mutate(Category = case_when(
    (pct_change_pop_2010_2020 > 0 & pct_change_mhi_2010_2020 < 0) ~ "Pop Growing, MHI Decreasing", 
    (pct_change_pop_2010_2020 < 0 & pct_change_mhi_2010_2020 < 0) ~ "Pop & MHI Decreasing", 
  TRUE ~ "No Concern Yet"))

table <- cws_of_interest %>% 
  as.data.frame() %>%
  group_by(Category) %>%
  summarize(CWS = n())
knitr::kable(table, caption = "Number of CWS in each Category")
```

While most CWS exhibit positive MHI and growth over the past decade, there are ~836 CWS in TX where the population has grown over time, but the MHI has decreased. There are 490 that have decreased in both population and MHI over the course of the past decade - indicating the system may be serving less people and with lower income now than in 2010. 

```{r fig.height = 9, fig.width = 10}
# plotting 
# ggplot(tx_decade, aes(x = pct_change_pop, y = pct_change_mhi)) + 
#   geom_point() + 
#   labs(x = "% Population Change", 
#        y = "% MHI Change") + 
#   epic_chart_theme

# combining all of the data: 
tx_2010_all <- tx_xwalk_2010 %>%
  as.data.frame() %>%
  select(-starts_with("geom"))

tx_2015_all <- tx_xwalk_2015 %>%
  as.data.frame() %>%
  select(-starts_with("geom"))

tx_2020_all <- tx_xwalk_2020 %>%
  as.data.frame() %>%
  select(-starts_with("geom")) %>%
  select(names(tx_2010_all)) %>%
  mutate(date_modified = as.character(date_modified), 
         verification_date = as.character(verification_date))

tx_decade_all <- bind_rows(tx_2010_all, tx_2015_all, 
                           tx_2020_all) %>%
  select(pwsid, estimate_total_pop:year) %>%
  select(-contains("age")) %>%
  select(pwsid, estimate_white_alone_per:estimate_hh_below_pov_per,
         estimate_pop_below_200_pov_per, year)


# pivoting to long for formatting:  
tx_decade_long <- tx_decade_all %>%
  pivot_longer(., estimate_white_alone_per:estimate_pop_below_200_pov_per)

ggplot(tx_decade_long, aes(x = year, 
                           y = value, color = year)) + 
  geom_jitter(width = 0.2, alpha = 0.1) +
  geom_violin(alpha = 0.8) +
  facet_wrap(~name, scales = "free") + 
  epic_chart_theme + 
  scale_color_manual(values = cat_palette(3), 
                     name = "") +
  labs(x = "Year", y = "Value")
```

Demographic data for CWS in TX are relatively similar across the study period. Broadly, the %mixed population served by CWS in TX have increased sinc 2010, and the % of households below the poverty level have decreased. 


## 3.2: What changes in violations do we see in fragile systems? 
```{r fig.height = 9, fig.width = 10}
# working directory: - using a direct file path to knit the rmd 
# grabbing latest data from sdwa
# file_loc <- "./data/raw/sdwa_download"
# download.file("https://echo.epa.gov/files/echodownloads/SDWA_latest_downloads.zip",
#               destfile = paste0(file_loc, ".zip"))
# unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc)
# file.remove(paste0(file_loc, ".zip"))

file_loc <- "/Users/emmalitsai/water-data/epa-sabs-comparison/data/sdwa_download"
# reading in violations: 
# NOTE - there are quite a few rows w/o violation_ids, but enforcement_ids
sdwa_viols <- read.csv(paste0(file_loc, "/SDWA_VIOLATIONS_ENFORCEMENT.csv")) %>%
  janitor::clean_names() 

# filtering for pwsids located in our CWS list: 
tx_viols <- sdwa_viols %>% 
  # tx_xwalk dataframe contails all CWS in TX (already filtered using SDWIS)
  filter(pwsid %in% tx_xwalk$pwsid) %>%
  mutate(rule_code = as.character(rule_code))

# merging in rule family reference codes:
sdwa_codes <- read.csv(paste0(file_loc, "/SDWA_REF_CODE_VALUES.csv")) %>%
  janitor::clean_names() 
rule_family_codes <- sdwa_codes %>% 
  filter(value_type %in% c("RULE_FAMILY_CODE")) %>%
  rename(rule = value_description)

# adding rule code: 
tx_viol_rulecode <- merge(tx_viols, rule_family_codes, 
                          by.x = "rule_family_code", 
                          by.y = "value_code", 
                          all.x = T) %>%
### START OF STEPS FROM EMAIL: ################################################
  # STEP ONE: concatenate PWSID and violation_id to create a unique identifier: 
  mutate(pwsid_viol_id = paste0(pwsid, "-", violation_id))

# STEP TWO: remove duplicates, keep distinct records using unique ID from 
# step 1: 
tx_viol_filt <- tx_viol_rulecode %>%
  distinct(pwsid_viol_id, .keep_all = T) %>%
# STEP THREE: filter for health viols- data has already been filtered for 
# CWS in TX
  filter(is_health_based_ind == "Y")

# STEP FOUR: group by compliance year, rulecode, and summarize totals: 
tx_viol_rule_summary <- tx_viol_filt %>%
  # converting to compl begin date to date format, and grabbing year for 
  # grouping: 
  mutate(compl_begin_date = as.Date(compl_per_begin_date, 
                                    tryFormats = c("%m/%d/%Y")), 
         start_year = year(compl_begin_date)) %>%
  # grouping: 
  group_by(start_year, rule) %>%
  summarize(total = length(unique(pwsid_viol_id)))

# plotting: 
health_viol_plot <- ggplot(tx_viol_rule_summary, aes(x = start_year, y = total, 
                        fill = rule)) + 
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = cat_palette(11), 
                    name = "") +
  theme_minimal() + 
  theme(legend.position = "right", 
        text = element_text(size = 13), 
        legend.text = element_text(size = 10), 
        legend.title = element_text(size = 11), 
        axis.text.x = element_text(margin = margin(t = 10, r = 0, 
                                                   b = 0, l = 0)), 
        axis.title.x = element_text(margin = margin(t = 10, r = 0, 
                                                    b = 0, l = 0)), 
        axis.text.y = element_text(margin = margin(t = 0, r = 10, 
                                                   b = 0, l = 0)), 
        axis.title.y = element_text(margin = margin(t = 0, r = 10, 
                                                    b = 0, l = 0))) +
  labs(x = "Year of Health Violation", y = "Total Number of Rule Violations")
plotly::ggplotly(health_viol_plot)
# write.csv(health_viol, "./data/raw/tx_cws_rule_viols.csv")
```

Prior to 2015, the majority of health-based violations in Texas appear to be inorganic chemicals. After 2015, the majority of violations switched to stage 2 disinfectants & inorganic chemicals. Health-based violations peaked in 2015, but has decreased over time. 

What do the total number of health-based violations since 2010 look like spatially?
```{r fig.height = 9, fig.width = 10}
pwsid_viol_summary <- tx_viol_filt %>%
  # converting to compl begin date to date format, and grabbing year for 
  # grouping: 
  mutate(compl_begin_date = as.Date(compl_per_begin_date, 
                                    tryFormats = c("%m/%d/%Y")), 
         start_year = year(compl_begin_date)) %>%
  # grouping: 
  filter(start_year >= 2010) %>%
  group_by(pwsid) %>%
  summarize(total = length(unique(pwsid_viol_id))) %>%
  left_join(., tx_geoms, by = "pwsid") %>%
  st_as_sf()

# plotting 
viol_pal <- colorBin(
  bins = c(0, 5, 10, 50, 100, 500),
  palette = cont_palette(8),
  domain = pwsid_viol_summary$total)

leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron, 
                   group = "Toner Lite") %>%
  addPolygons(data = pwsid_viol_summary,
           fillColor = ~viol_pal(total),
              # slight opacity
              fillOpacity = 0.8,
              stroke = T, 
              color = "black",
              opacity = 0.2, 
              weight = 1,
              # smoothFactor = 0.5, 
              label = paste0("pwsid: ", pwsid_viol_summary$pwsid, 
                             "; Health Violations since 2010 : ",
                             pwsid_viol_summary$total)) %>%
  addLegend("bottomright",
            pal = viol_pal,
            values = pwsid_viol_summary$total,
            title = "Health Violations since 2010",
            opacity = 1)
```


## 3.3: What changes in water rates and MHI do we see in fragile systems? 
```{r fig.height = 9, fig.width = 10}
# order of operations: 
# filter pwsids to those that have been around for the entire time period of interest
# populations, MHI, and water rates for census tracts through time (even when boundaries have changed !)
# crosswalk data to pwids for each year (which might've not even existed at a certain time period...?) this also involves writing a whole new xwalk function for just a couple 
# of variables 

# summarize sdwis violation data by appropriate year bins 

# -----
# water rates: B25134 - only go back to 2021, and only in 1yr acs which
# "The 1-year ACS provides data for geographies with populations of 65,000 and greater."
# so it only has data for large counties
# all_census_vars <- load_variables(2020, "acs5", cache = TRUE)
# TODO - we have this var for 2021 and 2022 tract level for ACS 5yr!!!!!

census_vars <- c(total_pop = "B01003_001", 
                 mhi = "B19013_001", 
                 total_water = "B25134_003", 
                 less_125 = "B25134_004", 
                 between_125_249 = "B25134_005", 
                 between_250_499 = "B25134_006", 
                 between_500_749 = "B25134_007", 
                 between_750_999 = "B25134_008", 
                 over_1000 = "B25134_009")

counties_2023 <- tidycensus::get_acs(
  geography = "county", 
  variables = census_vars, 
  survey = "acs1",
  year = 2023
) 

counties_2022 <- tidycensus::get_acs(
  geography = "county", 
  variables = census_vars, 
  survey = "acs1",
  year = 2022
)

counties_2021 <- tidycensus::get_acs(
  geography = "county", 
  variables = census_vars, 
  survey = "acs1",
  year = 2021
)

# quick combining for county data: 
tx_2023 <- counties_2023 %>%
  filter(grepl("^48", GEOID)) %>%
  select(-moe) %>%
  pivot_wider(., names_from = variable, values_from = estimate) %>%
  mutate(acs1yr_year = "2023")
tx_2022 <- counties_2022 %>%
  filter(grepl("^48", GEOID)) %>%
  select(-moe) %>%
  pivot_wider(., names_from = variable, values_from = estimate) %>%
  mutate(acs1yr_year = "2022")
tx_2021 <- counties_2021 %>%
  filter(grepl("^48", GEOID)) %>%
  select(-moe) %>%
  pivot_wider(., names_from = variable, values_from = estimate) %>%
  mutate(acs1yr_year = "2021")
# 2020 had a low response rate
# yeah - can confirm we don't have water rate data prior to 2021
# counties_2019 <- tidycensus::get_acs(
#   geography = "county", 
#   variables = census_vars, 
#   survey = "acs1",
#   year = 2019
# )


# calculating % of total population paying for water in each $ category: 
all_rates <- bind_rows(tx_2023, tx_2022, tx_2021) %>%
  mutate_at(vars(!c("GEOID", "NAME", "total_pop", 
                    "mhi", "acs1yr_year", "total_water")), 
            ~((./total_water)*100))

# pivoting to long for plotting: 
all_rates_long <- all_rates %>%
  pivot_longer(., cols = less_125:over_1000) %>%
  mutate(mhi_cat = case_when(
    mhi < 40000 ~ "Less than $40,000", 
    mhi >= 40000 & mhi < 60000 ~ "$40,000 - $59,999",
    mhi >= 60000 & mhi < 80000 ~ "$60,000 - $79,999",
    mhi >= 80000 & mhi < 100000 ~ "$80,000 - $99,999",
    mhi >= 100000 & mhi < 120000 ~ "$100,000 - $120,000",
    mhi >= 120000 ~ "Over $120,000"
  ), 
  mhi_acs = paste0(mhi_cat, acs1yr_year))

# plotting: 
# ggplot(all_rates_long, aes(x = mhi_cat, y = value, 
#                            color = acs1yr_year, 
#                            fill = acs1yr_year, 
#                            group = mhi_acs)) + 
#   geom_point( alpha = 0.2, 
#               position = position_dodge(width = 0.75)) + 
#   geom_boxplot(alpha = 0.4, outliers = F) + 
#   facet_wrap(~name) + 
#   coord_flip() + 
#   epic_chart_theme + 
#   labs(y = "% of Rate Payers", 
#        x = "MHI Category")

# rates_barplot <- ggplot(all_rates_long, aes(x = value, y = mhi_cat, fill = name)) + 
#   geom_bar(stat = "identity") + 
#   # geom_smooth(se = F) + 
#   facet_wrap(~acs1yr_year, ncol = 1) +
#   epic_chart_theme +  
#   labs(y = "% of Rate Payers", 
#        x = "MHI")
# 
# plotly::ggplotly(rates_barplot) 

# looking at counties where there is a high % of rate payers paying >750, 
# but income is < 40,000
low_mhi_high_rates <- all_rates_long %>%
  filter(name %in% c("between_750_999", "over_1000") & mhi < 60000) %>%
  pivot_wider(., names_from = name, values_from = value) %>%
  mutate(total_750_1000 = (between_750_999 + over_1000))

counties_of_interest <- unique(low_mhi_high_rates$NAME) 
counties_tidy <- counties_of_interest %>%
  gsub(" County, Texas", "", .)

# which ones serve Starr county?
starr_cws <- tx_xwalk %>% 
  filter(county_served == "Starr") 
starr_pwsid <- unique(starr_cws$pwsid)


of_interest <- ggplot(low_mhi_high_rates, 
                      aes(x = mhi, y = total_750_1000, 
                          color = NAME, 
                          text = NAME)) + 
  geom_point() + 
  facet_wrap(~acs1yr_year, ncol = 1) + 
  labs(y = "% of Rate Payers Paying > $750 Annually", 
       x = "MHI") + 
  theme_minimal() 
plotly::ggplotly(of_interest, tooltip = "NAME") 
```

Unfortunately, annual water & sewer rate data are only available from the American Community Survey 1-year estimates, and for years 2021 - 2023. Because of this limitation, data is only available at the county level and for counties where the total population is greater than 60,000 people. Since the census returns the number of people paying various bins of annual rates (for example, the number of people that pay >$1,000 annually for water), water rates were estimated using a weighted mean. 

Counties where the median household income is less than $60,000 and the annual rate for water and sewer is greater than 750/year: `r counties_tidy`. Starr county, in particular, had counties where the MHI was less than 45,000 but nearly 40% of rate payers were spending more than 750/year for water and sewer. CWS that serve this county include: `r starr_pwsid`

```{r fig.height = 9, fig.width = 10}
# can I calculate mean water rates?
all_rates_calc <- bind_rows(tx_2023, tx_2022, tx_2021) 

# one way to do this is multiply the midpoint of each bin by # of people, 
# sum that information, and then divide by total number of people 
# NOTE - this is not perfect! and could be improved! 
all_rates_calc <- all_rates_calc %>%
  rowwise() %>%
  mutate(
    # finding total $ contributed by each bin using the bin's midpoint:
    less_125_mp = less_125*62,
    between_125_249_mp = between_125_249*187, 
    between_250_499_mp = between_250_499*374.5,
    between_500_749_mp = between_500_749*624.5, 
    between_750_999_mp = between_750_999*874.5, 
    # NOTE - assuming "midpoint" for >100 cat is just a half of a full
    # bin above 1000
    over_1000_mp = over_1000*1124.5,
    # summing total amount: 
    total = sum(less_125_mp, between_125_249_mp, between_250_499_mp,
                between_500_749_mp, between_750_999_mp, over_1000_mp), 
    # diving by total people to get a ROUGH mean
    mean_rate = total/total_water, 
    # finding pct of mhi spent on water annually
    rate_div_mhi = (mean_rate/mhi)*100
  )

# combine with census geos: 
tx_county_geo <- tx_county_pop %>%
  select(NAME)

all_rates_geo <- all_rates_calc %>%
  left_join(tx_county_geo) %>%
  st_as_sf() 

# plotting: 
rate_pal <- colorNumeric(
  palette = cont_palette(5),
  domain = all_rates_geo$rate_div_mhi)

leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron, 
                   group = "Toner Lite") %>%
    addPolygons(data = tx_county_pop,
              # slight opacity
              fillOpacity = 0.2,
              color = "gray",
              weight = 1, 
              smoothFactor = 0) %>%
  addPolygons(data = all_rates_geo[all_rates_geo$acs1yr_year == "2023",],
              fillColor = ~rate_pal(rate_div_mhi),
              # slight opacity
              fillOpacity = 0.8,
              color = "white",
              weight = 1, 
              smoothFactor = 0, 
              group = "2023", 
              label = paste0("County: ", all_rates_geo[all_rates_geo$acs1yr_year == "2023",]$NAME, 
                             "; % MHI spent on Water & Sewer : ",
                            round(all_rates_geo[all_rates_geo$acs1yr_year == "2023",]$rate_div_mhi,
                                  3), "%")) %>%
    addPolygons(data = all_rates_geo[all_rates_geo$acs1yr_year == "2022",],
              fillColor = ~rate_pal(rate_div_mhi),
              # slight opacity
              fillOpacity = 0.8,
              color = "white",
              weight = 1, 
              smoothFactor = 0, 
              group = "2022", 
              label = paste0("County: ", all_rates_geo[all_rates_geo$acs1yr_year == "2022",]$NAME, 
                             "; % MHI spent on Water & Sewer : ",
                            round(all_rates_geo[all_rates_geo$acs1yr_year == "2022",]$rate_div_mhi,
                                  3), "%")) %>%
      addPolygons(data = all_rates_geo[all_rates_geo$acs1yr_year == "2021",],
              fillColor = ~rate_pal(rate_div_mhi),
              # slight opacity
              fillOpacity = 0.8,
              color = "white",
              weight = 1, 
              smoothFactor = 0, 
              group = "2021", 
              label = paste0("County: ", all_rates_geo[all_rates_geo$acs1yr_year == "2021",]$NAME, 
                             "; % MHI spent on Water & Sewer : ",
                            round(all_rates_geo[all_rates_geo$acs1yr_year == "2021",]$rate_div_mhi, 3), "%")) %>%
  addLegend("bottomright",
            pal = rate_pal,
            values = all_rates_geo$rate_div_mhi,
            # if working with money or percentages, use the 
            # follow label formats: 
            # labFormat = labelFormat(prefix = "$"),
            labFormat = labelFormat(prefix = "%"),
            title = "% MHI spent on Water",
            opacity = 1) %>% 
  addLayersControl(baseGroups = c('2023', '2022', "2021"),
                   options = layersControlOptions(collapsed = FALSE),
                   position = 'bottomleft')
```

Across the years and counties in which we have data, counties closer to the southern border appear to spend a larger percentage of their income of water in comparison to the rest of the counties included in our study. Starr county had the largest percentage in our study, where 17% of income was going towards sewer and water in 2021. 


```{r fig.height = 9, fig.width = 10}
# plotting by % change from 2021 to 2023 
all_rates_geo_simple <- all_rates_geo %>%
  select(NAME, acs1yr_year, rate_div_mhi) %>%
  pivot_wider(., names_from = acs1yr_year, values_from = rate_div_mhi) %>%
  mutate(pct_change =  100*((`2023` - `2021`)/`2023`))
     
# plotting: 
pct_mhi_rate_pal <- colorBin(
  bins = c(-40, -20, -5, 0, 5, 20),
    palette = rev(c("#eab28e", "#f4cbb3", "white", "#b1dac9", "#8cc7b0")),
  # palette = rev(c("#d73027", "#fc8d59", "white", "#d9ef8b", "#91cf60")),
  domain = all_rates_geo_simple$pct_change)

leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron, 
                   group = "Toner Lite") %>%
  addPolygons(data = tx_county_pop,
              # slight opacity
              fillOpacity = 0.2,
              color = "gray",
              weight = 1, 
              smoothFactor = 0) %>%
  addPolygons(data = all_rates_geo_simple,
              fillColor = ~pct_mhi_rate_pal(pct_change),
              # slight opacity
              fillOpacity = 0.8,
              stroke = T, 
              color = "black",
              opacity = 0.2, 
              weight = 1,
              # smoothFactor = 0.5, 
              label = paste0("County: ", all_rates_geo_simple$NAME, 
                             "; % change: ",
                             round(all_rates_geo_simple$pct_change, 3))) %>%
  addLegend("bottomright",
            pal = pct_mhi_rate_pal,
            values = all_rates_geo_simple$pct_change,
            # if working with money or percentages, use the 
            # follow label formats: 
            # labFormat = labelFormat(prefix = "$"),
            labFormat = labelFormat(prefix = "%"),
            title = "% Change in % of Income Spent on Water",
            opacity = 1)
```

When looking at the percentage in the % of income spent on water from 2021 - 2023, most counties in our study period decreased over time, indicating that the relative financial burden of water on annual income has decreased. There are a few counties where financial burden has increased, particularly in more rural areas in NE Texas. The largest % change in our study period was Tom Green County with an increase of 18.6%. 


# 4.0: What are examples of success stories? 
## 4.1: What are the trends [for this specific CWS] in terms of population, violations, water rates, and investments? 

```{r fig.height = 9, fig.width = 10}
# building some better code that will work for all following sections: 

# grabbing key of pwsid and county served for water rate information: 
tx_county <- tx_xwalk %>%
  as.data.frame() %>%
  select(pwsid, county_served) %>% 
  mutate(county_served = paste0(county_served, " County, Texas"))

# water rates: 
tx_county_rates <- all_rates_calc %>% 
  select(NAME, acs1yr_year, mean_rate, rate_div_mhi)

# mergin' 
tx_county_rates_all <- merge(tx_county, tx_county_rates, all.y = T, 
                             by.x = "county_served", 
                             by.y = "NAME") %>% 
  rename(year = acs1yr_year, 
         water_rate = mean_rate, 
         pct_rate_mhi = rate_div_mhi)


# grabbing violation by pwsid and year: 
pwsid_viol_summary <- tx_viol_filt %>%
  # converting to compl begin date to date format, and grabbing year for 
  # grouping: 
  mutate(compl_begin_date = as.Date(compl_per_begin_date, 
                                    tryFormats = c("%m/%d/%Y")), 
         start_year = year(compl_begin_date)) %>%
  group_by(pwsid, start_year) %>%
  summarize(total_health_viols = length(unique(pwsid_viol_id))) %>%
  rename(year = start_year)

# violations and water rates are the only two datasets that MAY overlap 
# after 2020: 
pwsid_viol_rate_trends <- merge(pwsid_viol_summary, tx_county_rates_all, 
                                by = c("pwsid", "year"), all = T)


# grabbing funding history: 
pwsid_srf <- tx_srf %>% 
  select(pwsid, year, assistance_amt, prin_forgive_amt) %>%
  group_by(pwsid, year) %>%
  summarize(assistance_amt = sum(assistance_amt), 
            prin_forgive_amt = sum(prin_forgive_amt))

# tidying demog information: 
pwsid_pop_mhi <- tx_decade %>% 
  select(pwsid, estimate_total_pop_2010, estimate_mhi_2010_inf_adj, 
         estimate_total_pop_2015, estimate_mhi_2015_inf_adj, estimate_mhi_2020, 
         estimate_total_pop_2020) %>%
  pivot_longer(., cols = estimate_total_pop_2010:estimate_total_pop_2020) %>%
  mutate(year = case_when(
    grepl("2010", name) ~ "2010", 
    grepl("2015", name) ~ "2015", 
    TRUE ~ "2020"
  ), 
  demo = case_when(
    grepl("pop", name) ~ "total_pop", 
    TRUE ~ "mhi"
  )) %>%
  select(-name) %>%
  pivot_wider(., names_from = demo, values_from = value) 


# combining info: 
pwsid_combo <- merge(pwsid_pop_mhi, pwsid_viol_rate_trends, 
                     by = c("pwsid", "year"), all = T) %>%
  merge(., pwsid_srf, by = c("pwsid", "year"), all = T) 

# wanting to also add flags for % population and % MHI trends: 
pop_mhi_trends <- tx_decade %>%
  select(pwsid, pct_change_pop_2010_2020, pct_change_mhi_2010_2020)

# lol this is such a weird df but accurately captures how odd this data is to 
# work with: 
pwsid_trends <- merge(pwsid_combo, pop_mhi_trends, 
                            by = "pwsid", all.x = T) %>%
  mutate(type_flag = case_when(
    pct_change_pop_2010_2020 < 0 & pct_change_mhi_2010_2020 < 0 ~ "Pop_MHI_dec", 
    pct_change_pop_2010_2020 < 0 & pct_change_mhi_2010_2020 > 0 ~ "Pop_dec_MHI_inc", 
    pct_change_pop_2010_2020 > 0 & pct_change_mhi_2010_2020 < 0 ~ "Pop_inc_MHI_dec", 
    pct_change_pop_2010_2020 > 0 & pct_change_mhi_2010_2020 > 0 ~ "Pop_MHI_inc", 
    # these are either pwsids that are both tier 1 but missing MHI for one year
    # or instances where the tier of the crosswalk was different from 2010 and 
    # 2020 and therefore may not be fullllly comparable 
    TRUE ~ "incomplete_data"
  ), 
  funded = case_when(
    pwsid %in% pwsid_srf$pwsid ~ "YES", 
    TRUE ~ "NO"))

# filling in all county information: 
pwsid_trends_final <- merge(pwsid_trends, tx_county, 
              by = c("pwsid"), all.x = T) %>%
  select(-county_served.x) %>%
  rename(county_served = county_served.y) %>%
  relocate(county_served, .after = pwsid) %>%
  mutate(year = as.numeric(year))

# adding a long version for easy plotting: 
pwsid_trends_long <- pivot_longer(pwsid_trends_final, 
                                  cols = c(mhi, total_health_viols, 
                                           assistance_amt,
                                           total_pop))

# checking out trends for a specific county: 
county <- "Jefferson County, Texas"
county_trends <- pwsid_trends_long %>%
  filter(county_served == county) %>%
  filter(year >= 2000)

county_plot <- ggplot(county_trends, aes(x = year, y = value, 
                                         group = pwsid, 
                                         color = pwsid)) +
  geom_point() + 
  geom_smooth(se = F) + 
  theme_minimal() + 
  facet_grid(name~funded, scales = "free_y") + 
  theme(legend.position = "right",
        text = element_text(size = 13),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 11),
        axis.text.x = element_text(margin = margin(t = 10, r = 0,
                                                   b = 0, l = 0)),
        axis.title.x = element_text(margin = margin(t = 10, r = 0,
                                                    b = 0, l = 0)),
        axis.text.y = element_text(margin = margin(t = 0, r = 10,
                                                   b = 0, l = 0)),
        axis.title.y = element_text(margin = margin(t = 0, r = 10,
                                                    b = 0, l = 0))) +
  labs(x = "Year", y = "Value") + 
  ggtitle(paste0("CWS Trends for ", county))

ggplotly(county_plot)
``` 

In this interactive plot, you're seeing trends in median household income, health violations since 2000, and population trends for CWS that serve Jefferson County. CWS plotted on the left have not received SRF funds, and those plotting on the right have received SRF funds. Jefferson county, as indicated earlier, is a county where the relative financial burden of water has increased from 2021 - 2023. Let's take a closer look at the CWS that have received funding: 

```{r fig.height = 9, fig.width = 10}
funded_county <- county_trends %>%
  filter(funded == "YES")

# grabbing years of funding for each pwsid: 
pwsid_yr_funded <- funded_county %>%
  filter(name == "assistance_amt" & value != 0) %>%
  group_by(pwsid) %>%
  summarize(yr_funded = (unique(year)))

funded_county_filter <- funded_county %>%
  filter(name != "assistance_amt")  

jeff_county <- ggplot(funded_county_filter, aes(x = year, y = value, 
                                         group = pwsid, 
                                         color = pwsid)) +
  geom_point() + 
  geom_smooth(se = F) + 
  geom_vline(data = pwsid_yr_funded, 
             aes(xintercept = yr_funded, 
                 color = pwsid), 
             size=1, 
             linetype = "dashed") +
  theme_minimal() + 
  facet_wrap(~name, scales = "free_y", ncol = 1) + 
  ggtitle(paste0("Utilities Funded in ", county)) + 
  theme(legend.position = "right", 
        text = element_text(size = 13), 
        legend.text = element_text(size = 10), 
        legend.title = element_text(size = 11), 
        axis.text.x = element_text(margin = margin(t = 10, r = 0, 
                                                   b = 0, l = 0)), 
        axis.title.x = element_text(margin = margin(t = 10, r = 0, 
                                                    b = 0, l = 0)), 
        axis.text.y = element_text(margin = margin(t = 0, r = 10, 
                                                   b = 0, l = 0)), 
        axis.title.y = element_text(margin = margin(t = 0, r = 10, 
                                                    b = 0, l = 0))) +
  labs(x = "Year", y = "Value")
ggplotly(jeff_county)
```

Here, we can take a closer look at CWS that have received funding in Jefferson County, Texas. Colored dashed lines indicate the year in which a particular CWS received funding. Of the four plotted here, TX1230009 serves the largest population with the lowest MHI of ~35k. Based on our county-level rate data, the relative financial burden of water in Jefferson county has increased from 2021 to 2023. While health-based violations have been relatively low through time, the SRF funding this CWS received in 2015 may help support this CWS' large (and growing) but relatively low income community. 

Zooming in on CWS in a different county: 
```{r fig.height = 9, fig.width = 10}
# looking at pwsids in a particular county: 

# checking out trends for a specific county: 
county <- "Tarrant County, Texas"
county_trends <- pwsid_trends_long %>%
  filter(county_served == county) %>%
  filter(year >= 2000)

county_plot <- ggplot(county_trends, aes(x = year, y = value, 
                                         group = pwsid, 
                                         color = pwsid)) +
  geom_point() + 
  geom_smooth(se = F) + 
  theme_minimal() + 
  facet_grid(name~funded, scales = "free_y") + 
  theme(legend.position = "right",
        text = element_text(size = 13),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 11),
        axis.text.x = element_text(margin = margin(t = 10, r = 0,
                                                   b = 0, l = 0)),
        axis.title.x = element_text(margin = margin(t = 10, r = 0,
                                                    b = 0, l = 0)),
        axis.text.y = element_text(margin = margin(t = 0, r = 10,
                                                   b = 0, l = 0)),
        axis.title.y = element_text(margin = margin(t = 0, r = 10,
                                                    b = 0, l = 0))) +
  labs(x = "Year", y = "Value") + 
  ggtitle(paste0("CWS Trends for ", county))

ggplotly(county_plot)
```

Above, you can see the CWS trends for Tarrant county for comparisons. 

```{r fig.height = 9, fig.width = 10}
# zooming in a particular pwsid: 
funded_county <- county_trends %>%
  filter(funded == "YES")

# grabbing years of funding for each pwsid: 
pwsid_yr_funded <- funded_county %>%
  filter(name == "assistance_amt" & value != 0) %>%
  group_by(pwsid) %>%
  summarize(yr_funded = (unique(year)))

funded_county_filter <- funded_county %>%
  filter(name != "assistance_amt")  

tarrant_county <- ggplot(funded_county_filter, aes(x = year, y = value, 
                                         group = pwsid, 
                                         color = pwsid)) +
  geom_point() + 
  geom_smooth(se = F) + 
  geom_vline(data = pwsid_yr_funded, 
             aes(xintercept = yr_funded, 
                 color = pwsid), 
             size=1, 
             linetype = "dashed") +
  theme_minimal() + 
  facet_wrap(~name, scales = "free_y", ncol = 1) + 
  ggtitle(paste0("Utilities Funded in ", county)) + 
  theme(legend.position = "right", 
        text = element_text(size = 13), 
        legend.text = element_text(size = 10), 
        legend.title = element_text(size = 11), 
        axis.text.x = element_text(margin = margin(t = 10, r = 0, 
                                                   b = 0, l = 0)), 
        axis.title.x = element_text(margin = margin(t = 10, r = 0, 
                                                    b = 0, l = 0)), 
        axis.text.y = element_text(margin = margin(t = 0, r = 10, 
                                                   b = 0, l = 0)), 
        axis.title.y = element_text(margin = margin(t = 0, r = 10, 
                                                    b = 0, l = 0))) +
  labs(x = "Year", y = "Value")
ggplotly(tarrant_county)
```

The colored dashed line indicates the year in which the utility received funding. For TX2200069, violations dipped after the utility received funding (no reported violations since 2017), populations continued to increase, and MHI started increasing since 2015. Utilities serving Tarrant county have also exhibited a decrease in the % of income spent on MHI from 2021-2023, indicating either lower rates or higher MHI over that time period. 


# 5.0 What are examples of cautionary tales?
## 5.1: What are the trends [for this specific CWS] in terms of population, violations, water rates, and investments? 
```{r fig.height = 9, fig.width = 10}
# grabbing high violating CWS: 
high_viols <- pwsid_trends_final %>% 
  filter(year >= 2005) %>%
  group_by(pwsid) %>%
  summarize(total_viols = sum(total_health_viols, na.rm = T)) %>% 
  filter(total_viols > 50)

# zooming in on utilities that have been declining in pop and MHI by >= 10%, 
# have MHI <= 60,000, and have not received funding 
pwsid_caution <- pwsid_trends_final %>%
  filter(type_flag %in% c("Pop_MHI_dec", "Pop_inc_MHI_dec")) %>%
  filter(funded == "NO") %>%
  filter(abs(pct_change_pop_2010_2020) >= 10) %>%
  filter(abs(pct_change_mhi_2010_2020) >= 10) %>%
  filter(year >= 2005) %>%
  filter(pwsid %in% high_viols$pwsid)

# adding a long version for easy plotting: 
caution_trends_long <- pivot_longer(pwsid_caution, 
                                    cols = c(mhi, total_health_viols, 
                                             total_pop))

# checking out trends:
caution_trends <- caution_trends_long 

caution_plot <- ggplot(caution_trends, aes(x = year, y = value, 
                                         group = pwsid, 
                                         color = pwsid)) +
  geom_point() + 
  geom_smooth(se = F) + 
  facet_wrap(~name, scales = "free_y", ncol = 1) + 
  labs(x = "Year", y = "Value") + 
  theme_minimal() + 
  theme(legend.position = "right", 
        text = element_text(size = 13), 
        legend.text = element_text(size = 10), 
        legend.title = element_text(size = 11), 
        axis.text.x = element_text(margin = margin(t = 10, r = 0, 
                                                   b = 0, l = 0)), 
        axis.title.x = element_text(margin = margin(t = 10, r = 0, 
                                                    b = 0, l = 0)), 
        axis.text.y = element_text(margin = margin(t = 0, r = 10, 
                                                   b = 0, l = 0)), 
        axis.title.y = element_text(margin = margin(t = 0, r = 10, 
                                                    b = 0, l = 0))) +  ggtitle(paste0("Trends for CWS in potential need of Support"))

ggplotly(caution_plot)
```

Chart above shows utilizes where populations and/or MHI have been decreasing from 2010 - 2020, have not received funding, and have more than 50 total health violations from 2005 - 2024. TX0030030 located in Zavalla TX, for example, has been increasing in violations since 2017, had 22 health-based violations in 2022, and has the lowest (& declining!) MHI out of our collection of CWS, and moderate population loss. These characteristics may represent a CWS in need of funding. 

```{r fig.height = 9, fig.width = 10}
of_caution_geo <- tx_xwalk %>% 
  filter(pwsid %in% caution_trends$pwsid)

leaflet() %>%
    addProviderTiles(providers$CartoDB.Positron, 
                   group = "Toner Lite") %>%
  addPolygons(data = of_caution_geo,
              # slight opacity
              fillOpacity = 0.8,
              color = "red",
              weight = 1, 
              smoothFactor = 0, 
              label = paste0("pwsid: ", of_caution_geo$pwsid))
```

Quite a few located East of Waco, around Livingston, and Zavalla TX. 

# 6.0 What are the overall trends of CWS since 2010 in terms of population, MHI, violations, and funding?
```{r fig.height = 9, fig.width = 10}
# can I just build a summary df of the main trends we care about?
# thinking about - pop change, total violations, mhi change, total funding
tx_viol_trends <- tx_viol_rulecode %>%
  mutate(start_date = as.Date(viol_first_reported_date, 
                              tryFormats = c("%m/%d/%Y")), 
         start_year = year(start_date)) %>%
  filter(is_health_based_ind == "Y") %>%
  # filtering to the years we have
  filter(start_year >= 2010 & start_year <= 2020) %>%
  group_by(pwsid) %>%
  summarize(health_viols_since_2010 = length(unique(violation_id)))

# funding: 
tx_srf_trends <- tx_srf %>%
    # filtering to the years we have
  filter(year >= 2010) %>%
  group_by(pwsid) %>%
  summarize(times_funded_since_2010 = n(), 
            total_assistance = sum(assistance_amt), 
            total_pf = sum(prin_forgive_amt))

tx_summary_trends <- tx_decade %>%
  select(pwsid, 
         pct_change_pop_2010_2020:pct_change_mhi_2010_2020) %>%
  left_join(tx_viol_trends, by = "pwsid") %>%
  left_join(tx_srf_trends, by = "pwsid") %>%
  mutate_if(is.numeric, ~replace_na(., 0)) %>%
  mutate(pct_change_pop_2010_2020 = round(pct_change_pop_2010_2020, 2), 
         pct_change_mhi_2010_2020 = round(pct_change_mhi_2010_2020, 2))

# table for people to explore: 
DT::datatable(tx_summary_trends)
```

# 7.0 Violation summary stats
```{r message = F, results='hide'}
# Taking code from Oregon and running summary stats for TX
# on violations: 

# breaking data and summarizing: 
# 4,616 SABs turns to 4,559 --> removed 57 SABs
tx_viols <- tx_xwalk %>%
  filter(estimate_hh_below_pov_per >= 0) %>%
  filter(!is.na(estimate_mhi)) %>%
  filter(!is.na(estimate_hh_below_pov_per)) %>%
  as.data.frame() 
         
### % of households below poverty ###
pov_viols <- tx_viols %>%
  # eight have % below 0, likely due to errors or difficulties associated with interpolation
  mutate(pct_hh_below_pov_bin = cut(estimate_hh_below_pov_per, 
                                    breaks = c(0, 5, 10, 20, 
                                               30, 40, 50, 60), 
                                    include.lowest = T))

# summarize by binned values
pov_viols %>%
  group_by(pct_hh_below_pov_bin) %>%
  summarize(total_sabs = n(), 
            mean_health_viols_10yr = mean(health_viols_10yr), 
            sd_health_viols_10yr = sd(health_viols_10yr), 
            min = min(health_viols_10yr), 
            max = max(health_viols_10yr)) %>%
  rename("% Househlds Below Poverty" = pct_hh_below_pov_bin, 
         "Total SABs" = total_sabs, 
         "Mean Health Violations, 10yr" = mean_health_viols_10yr,
         "SD Health Violations, 10yr" = sd_health_viols_10yr)


# summarize by cutoff
pov_viols %>%
  filter(estimate_hh_below_pov_per < 15) %>%
  summarize(total_sabs = n(), 
            mean_below_15 = mean(health_viols_10yr), 
            sd_below_15 = sd(health_viols_10yr), 
            min = min(health_viols_10yr), 
            max = max(health_viols_10yr))

pov_viols %>%
  filter(estimate_hh_below_pov_per >= 15) %>%
  summarize(total_sabs = n(),
            mean_greater_equal_15 = mean(health_viols_10yr), 
            sd_greater_equal_15 = sd(health_viols_10yr), 
            min = min(health_viols_10yr), 
            max = max(health_viols_10yr))


### Median household income ### 
mhi_viols <- tx_viols %>%
  # eight have % below 0, likely due to errors or difficulties associated with interpolation
  mutate(mhi_bin = cut(estimate_mhi, breaks = c(0, 25000, 50000, 75000, 
                                                100000, 120000, 250001), 
                       labels = c("$0-25,000", "$25,001-50,000", 
                                  "$50,001-75,000", "$75,001-100,000", 
                                  "$100,001-120,000", "$120,001-250,001"),
                       include.lowest = T))


# summarize by binned mhi
mhi_viols %>%
  group_by(mhi_bin) %>%
  summarize(total_sabs = n(), 
            mean_health_viols_10yr = mean(health_viols_10yr), 
            sd_health_viols_10yr = sd(health_viols_10yr), 
            min = min(health_viols_10yr), 
            max = max(health_viols_10yr)) %>%
  rename("Median Household Income" = mhi_bin,
         "Total SABs" = total_sabs, 
         "Mean Health Violations, 10yr" = mean_health_viols_10yr, 
         "SD Health Violations, 10yr" = sd_health_viols_10yr, 
         "Minimum Health Violations, 10yr" = min, 
         "Maximum Health Violations, 10yr" = max)

# summarize by cutoff: 
mhi_viols %>%
  filter(estimate_mhi < 50000) %>%
  summarize(total_sabs = n(), 
            mean_below_50 = mean(health_viols_10yr), 
            sd_below_50 = sd(health_viols_10yr), 
            min = min(health_viols_10yr), 
            max = max(health_viols_10yr))
# open viols?
mhi_viols %>%
  filter(estimate_mhi < 50000) %>%
  filter(open_health_viol == "Yes") %>%
  tally()


mhi_viols %>%
  filter(estimate_mhi >= 50000) %>%
  summarize(total_sabs = n(),
            mean_greater_equal_50 = mean(health_viols_10yr), 
            sd_greater_equal_50 = sd(health_viols_10yr), 
            min = min(health_viols_10yr), 
            max = max(health_viols_10yr))
# open viols?
mhi_viols %>%
  filter(estimate_mhi >= 50000) %>%
  filter(open_health_viol == "Yes") %>%
  tally()



# fonts: 
sysfonts::font_add_google("Lato")
showtext::showtext_auto()

# palettes: 
# continuous palettes - based on EPIC's primary colors: 
cont_palette <- colorRampPalette(c("#172f60","#4ea324"))
cat_palette <- colorRampPalette(c("#172f60","#1054a8",
                                  "#791a7b","#de9d29", 
                                  "#b15712","#4ea324"))
# plotting: 
epic_chart_theme <- theme_minimal() + 
  theme(legend.position = "right", 
        text = element_text(size = 13, family = "Lato"), 
        legend.text = element_text(size = 10, family = "Lato"), 
        legend.title = element_text(size = 11, family = "Lato"), 
        axis.text.x = element_text(margin = margin(t = 10, r = 0, 
                                                   b = 0, l = 0)), 
        axis.title.x = element_text(margin = margin(t = 10, r = 0, 
                                                    b = 0, l = 0)), 
        axis.text.y = element_text(margin = margin(t = 0, r = 10, 
                                                   b = 0, l = 0)), 
        axis.title.y = element_text(margin = margin(t = 0, r = 10, 
                                                    b = 0, l = 0))) 

ggplot(mhi_viols, aes(x = estimate_mhi, 
                      y = estimate_hh_below_pov_per, 
                      size = health_viols_10yr, 
                      color = health_viols_10yr)) + 
  geom_point(alpha = 0.7) +
  scale_colour_gradientn(colours = cont_palette(3), 
                         name = "Health Violations \n10yr") + 
  epic_chart_theme


# trying to facet to clean up relationships:
library(ggridges)

df_plotting <- mhi_viols %>%
  select(pwsid, estimate_mhi, estimate_hh_below_pov_per, 
         health_viols_10yr) %>%
  pivot_longer(., cols = c(estimate_mhi, estimate_hh_below_pov_per)) %>%
  mutate(clean_name = case_when(
    name == "estimate_mhi" ~ "Median Household Income ($)", 
    TRUE ~ "% Households Below Poverty"
  ))

ggplot(df_plotting, aes(x = value, 
                        y = health_viols_10yr, 
                        color = clean_name)) + 
  geom_jitter(width = 0.001, size = 3, alpha = 0.7) + 
  scale_colour_manual(values = cat_palette(2), 
                      name = "") + 
  facet_wrap(~clean_name, scales = "free_x", ncol = 1) + 
  epic_chart_theme + 
  theme(legend.position = "none") + 
  labs(x = "", y = "Health Violations, 10 Years")

```

# 8.0 Bonus water affordability analysis 
```{r}
library(sf)
library(leaflet)

# read in latest xwalk with water rates
xwalk_new <- aws.s3::s3read_using(st_read, 
                                  object = "/service_area_boundaries/epa-sabs/epa-sabs-crosswalk-acs2021.geojson",
                                  bucket = "tech-team-data",
                                  quiet = TRUE)

# filter for those in TX project: 
tx_xwalk_new <- xwalk_new %>%
  filter(pwsid %in% tx_xwalk$pwsid)

# mutating to get estimate water rates using a weighted mean 
wr <- tx_xwalk_new %>%
  select(pwsid, mhi, total_water_rates, health_viols_10yr, 
         water_rate_less_125:water_rate_over_1000) %>%
  rowwise() %>%
  mutate(
    # finding total $ contributed by each bin using the bin's midpoint:
    less_125_mp = water_rate_less_125*62,
    between_125_249_mp = water_rate_between_125_249*187, 
    between_250_499_mp = water_rate_between_250_499*374.5,
    between_500_749_mp = water_rate_between_500_749*624.5, 
    between_750_999_mp = water_rate_between_750_999*874.5, 
    # NOTE - assuming "midpoint" for >100 cat is just a half of a full
    # bin above 1000
    over_1000_mp = water_rate_over_1000*1124.5,
    # summing total amount: 
    total = sum(less_125_mp, between_125_249_mp, between_250_499_mp,
                between_500_749_mp, between_750_999_mp, over_1000_mp), 
    # diving by total people to get a ROUGH mean
    mean_rate = total/total_water_rates, 
    # finding pct of mhi spent on water annually
    rate_div_mhi = (mean_rate/mhi)*100
  )


# mappin'
cont_palette <- colorRampPalette(c("#172f60","#4ea324"))

rate_pal <- colorBin(
  bins = c(0, 0.5, 0.75, 1, 3, 6),
  palette = cont_palette(5),
  domain = wr$rate_div_mhi)

leaflet() %>%
  addProviderTiles(providers$CartoDB.Positron, 
                   group = "Toner Lite") %>%
  addPolygons(data = wr,
              fillColor = ~rate_pal(rate_div_mhi),
              # slight opacity
              fillOpacity = 0.95,
              color = "black",
              weight = 1, 
              smoothFactor = 0, 
              label = paste0("% MHI spent on Water & Sewer : ",
                             round(wr$rate_div_mhi, 3), "%")) %>%
    addLegend("bottomright",
              pal = rate_pal,
              values = wr$rate_div_mhi,
              # if working with money or percentages, use the 
              # follow label formats: 
              # labFormat = labelFormat(prefix = "$"),
              labFormat = labelFormat(prefix = "%"),
              title = "% MHI spent on Water",
              opacity = 1) 


```
