---
title: "Texas Drinking Water Data Collating & Merging"
author: "EmmaLi Tsai"
date: "2024-02-09"
output: html_document
---

Packages: 
```{r packages}
library(tidyverse)
library(leaflet)
library(sf)
library(aws.s3) 
library(janitor)
library(tidycensus)
library(areal)
library(mapview)
library(httr)
library(readxl)
library(jsonlite)
library(googlesheets4)
library(googledrive)
# remotes::install_github("environmental-policy-innovation-center/stateDW", 
#                         auth_token = "insert-github-pat-here")
library(stateDW)

# Store keys on local machine in .Renviron file! 
```

Functions: 
```{r functions}
# Migrated to R package workflow but keeping these here as a backup 
# source("./code/functions/get_sab.R")
# source("./code/functions/get_SDWIS.R")
# source("./code/functions/interp_sab_census.R")
```

_______________________________________________________________________________
### Data Collating - pulling data from APIs, web links, or S3 buckets:  Data here are either pulled from S3 (see code above for how they got there) or downloaded easily from websites or APIs
_______________________________________________________________________________
## Demographic
```{r data collating}
## Demographic #################################################################
# SABs
sab_census <- aws.s3::s3read_using(st_read, 
                           object = "state-drinking-water/TX/raw/TX_sab_census.geojson",
                           bucket = "tech-team-data")
# CEJST
cejst_raw <- aws.s3::s3read_using(read.csv, object = "s3://tech-team-data/ej-data/cejest-v1/1.0-communities.csv") %>%
  janitor::clean_names() %>%
  filter(state_territory == "Texas") %>%
  select(c("census_tract_2010_id","total_threshold_criteria_exceeded",
           "total_categories_exceeded","identified_as_disadvantaged",
           "percentage_of_tract_that_is_disadvantaged_by_area","total_population"))%>%
  rename("total_population_cejst" = "total_population")

# Social Vulnerability Index 
svi <- aws.s3::s3read_using(read.csv, 
                            object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_SVI_2020.csv") %>%
  janitor::clean_names()

# Climate Vulnerability Index: 
cvi <- aws.s3::s3read_using(read.csv, 
                            object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_CVI.csv")

# TX EJ Screen: 
tx_ejscreen <- aws.s3::s3read_using(read.csv, 
                                    object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_ejscreen.csv")

# Adding this into a larger list: 
TX_demographic <- list()
TX_demographic[[1]] <- sab_census
TX_demographic[[2]] <- cejst_raw
TX_demographic[[3]] <- tx_ejscreen
TX_demographic[[4]] <- svi
TX_demographic[[5]] <- cvi
names(TX_demographic) <- c("census", "cejst", "ejscreen", "svi", "cvi")
saveRDS(TX_demographic, file="./data/raw/TX_demographic_list.RData")
```
## Environmental
```{r}
## Environmental ###############################################################
## Surface water intake locations: https://gis-tceq.opendata.arcgis.com/datasets/TCEQ::surface-intakes/api
# Pulled Jan 22 2024 
sw_intake <- GET("https://gisweb.tceq.texas.gov/arcgis/rest/services/Public/PWS/MapServer/1/query?where=1%3D1&outFields=*&outSR=4326&f=json")
sw_intake_tidy <- fromJSON(rawToChar(sw_intake$content))$features$attribute %>%
  as.data.frame() %>%
  janitor::clean_names() %>%
  rename(pwsid = pws_id) %>%
  mutate(pwsid = paste0("TX", pwsid))

## groundwater quality & levels: 
# function for reading gwdb tables, which have unique separators: 
read_gwdb_table <- function(x) {
     read.table(x, header = TRUE, fill = TRUE, sep = "|")
}
# grabbing well metadata: 
well_main <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WellMain.txt") %>%
  janitor::clean_names()

# water quality information: 
wq_major <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WaterQualityMajor.txt") %>%
  janitor::clean_names()
wq_minor <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WaterQualityMinor.txt") %>%
  janitor::clean_names()
wq_other <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WaterQualityOtherUnassigned.txt") %>%
  janitor::clean_names()
# let's just combine all of these 
well_wq <- rbind(wq_major, wq_minor, wq_other)

# water level information: 
wl_major <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WaterLevelsMajor.txt") %>%
  janitor::clean_names()
wl_minor <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WaterLevelsMinor.txt") %>%
  janitor::clean_names()
wl_other <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WaterLevelsOtherUnassigned.txt") %>%
  janitor::clean_names()
# let's just combine all of these 
well_wl <- rbind(wl_major, wl_minor, wl_other)

## Major aquifers: https://www.twdb.texas.gov/mapping/gisdata/doc/major_aquifers.zip
# downloaded 1/23/2024
file_loc <- "./data/raw/TX_major_aquifers"
download.file("https://www.twdb.texas.gov/mapping/gisdata/doc/major_aquifers.zip", 
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
file.remove(paste0(file_loc, ".zip"))
maj_aquifers <- st_read(paste0(file_loc, "/NEW_major_aquifers_dd.shp")) %>%
  janitor::clean_names()

## Minor aquifers: https://www.twdb.texas.gov/mapping/gisdata/doc/minor_aquifers.zip
# downloaded 1/23/2024
file_loc <- "./data/raw/TX_minor_aquifers"
download.file("https://www.twdb.texas.gov/mapping/gisdata/doc/minor_aquifers.zip", 
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
file.remove(paste0(file_loc, ".zip"))
minor_aquifers <- st_read(paste0(file_loc, "/Minor_Aquifers.shp")) %>%
  janitor::clean_names()

## Major river basins: https://www.twdb.texas.gov/mapping/gisdata/doc/Major_River_Basins_Shapefile.zip
# downloaded 1/23/2024
file_loc <- "./data/raw/TX_major_river_basins"
download.file("https://www.twdb.texas.gov/mapping/gisdata/doc/Major_River_Basins_Shapefile.zip", 
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
file.remove(paste0(file_loc, ".zip"))
major_river_basins <- st_read(paste0(file_loc, "/TWDB_MRBs_2014.shp")) %>%
  janitor::clean_names()

## Existing reservoirs: https://www.twdb.texas.gov/mapping/gisdata/doc/Existing_Reservoirs.zip
# downloaded 1/23/2024
file_loc <- "./data/raw/TX_reservoirs"
download.file("https://www.twdb.texas.gov/mapping/gisdata/doc/Existing_Reservoirs.zip", 
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
file.remove(paste0(file_loc, ".zip"))
reservoirs <- st_read(paste0(file_loc, "/TWDB_SWP2012_Major_Reservoirs.shp")) %>%
  janitor::clean_names()

## Major rivers: https://www.twdb.texas.gov/mapping/gisdata/doc/Major_Rivers_dd83.zip
# downloaded 2/14/2024
file_loc <- "./data/raw/TX_rivers"
download.file("https://www.twdb.texas.gov/mapping/gisdata/doc/Major_Rivers_dd83.zip", 
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
file.remove(paste0(file_loc, ".zip"))
rivers <- st_read(paste0(file_loc, "/MajorRivers_dd83.shp")) %>%
  janitor::clean_names()

## Surface water 303(d) river basin impairment: https://texaswaterexplorer.tnc.org/index.html
# downloaded 1/24/2024
file_loc <- "./data/raw/TX_basin_impairment_2012"
download.file("https://texaswaterexplorer.tnc.org/downloads/Data_WQImpairment.zip", 
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
file.remove(paste0(file_loc, ".zip"))
basin_impairment <- readxl::read_excel(paste0(file_loc, "/Data_WQImpairment.xlsx"), sheet = "Basins")%>%
  janitor::clean_names()

# reservoir levels - recent
recent_reservoir_levels <- aws.s3::s3read_using(read.csv, 
                                         object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_reservoir_levels_recent.csv") %>%
  select(-X)

# reservoir levels - historic
reservoir_levels_10yr <- aws.s3::s3read_using(read.csv, 
                                         object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_reservoir_levels_10yr.csv") %>%
  select(-X)

# groundwater levels
recent_well_levels <- aws.s3::s3read_using(read.csv, 
                                           object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_recent_well_levels.csv") %>%
  select(-X)

# Adding this into a larger list
# surface water: 
TX_enviro_surface <- list()
TX_enviro_surface[[1]] <- sw_intake_tidy
TX_enviro_surface[[2]] <- rivers
TX_enviro_surface[[3]] <- reservoirs
TX_enviro_surface[[4]] <- recent_reservoir_levels
TX_enviro_surface[[5]] <- reservoir_levels_10yr
TX_enviro_surface[[6]] <- major_river_basins
TX_enviro_surface[[7]] <- basin_impairment
names(TX_enviro_surface) <- c("surface_water_intakes", "rivers", "reservoirs",
                              "reservoir_levels_recent", "reservoir_levels_10yr", 
                              "major_river_basins", "impaired_basins")
saveRDS(TX_enviro_surface, file="./data/raw/TX_enviro_surface_list.RData")


# ground water: 
TX_enviro_ground <- list()
TX_enviro_ground[[1]] <- minor_aquifers
TX_enviro_ground[[2]] <- maj_aquifers
TX_enviro_ground[[3]] <- well_main
TX_enviro_ground[[4]] <- recent_well_levels
TX_enviro_ground[[5]] <- well_wl
TX_enviro_ground[[6]] <- well_wq
names(TX_enviro_ground) <- c("minor_aquifers", "major_aquifers",
                              "well_metadata",
                              "well_levels_recent", "well_levels_historic", 
                              "well_quality_historic")
saveRDS(TX_enviro_ground, file="./data/raw/TX_enviro_ground_list.RData")
```
## Water Delivery System
```{r}
## Water Delivery System ######################################################
# historic use estimates: 
historic_use_estimates <-  aws.s3::s3read_using(read.csv, 
                                                object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_historic_use_estimates.csv") %>%
  select(-X)
# need to fix the pwsid column: 
for(i in 1:nrow(historic_use_estimates)){
  historic_use_estimates$pwsid[i] <- paste0("TX", paste(rep(0, (7-nchar(historic_use_estimates$tceq_pws_code[i]))), 
                                                        collapse = ""), 
                                            historic_use_estimates$tceq_pws_code[i], collapse = "")
}

# water loss audits
# vector of cleaned column names 
waterlosscolnames <- c("Name of Utility",
                       "Real Loss GMD (<32 conn/mi)",
                       "Real Loss GCD",
                       "Apparent Loss GCD",
                       "Water Loss GCD",
                       "ILI (>= 3,000 connections)", 
                       "Total GPCD", 
                       "GPCD Loss",
                       "Real Loss Cost in dollars",
                       "Apparent Loss Cost in dollars","Year")

#making empty dataframe 
waterloss_all <- setNames(data.frame(matrix(ncol = length(waterlosscolnames), nrow = 0)), waterlosscolnames)

year <- 2010 

# ! edit this when adding additional years 
stopyear <- 2022 

while(year <= stopyear)
{
print(year)
waterloss_raw <- aws.s3::s3read_using(read_csv, 
                               object = paste0("s3://tech-team-data/state-drinking-water/TX/raw/TCEQ_Loss_Audits/SummaryAuditsByCategory_",year,".csv"))

waterloss <- waterloss_raw %>%
             select(1:10)

colnames(waterloss) <- waterlosscolnames

waterloss <- waterloss %>%
             janitor::clean_names()%>%
             filter(name_of_utility != "UtilityName")%>%
             mutate(Year = year)

waterloss_all <- rbind(waterloss_all,waterloss)
year <- year + 1
}

## 2022 TX State Water Plan: https://2022.texasstatewaterplan.org/statewide?
# downloaded Jan 29 2024
pop <- read.csv("https://2022.texasstatewaterplan.org/download/statewide/population") %>%
  janitor::clean_names()
demands <- read.csv("https://2022.texasstatewaterplan.org/download/statewide/demands") %>%
  janitor::clean_names()
supp <- read.csv("https://2022.texasstatewaterplan.org/download/statewide/supplies") %>%
  janitor::clean_names()
needs <- read.csv("https://2022.texasstatewaterplan.org/download/statewide/needs") %>%
  janitor::clean_names()
strat <- read.csv("https://2022.texasstatewaterplan.org/download/statewide/strategies") %>%
  janitor::clean_names()

# TCEQ DWW system data: 
# NOTE: there is A TON more information in s3 for us to pull like enforcement 
# actions, etc. but starting the structure here for grabbing that information
# pulling data from two different webscraping events and locating pwsids that 
# are community water systems: 
water_details_dec17 <- aws.s3::s3read_using(read.csv, 
                                      object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_TCEQ_DWW/TX_20231217/WaterSystemDetail/TX_Water System Details_20231217_20231229203427.csv") 
water_details_dec3rd <- aws.s3::s3read_using(read.csv, 
                                      object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_TCEQ_DWW/TX_20231203/WaterSystemDetail/TX_Water System Details_20231203.csv") 
combined_details <- rbind(water_details_dec17, water_details_dec3rd)
water_details <- pivot_wider(combined_details, names_from = "Type", values_from = "Value") %>%
  janitor::clean_names()
# grabbing just community water systems (tceq has 10k that are not cws): 
tceq_cws <- water_details %>%
  filter(system_type == "C") # 4666

# pulling violations: 
viol_dec17 <- aws.s3::s3read_using(read.csv, 
                                      object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_TCEQ_DWW/TX_20231217/Violations/TX_Individual Violations_20231217_20231229203427.csv") 
viol_dec3rd <- aws.s3::s3read_using(read.csv, 
                                      object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_TCEQ_DWW/TX_20231203/Violations/TX_Individual Violations_20231203.csv") 
combined_viols <- rbind(viol_dec17, viol_dec3rd) %>%
  janitor::clean_names() %>%
  filter(water_system_no %in% tceq_cws$water_system_no) %>% # 118,659
  rename(pwsid = water_system_no)
# SDWIS contains 20,000 more violations

# TX SDWIS violations: 
sdwis <- aws.s3::s3read_using(read.csv, 
                                      object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_sdwis.csv") %>%
  select(-X)

# throwing this into a list: 
TX_water_delivery <- list()
TX_water_delivery[[1]] <- sdwis
TX_water_delivery[[2]] <- combined_viols
TX_water_delivery[[3]] <- historic_use_estimates
TX_water_delivery[[4]] <- waterloss_all
TX_water_delivery[[5]] <- pop 
TX_water_delivery[[6]] <- demands
TX_water_delivery[[7]] <- supp
TX_water_delivery[[8]] <- needs
TX_water_delivery[[9]] <- strat

names(TX_water_delivery) <- c("sdwis", "tceq_viols",
                              "historic_use_estimates",
                              "water_loss", "TXSWP_pop", "TXSWP_demands",
                              "TXSWP_supp", "TXSWP_needs", "TXSWP_strat")

saveRDS(TX_water_delivery, file="./data/raw/TX_water_delivery_list.RData")
```
## Financial 
```{r}
## Financial ##################################################################
## Water affordability - nixed b/c folded into the TX DAC definition - which 
# incorporates TX water bills with mhi to estimate household cost factor 
# https://github.com/NIEPS-Water-Program/water-affordability/raw/main/data/rates_data/rates_tx.xlsx
# downloaded Jan 24 2024
# link <- "https://github.com/NIEPS-Water-Program/water-affordability/raw/main/data/rates_data/rates_tx.xlsx"
# temp_file <- tempfile(fileext = ".xlsx")
# water_aff <- GET(link, write_disk(path = temp_file))
# water_aff <- readxl::read_excel(temp_file) %>%
#   janitor::clean_names()

# PPL data & tidying 
# TODO: Update this when Phil finishes scraping TX's PPL
TX_PPL <- aws.s3::s3read_using(read.csv, 
                               object = "s3://water-team-data/clean_data/srf_project_priority_lists/web_ppl_combined_clean.csv") %>%
  mutate(across('Project.Type', 
                str_replace, 'Other', 'General')) %>%
  filter(State == "Texas") %>%
  janitor::clean_names()

## Danielle's SRF analysis: had to download it using a different method because 
# it's an excel file that's stored on google drive - so it's not actually a 
# google sheet. 
# downloaded 2/1/2024 
URL <- "https://docs.google.com/spreadsheets/d/1kUHLdzFjwmKx4r6YTu2qjxwNuuG5JeUi/edit#gid=252194392"
dl <- googledrive::drive_download(
  as_id(URL),
  path = 'temp1.xlsx', 
  type = "xlsx")
cwsrf_by_rank <- readxl::read_excel('temp1.xlsx', sheet = "CWSRF By Rank", 
                                        skip = 1, col_names = TRUE) %>%
  janitor::clean_names() %>%
  # selecting columns of interest: 
  select(rank:sfy)
file.remove("temp1.xlsx")

## TX DAC - whether they are likely to qualify for DAC status - requires pulling water affordability data and combining this with mhi
TX_DAC <- aws.s3::s3read_using(st_read, 
                               object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_DAC.geojson")

# throwing this into a list: 
TX_financial <- list()
TX_financial[[1]] <- TX_DAC
TX_financial[[2]] <- TX_PPL
TX_financial[[3]] <- cwsrf_by_rank
names(TX_financial) <- c("TX_affordability_DAC", "TX_PPL",
                         "SRF_analysis")

saveRDS(TX_financial, file="./data/raw/TX_financial_list.RData")
```

## Loading lists 
```{r}
# reading demographic, environmental (gw and sw), water delivery system (wds)
# and financial lists generated above: 
demo <- readRDS("./data/raw/TX_demographic_list.RData")
enviro_sw <- readRDS("./data/raw/TX_enviro_surface_list.RData")
enviro_gw <- readRDS("./data/raw/TX_enviro_ground_list.RData")
wds <- readRDS("./data/raw/TX_water_delivery_list.RData")
fin <- readRDS("./data/raw/TX_financial_list.RData")
```

## Merging: surface water (reservoirs, rivers, basins) to pwsids:
```{r}
# merging impaired water bodies with basin information: 
basins <- enviro_sw$major_river_basins %>% 
  left_join(., enviro_sw$impaired_basins, by = "basin_num") %>%
  rename(basin_name_short = basin_name.x, 
         basin_name_long = basin_name.y) %>%
  select(!c("id", "label_no"))

# st_intersection with reservoir boundaries to identify which reservoirs 
# are in what basin: 
res <- enviro_sw$reservoirs %>%
  # dropping z coordinates: 
  st_zm(., drop = TRUE) %>%
  st_as_sf(.) %>%
  st_transform(., crs = st_crs(basins)) 
# switch off spherical geometry 
sf_use_s2(FALSE)
res_basin <- st_intersection(res, basins)

# st_intersection with rivers to identify which basin they fall under: 
rivers <- enviro_sw$rivers %>%
  mutate(name = toupper(name)) %>%
  rename(river_name = name)
riv_basin <- st_intersection(rivers, basins) %>%
  select(river_name:geometry) %>%
  select(-shape_len) %>%
  rename(res_name = river_name) %>%
  mutate(type = "River", 
         status = "Existing") %>%
  unique()

# combining rivers and reservoirs: 
riv_res_basin <- rbind(riv_basin, res_basin)

# matching reservoir, basins, and river with surface water intakes: 
sw <- enviro_sw$surface_water_intakes %>%
  select(-objectid) %>%
  rename()

sw_basins <- merge(riv_res_basin, sw, by.x = "res_name", by.y = "waterbody",
                   all = TRUE) %>%
  # removing reservoirs that are not a water supply (and therefore 
  # no matching sw_intake)
  filter(type == "Water Supply" | is.na(type) | type == "River")
# missing pwsid = reservoir or river that might not have a surface water intake 
# missing type = pwsid that does not match to a reservoir or river

# plotting surface water intakes to visualize gaps
points_sf <- sw_basins %>%
  as.data.frame() %>%
  select(-geometry) %>%
  filter(!(is.na(lat_dd)))  %>%
  st_as_sf(., coords = c("long_dd", "lat_dd"), 
           crs = st_crs(res_basin)) 
         
# where are we missing the links between surface water intakes and pwsids? 
viol_pal <- colorFactor(
  palette = viridis::plasma(23),
  domain = basins$baisn_name_short)

leaflet() %>%
  addProviderTiles(providers$CartoDB.VoyagerNoLabels, group = "Toner Lite") %>%
  addPolygons(data = basins,
              opacity = 9,
              color = ~viol_pal(basin_name_short),
              weight = 1,
              label = paste0("basin: ", basins$basin_name_short)) %>%
  addLegend("bottomright",
            pal = viol_pal,
            values = basins$basin_name_short,
            title = "Basin",
            opacity = 1) %>%
  addPolygons(data = res_basin,
              opacity = 9,
              color = ~viol_pal(basin_name_short),
              weight = 1,
              label = paste0("res: ", res_basin$res_name, "; basin: ",
                             res_basin$basin_name_short)) %>%
  addCircleMarkers(data = points_sf, 
                   fill = ~viol_pal(basin_name_short), 
                   color = ~viol_pal(basin_name_short), 
                   weight = 3, 
                   opacity = 9, 
                   radius = 3, 
                   label = paste0("pwsid: ", points_sf$pwsid, "; waterbody: ", 
                                  points_sf$res_name, 
                                  "; basin name: ", 
                                  points_sf$basin_name_short)) %>%
  addPolylines(data = riv_basin,
               opacity = 9,
               color = ~viol_pal(basin_name_short),
               weight = 2,
               label = paste0("river: ", riv_basin$res_name))
  
# TODO: need some fuzzy string matching with reservoirs and nearby 
# surface water intakes 

# crosswalking key - completing this step once initial merging is complete
# cross_key <- sab_census %>%
#   as.data.frame(.) %>%
#   select(pwsid:county_served) 

```

## East TX counties
```{r}
east_tx <- read.csv("./data/raw/east_TX_counties.csv") %>%
  select(-X)

# grabbing census geographies: 
east_tx_geom <- census_tidy %>%
  filter(counties %in% paste0(east_tx$county_name, " County")) %>%
  group_by(counties) %>%
  summarize()

# also grabbing the larger geography of TX
tx_geom <- census_tidy %>%
  summarize()

leaflet() %>%
  addProviderTiles(providers$CartoDB.VoyagerNoLabels, group = "Toner Lite") %>%
  addPolygons(data = tx_geom, 
              color = "grey", 
              weight = 1) %>%
  addPolygons(data = east_tx_geom,
              opacity = 9,
              color = "#333c6bff",
              weight = 1)
```


NEXT STEPS: 
questions: 
- Well numbers to pwsids 
- % people on well water in east TX 
- do people think about water at the county scale?
- Incorporating Danielle's SRF analysis 
- TX SRF data - next Friday 

steps: 
- adding this to a list and the onto S3
- breaking collection & merging script 
- function that takes list from s3 and queries for categories 
- crosswalk data frame included as metadata in the list 

data collection 
- nixing cvi
- incorporating CEJST specific metrics regarding whether disadv or not 
- incorporate EJ screen (R package and then API) regarding water-specfic hazards 