---
title: "Texas Drinking Water Data Collating & Merging"
author: "EmmaLi Tsai"
date: "2024-02-09"
output: html_document
---

Packages: 
```{r packages}
library(tidyverse)
library(leaflet)
library(sf)
library(aws.s3) 
library(janitor)
library(tidycensus)
library(areal)
library(mapview)
library(httr)
library(readxl)
library(jsonlite)
library(googlesheets4)
# remotes::install_github("environmental-policy-innovation-center/stateDW", 
#                         auth_token = "insert-github-pat-here")
library(stateDW)

# Store keys on local machine in .Renviron file! 
```

Functions: 
```{r functions}
# Migrated to R package workflow but keeping these here as a backup 
# source("./code/functions/get_sab.R")
# source("./code/functions/get_SDWIS.R")
# source("./code/functions/interp_sab_census.R")
```

_______________________________________________________________________________
### Data Collating - pulling data from APIs, web links, or S3 buckets:  Data here are either pulled from S3 (see code above for how they got there) or downloaded easily from websites or APIs
_______________________________________________________________________________
## Demographic
```{r data collating}
## Demographic #################################################################
# SABs
sab_census <- aws.s3::s3read_using(st_read, 
                           object = "state-drinking-water/TX/raw/TX_sab_census.geojson",
                           bucket = "tech-team-data")
# CEJST
cejst_raw <- aws.s3::s3read_using(read.csv, object = "s3://tech-team-data/ej-data/cejest-v1/1.0-communities.csv") %>%
  janitor::clean_names() %>%
  filter(state_territory == "Texas") %>%
  select(c("census_tract_2010_id","total_threshold_criteria_exceeded",
           "total_categories_exceeded","identified_as_disadvantaged",
           "percentage_of_tract_that_is_disadvantaged_by_area","total_population"))%>%
  rename("total_population_cejst" = "total_population")

# Social Vulnerability Index 
svi <- aws.s3::s3read_using(read.csv, 
                            object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_SVI_2020.csv")

# Climate Vulnerability Index: 
cvi <- aws.s3::s3read_using(read.csv, 
                            object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_CVI.csv")

# TX EJ Screen: 
tx_ejscreen <- aws.s3::s3read_using(read.csv, 
                                    object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_ejscreen.csv")
```
## Environmental
```{r}
## Environmental ###############################################################
## Surface water intake locations: https://gis-tceq.opendata.arcgis.com/datasets/TCEQ::surface-intakes/api
# Pulled Jan 22 2024 
sw_intake <- GET("https://gisweb.tceq.texas.gov/arcgis/rest/services/Public/PWS/MapServer/1/query?where=1%3D1&outFields=*&outSR=4326&f=json")
sw_intake_tidy <- fromJSON(rawToChar(sw_intake$content))$features$attribute %>%
  as.data.frame() %>%
  janitor::clean_names() %>%
  rename(pwsid = pws_id) %>%
  mutate(pwsid = paste0("TX", pwsid))

## groundwater quality & levels: 
# function for reading gwdb tables, which have unique separators: 
read_gwdb_table <- function(x) {
     read.table(x, header = TRUE, fill = TRUE, sep = "|")
}
# grabbing well metadata: 
well_main <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WellMain.txt") %>%
  janitor::clean_names()

# water quality information: 
wq_major <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WaterQualityMajor.txt") %>%
  janitor::clean_names()
wq_minor <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WaterQualityMinor.txt") %>%
  janitor::clean_names()
wq_other <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WaterQualityOtherUnassigned.txt") %>%
  janitor::clean_names()
# let's just combine all of these 
well_wq <- rbind(wq_major, wq_minor, wq_other)

# water level information: 
wl_major <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WaterLevelsMajor.txt") %>%
  janitor::clean_names()
wl_minor <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WaterLevelsMinor.txt") %>%
  janitor::clean_names()
wl_other <- aws.s3::s3read_using(read_gwdb_table, 
                     object = "s3://tech-team-data/s3://tech-team-data/state-drinking-water/TX/raw/TX_GWDB/WaterLevelsOtherUnassigned.txt") %>%
  janitor::clean_names()
# let's just combine all of these 
well_wl <- rbind(wl_major, wl_minor, wl_other)

## Major aquifers: https://www.twdb.texas.gov/mapping/gisdata/doc/major_aquifers.zip
# downloaded 1/23/2024
file_loc <- "./data/raw/TX_major_aquifers"
download.file("https://www.twdb.texas.gov/mapping/gisdata/doc/major_aquifers.zip", 
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
file.remove(paste0(file_loc, ".zip"))
maj_aquifers <- st_read(paste0(file_loc, "/NEW_major_aquifers_dd.shp")) %>%
  janitor::clean_names()

## Minor aquifers: https://www.twdb.texas.gov/mapping/gisdata/doc/minor_aquifers.zip
# downloaded 1/23/2024
file_loc <- "./data/raw/TX_minor_aquifers"
download.file("https://www.twdb.texas.gov/mapping/gisdata/doc/minor_aquifers.zip", 
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
file.remove(paste0(file_loc, ".zip"))
minor_aquifers <- st_read(paste0(file_loc, "/Minor_Aquifers.shp")) %>%
  janitor::clean_names()

## Major river basins: https://www.twdb.texas.gov/mapping/gisdata/doc/Major_River_Basins_Shapefile.zip
# downloaded 1/23/2024
file_loc <- "./data/raw/TX_major_river_basins"
download.file("https://www.twdb.texas.gov/mapping/gisdata/doc/Major_River_Basins_Shapefile.zip", 
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
file.remove(paste0(file_loc, ".zip"))
major_river_basins <- st_read(paste0(file_loc, "/TWDB_MRBs_2014.shp")) %>%
  janitor::clean_names()

## Existing reservoirs: https://www.twdb.texas.gov/mapping/gisdata/doc/Existing_Reservoirs.zip
# downloaded 1/23/2024
file_loc <- "./data/raw/TX_reservoirs"
download.file("https://www.twdb.texas.gov/mapping/gisdata/doc/Existing_Reservoirs.zip", 
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
file.remove(paste0(file_loc, ".zip"))
reservoirs <- st_read(paste0(file_loc, "/TWDB_SWP2012_Major_Reservoirs.shp")) %>%
  janitor::clean_names()

## Surface water 303(d) river basin impairment: https://texaswaterexplorer.tnc.org/index.html
# downloaded 1/24/2024
file_loc <- "./data/raw/TX_basin_impairment_2012"
download.file("https://texaswaterexplorer.tnc.org/downloads/Data_WQImpairment.zip", 
              destfile = paste0(file_loc, ".zip"))
unzip(zipfile = paste0(file_loc, ".zip"), exdir = file_loc) 
file.remove(paste0(file_loc, ".zip"))
basin_impairment <- readxl::read_excel(paste0(file_loc, "/Data_WQImpairment.xlsx"), sheet = "Basins")%>%
  janitor::clean_names()

# reservoir levels - recent
recent_reservoir_levels <- aws.s3::s3read_using(read.csv, 
                                         object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_reservoir_levels_recent.csv") %>%
  select(-X)

# reservoir levels - historic
reservoir_levels_10yr <- aws.s3::s3read_using(read.csv, 
                                         object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_reservoir_levels_10yr.csv") %>%
  select(-X)

# groundwater levels
recent_well_levels <- aws.s3::s3read_using(read.csv, 
                                           object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_recent_well_levels.csv") %>%
  select(-X)

# TCEQ DWW system data: 
# NOTE: there is A TON more information in s3 for us to pull like enforcement 
# actions, etc. but starting the structure here for grabbing that information
# pulling data from two different webscraping events and locating pwsids that 
# are community water systems: 
water_details_dec17 <- aws.s3::s3read_using(read.csv, 
                                      object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_TCEQ_DWW/TX_20231217/WaterSystemDetail/TX_Water System Details_20231217_20231229203427.csv") 
water_details_dec3rd <- aws.s3::s3read_using(read.csv, 
                                      object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_TCEQ_DWW/TX_20231203/WaterSystemDetail/TX_Water System Details_20231203.csv") 
combined_details <- rbind(water_details_dec17, water_details_dec3rd)
water_details <- pivot_wider(combined_details, names_from = "Type", values_from = "Value") %>%
  janitor::clean_names()
# grabbing just community water systems (tceq has 10k that are not cws): 
tceq_cws <- water_details %>%
  filter(system_type == "C") # 4666

# pulling violations: 
viol_dec17 <- aws.s3::s3read_using(read.csv, 
                                      object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_TCEQ_DWW/TX_20231217/Violations/TX_Individual Violations_20231217_20231229203427.csv") 
viol_dec3rd <- aws.s3::s3read_using(read.csv, 
                                      object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_TCEQ_DWW/TX_20231203/Violations/TX_Individual Violations_20231203.csv") 
combined_viols <- rbind(viol_dec17, viol_dec3rd) %>%
  janitor::clean_names() %>%
  filter(water_system_no %in% tceq_cws$water_system_no) # 118,659
# SDWIS contains 20,000 more violations  
```
## Water Delivery System
```{r}
## Water Delivery System ######################################################
# historic use estimates: 
historic_use_estimates <-  aws.s3::s3read_using(read.csv, 
                                                object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_historic_use_estimates.csv") %>%
  select(-X)
# need to fix the pwsid column: 
for(i in 1:nrow(historic_use_estimates)){
  historic_use_estimates$pwsid[i] <- paste0("TX", paste(rep(0, (7-nchar(historic_use_estimates$tceq_pws_code[i]))), 
                                                        collapse = ""), 
                                            historic_use_estimates$tceq_pws_code[i], collapse = "")
}

# water loss audits
# vector of cleaned column names 
waterlosscolnames <- c("Name of Utility",
                       "Real Loss GMD (<32 conn/mi)",
                       "Real Loss GCD",
                       "Apparent Loss GCD",
                       "Water Loss GCD",
                       "ILI (>= 3,000 connections)", 
                       "Total GPCD", 
                       "GPCD Loss",
                       "Real Loss Cost in dollars",
                       "Apparent Loss Cost in dollars","Year")

#making empty dataframe 
waterloss_all <- setNames(data.frame(matrix(ncol = length(waterlosscolnames), nrow = 0)), waterlosscolnames)

year <- 2010 

# ! edit this when adding additional years 
stopyear <- 2022 

while(year <= stopyear)
{
print(year)
waterloss_raw <- aws.s3::s3read_using(read_csv, 
                               object = paste0("s3://tech-team-data/state-drinking-water/TX/raw/TCEQ_Loss_Audits/SummaryAuditsByCategory_",year,".csv"))

waterloss <- waterloss_raw %>%
             select(1:10)

colnames(waterloss) <- waterlosscolnames

waterloss <- waterloss %>%
             janitor::clean_names()%>%
             filter(name_of_utility != "UtilityName")%>%
             mutate(Year = year)

waterloss_all <- rbind(waterloss_all,waterloss)
year <- year + 1
}

## 2022 TX State Water Plan: https://2022.texasstatewaterplan.org/statewide?
# downloaded Jan 29 2024
pop <- read.csv("https://2022.texasstatewaterplan.org/download/statewide/population") %>%
  janitor::clean_names()
demands <- read.csv("https://2022.texasstatewaterplan.org/download/statewide/demands") %>%
  janitor::clean_names()
supp <- read.csv("https://2022.texasstatewaterplan.org/download/statewide/supplies") %>%
  janitor::clean_names()
needs <- read.csv("https://2022.texasstatewaterplan.org/download/statewide/needs") %>%
  janitor::clean_names()
strat <- read.csv("https://2022.texasstatewaterplan.org/download/statewide/strategies") %>%
  janitor::clean_names()
```
## Financial 
```{r}
## Financial ##################################################################
## Water affordability: https://github.com/NIEPS-Water-Program/water-affordability/raw/main/data/rates_data/rates_tx.xlsx
# downloaded Jan 24 2024
link <- "https://github.com/NIEPS-Water-Program/water-affordability/raw/main/data/rates_data/rates_tx.xlsx"
temp_file <- tempfile(fileext = ".xlsx")
water_aff <- GET(link, write_disk(path = temp_file))
water_aff <- readxl::read_excel(temp_file) %>%
  janitor::clean_names()

# PPL data & tidying 
TX_PPL <- aws.s3::s3read_using(read.csv, 
                               object = "s3://water-team-data/clean_data/srf_project_priority_lists/web_ppl_combined_clean.csv") %>%
  mutate(across('Project.Type', 
                str_replace, 'Other', 'General')) %>%
  filter(State == "Texas") %>%
  janitor::clean_names()

## Danielle's SRF analysis: had to download it using a different method because 
# it's an excel file that's stored on google drive - so it's not actually a 
# google sheet. 
# downloaded 2/1/2024 
URL <- "https://docs.google.com/spreadsheets/d/1kUHLdzFjwmKx4r6YTu2qjxwNuuG5JeUi/edit#gid=252194392"
dl <- googledrive::drive_download(
  as_id(URL),
  path = 'temp1.xlsx', 
  type = "xlsx")
cwsrf_by_rank <- readxl::read_excel('temp1.xlsx', sheet = "CWSRF By Rank", 
                                        skip = 1, col_names = TRUE)

## TX DAC - whether they are likely to qualify for DAC status - requires pulling water affordability data and combining this with mhi
TX_DAC <- aws.s3::s3read_using(st_read, 
                               object = "s3://tech-team-data/state-drinking-water/TX/raw/TX_DAC.geojson")
```
_______________________________________________________________________________
### Organizing data into a list:
_______________________________________________________________________________
```{r}
# TX_data <- list()
# TX_data[[1]] <- sab_census
# TX_data[[2]] <- sdwis_tidy
# names(TX_data) <- c("census", "sdwis")
# ... 
# # here's how you get the list entry out and into a data frame: 
# sdwis <- do.call(data.frame, TX_data[[2]])
```

NEXT STEPS: 
questions: 
- Well numbers to pwsids 
- % people on well water in east TX 
- do people think about water at the county scale?
- Incorporating Danielle's SRF analysis 
- TX SRF data - next Friday 

steps: 
- adding this to a list and the onto S3
- breaking collection & merging script 
- function that takes list from s3 and queries for categories 
- crosswalk data frame included as metadata in the list 

data collection 
- nixing cvi
- incorporating CEJST specific metrics regarding whether disadv or not 
- incorporate EJ screen (R package and then API) regarding water-specfic hazards 



